---
title: "Derived Variables"
author: "Diego Vitali"
output: 
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib

---
```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(NbClust)
library(cluster)
library(readr)

source("Sources/Paper1_func.R")

Imputed_datasets<-readRDS("~/private_WP5/WP5_data/rds/mice_imputationlong_VH.rds")
paper1_rds_path<-"~/private_WP5/WP5_data/rds/"
paper1_MODELS_rds_path<-"~/private_WP5/WP5_data/model_fits/"

# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

```

# Propagate derived variables 
Here we set up our derived variables and we propagate in all imputed datasets

# Import imputations
```{r echo=FALSE}


library(mice)
imputed.df<-mice::complete(Imputed_datasets,1)
```
incomplete variables that were imputed:
* "painlv", 
* "painchr_ext", 
* "shlt", 
* "lnlys",
* "ssupport6_log1p", 
* "nettotw_bu_s_log"

## Pass-through variables from upstream pipeline

The following LTA variables are not derived in this file — they enter via
`mice_imputationlong_VH.rds` and are propagated as-is through `updated_mids`:

| Variable | Derived in | Location |
|---|---|---|
| `raeduc_e` | `PAPER_1.pt1_Data_prep.Rmd` | line 375: `-15` recoded to `2` ("Other education"); imputed with `2lonly.pmm` in pt2 |
| `MLTC` | `PAPER_1.pt1_Data_prep.Rmd` | lines 683–702: `create_MLTC_variable()` — Poole 2020 definition + psyche; imputed with `2l.pmm` in pt2 |
| `cesd_lat_shifted` | `PAPER_1.pt2_IMPUTATION_MICE_long.Rmd` | lines 174–176: `rXcesd_lat - global_min_cesd_lat` (shift before imputation); imputed with `2l.pmm` |

## Derived variables

### Health impact scores
Impact is a derived variable based on lvimp_shifted and this was a complete variable

We use wave 4 as the baseline to get the reference medoid from this

```{r}
#Use first wave as reference (first wave)
imputed.df4<-imputed.df%>%filter(wave == 4)
nonzero_data_ref <- na.omit(imputed.df$lvimp_shifted) # there is no missingness but I am checking this 

# Exclude near-zero IRT scores before clustering.
# Under partial scalar invariance, two items (lifta, hlthlm) have freed
# intercepts, which means respondents with genuinely zero health impact
# can still receive a small positive IRT score (~0.00079) rather than
# exactly 0. The threshold 0.1 sits well above this artefactual floor
# and cleanly separates true "no impact" respondents (coded 0 in
# lvimp_shifted_cat) from those with measurable impact.
nonzero_data_ref<- subset(imputed.df4, lvimp_shifted > 0.1)$lvimp_shifted

library(mclust)
  fit <- Mclust(nonzero_data_ref, G = 1:4)  # BIC selects k
  summary(fit)
# Mclust suggests k=3 by BIC, but the poor cluster separation (BIC–ICL gap of ~733) undermines that solution. 
# Proceeding with K=2


# TWO-PART PARAMETRISATION OF lvimp_shifted
# -----------------------------------------
# lvimp_shifted is zero-inflated. The zero mass represents respondents with
# no measurable health impact — a qualitatively distinct state defined by the
# scale itself, not a low value on the same continuum as non-zero scores.
# Category 0 ("no impact") is therefore not discovered by clustering; it is
# assigned a priori from substantive knowledge of the instrument.
#
# Clustering operates only on the non-zero subspace (lvimp_shifted > 0.1),
# characterising the distribution of measurable impact. The resulting clusters
# are then rejoined with the zero group to form the final ordered indicator:
#   0 = no impact     (structural zeros — a priori)
#   1 = lower impact  \
#   2 = higher impact  > empirically determined by PAM below
#  [3 = highest impact if k=3 is selected]
#
# This mirrors the logic of two-part / hurdle models for zero-inflated outcomes.
#
# PAM (Partitioning Around Medoids) is preferred over k-means because:
#   - Medoids are actual observed IRT scores — interpretable on the original scale
#   - Does not assume normality (k-means implicitly does via L2 minimisation)
#   - More robust to outliers than k-means
# Note: for univariate data, Manhattan and Euclidean distance are equivalent;
# metric choice has no effect on the result here.
library(cluster)
################  
n_clusters<-3 ##
################
pam_result_3 <- pam(nonzero_data_ref, k = n_clusters, metric = "manhattan")
pam_result_3$medoids   # cluster medoids
pam_result_3$clustering %>% table  # cluster assignments

sil_3 <- silhouette(pam_result_3$clustering, dist(nonzero_data_ref, method = "manhattan"))
average_sil_3<-mean(sil_3[,3])
plot(sil_3, main = "Silhouette Plot for PAM Clustering", 
     col = 1:n_clusters, border = NA)

summary(pam_result_3)
average_sil_3
################
n_clusters<-2 ##
################
pam_result_2 <- pam(nonzero_data_ref, k = n_clusters, metric = "manhattan")
pam_result_2$medoids   # cluster medoids
pam_result_2$clustering %>% table  # cluster assignments
summary(pam_result_2)

sil_2 <- silhouette(pam_result_2$clustering, dist(nonzero_data_ref, method = "manhattan"))
average_sil_2<-mean(sil_2[,3])
plot(sil_2, main = "Silhouette Plot for PAM Clustering", 
     col = 1:n_clusters, border = NA)
summary(pam_result_2)
average_sil_2
```

<!-- CLUSTERING DECISION: k=3, with top two clusters collapsed for modelling

k=2 vs k=3 PAM comparison on non-zero lvimp_shifted scores (n=3049):

  k=2: overall silhouette = 0.602
       cluster 1 (medoid 1.620, n=1616): silhouette = 0.364  ← weak
       cluster 2 (medoid 0.879, n=1433): silhouette = 0.870

  k=3: overall silhouette = 0.701
       cluster 1 (medoid 1.392, n=1056): silhouette = 0.611
       cluster 2 (medoid 2.025, n=687):  silhouette = 0.440  ← fuzzy boundary
       cluster 3 (medoid 0.879, n=1306): silhouette = 0.911
       inter-cluster separation (clusters 1–2): 0.016         ← very small

  Mclust (G=1:4) independently selected k=3 by BIC, but with a large
  BIC–ICL gap (~733), indicating substantial overlap between components —
  consistent with the poor separation between the mid and high clusters above.

DECISION: pull k=3 medoids (0.879 / 1.392 / 2.025) for parametrisation.
The mid (1.392) and high (2.025) impact clusters are empirically identified
but merge into lvimp_shifted_cat3 (categories 2+3 → 2) for modelling, because
their boundary is insufficiently sharp (silhouette 0.440, separation 0.016).
This merge is data-driven, not assumed: k=2 would have obscured the distinction
entirely; k=3 documents it and then collapses it with justification.

Final lvimp_shifted_cat categories entering LTA:
  0 = no impact     (structural zeros — a priori, not clustered)
  1 = not high impact    (PAM cluster 3, medoid 0.879)
  2 = mid/high impact    (PAM cluster 1, medoid 1.392)
  3 = high impact   (PAM cluster 2, medoid 2.025)

lvimp_shifted_cat3 (used in LTA model):
  0 = no impact
  1 = some but not "high impact"
  2 = high impact  (categories 2+3 merged due to poor separation)
-->

```{r,echo=FALSE, eval = F, warning=FALSE,message=FALSE,error=FALSE}
# eval=FALSE: PAM bootstrap is expensive. Pre-computed result stored in
# Impact_clustStability_boot2000.rds. Set eval=TRUE to regenerate.

set.seed(200720)

n_boot <- 2000
# Bootstrap stability validation for k=3 (selected above)
# Nb clust and our theory about these cluster to us to select 3 clusters
n_clusters <- 3

library(fpc)
## define the pam function to be used by clusterboot (as per documentation)
pamCBI <- function(x, k) {
  pamfit <- pam(x, k, metric = "manhattan")
  partition <- pamfit$clustering
  nclusters <- length(unique(partition))
  clusterlist <- lapply(1:nclusters, function(i) partition == i)
  return(list(result = pamfit,
              nc = nclusters,
              clusterlist = clusterlist,
              partition = partition,
              clustermethod = "pam"))
}

ImpactBootstability_results <- clusterboot(nonzero_data_ref,
                          clustermethod = pamCBI,
                          k = n_clusters,
                          B = n_boot,          # number of bootstrap samples
                          seed = 200720)     # for reproducibility

saveRDS(ImpactBootstability_results,paste0(paper1_MODELS_rds_path,"Impact_2_clust_Stability_boot2000.rds"))

```

```{r,echo=FALSE, eval = T, warning=FALSE,message=FALSE,error=FALSE}

ImpactBootstability_results<-readRDS(paste0(paper1_MODELS_rds_path,"Impact_clustStability_boot2000.rds"))

# Print bootstrap mean Jaccard similarity (cluster stability)
# Hennig, C. (2007) Cluster-wise assessment of cluster stability. Computational Statistics and Data Analysis, 52, 258-271.ara
cat("Bootstrap mean Jaccard similarity (cluster stability) Hennig, C. (2007)")
cat(sprintf("  Cluster 1 Stability: %.4f\n", ImpactBootstability_results$bootmean[1]))
cat(sprintf("  Cluster 2 Stability: %.4f\n", ImpactBootstability_results$bootmean[2]))
#ImpactBootstability_results$result
```


```{r}
imputed.df4<-imputed.df%>%filter(wave == 4)
nonzero_data_ref <- na.omit(subset(imputed.df4, lvimp_shifted > 0.1)$lvimp_shifted)
pam_result <- pam(nonzero_data_ref, k = 3, metric = "manhattan")
pam_result$medoids   # cluster medoids

ref_centers_IMPACT <- sort(pam_result$medoids)

# Cross-wave parametrisation using wave 4 reference medoids.
# Partial scalar measurement invariance (confirmed in Measurement_invariance_IRT.Rmd)
# justifies applying the same medoid thresholds to waves 5 and 6: the IRT scores
# are on a common metric across waves, so wave 4 boundaries are directly comparable.
#
# categorize_wave() assigns each observation to the nearest medoid by absolute
# distance, with boundaries at the midpoints between consecutive medoids:
#   0 / 1 boundary: 0.1        (hard threshold — structural zeros, see above)
#   1 / 2 boundary: (0.879 + 1.392) / 2 = 1.136
#   2 / 3 boundary: (1.392 + 2.025) / 2 = 1.709
#
# lvimp_shifted_cat (4 levels, used as LTA indicator):
#   0 = no impact       (structural zeros, a priori)
#   1 = low impact      (medoid 0.879)
#   2 = mid impact      (medoid 1.392)
#   3 = high impact     (medoid 2.025)
#
# lvimp_shifted_cat3 (3 levels, used for descriptives and state ordering):
#   0 = no impact
#   1 = low impact
#   2 = mid/high impact (categories 2+3 merged — poor inter-cluster separation:
#                        silhouette 0.440, separation 0.016)
#
# Observed wave-level distributions (imputation 1):
#   wave 4:  0=4965  1=1306  2=1056  3=687
#   wave 5:  0=4821  1=1226  2=1134  3=833
#   wave 6:  0=4620  1=1258  2=1116  3=1020
# Category 0 decreases and categories 2+3 grow across waves, consistent with
# progressive functional decline in an ageing cohort. Category 1 is stable,
# consistent with its role as a transition zone. No ordering violations.

updated_mids <- IMPACT_parametrisation_to_mice_objs(Imputed_datasets, ref_centers_IMPACT)
# the mids now contain:
#   * lvimp_shifted_cat3 : the three level ordinal for pain impact (see above)
#   * lvimp_shifted_cat : the four level ordinal for pain impact  (see above)

```



```{r}
# Step 1: prepare both tables
freq_cat3 <- mice::complete(updated_mids, 1) %>%
  filter(wave %in% c(4, 5, 6)) %>%
  dplyr::count(wave, lvimp_shifted_cat3) %>%
  mutate(pct  = round(100 * n / sum(n), 1), .by = wave) %>%
  mutate(cell = paste0(n, " (", pct, "%)")) %>%
  select(wave, lvimp_shifted_cat3, cell) %>%
  pivot_wider(names_from = lvimp_shifted_cat3, values_from = cell, names_prefix = "Cat ")

freq_cat <- mice::complete(updated_mids, 1) %>%
  filter(wave %in% c(4, 5, 6)) %>%
  dplyr::count(wave, lvimp_shifted_cat) %>%
  mutate(pct  = round(100 * n / sum(n), 1), .by = wave) %>%
  mutate(cell = paste0(n, " (", pct, "%)")) %>%
  select(wave, lvimp_shifted_cat, cell) %>%
  pivot_wider(names_from = lvimp_shifted_cat, values_from = cell, names_prefix = "Cat ")

# Step 2: render side by side
kable(list(freq_cat, freq_cat3),
      caption = c("lvimp_shifted_cat", "lvimp_shifted_cat3")) 
# Categories 2 and 3 in the 4-level model are collapsed into a single "mid/high
# impact" category in lvimp_shifted_cat3. Both k=2 and k=3 PAM solutions are
# bootstrap-stable (Jaccard > 0.60), but k=3 is preferred because:
#   (a) it identifies three empirically distinct non-zero groups rather than
#       assuming a single binary split;
#   (b) the poor separation between categories 2 and 3 (silhouette 0.440,
#       inter-cluster separation 0.016) provides a principled, data-driven
#       justification for merging them — the collapse follows empirical observation
#       and it follow clinical principles by which we privilege the detection of 
#       high impact on health more inclusively rather than missing it a false negative
```

### Net total Wealth Quintiles

We wave 4 quantile breaks calculated from one reference dataset to ensure consistent categorization across all imputations
```{r, eval=T, echo=FALSE}
# Calculate wave 4 quintile breaks from complete case or first imputation
data_imp1 <- mice::complete(updated_mids, 1)

global_min <- -436410

data_imp1$nettotw_bu_s_original<- exp(data_imp1$nettotw_bu_s_log) + global_min - 1
cutpoints <- quantile(data_imp1$nettotw_bu_s_original,
                     probs = seq(0, 1, 0.2),
                     na.rm = TRUE)
print(cutpoints)

quintile_breaks_log <- quantile(data_imp1$nettotw_bu_s_log, # pooled quintile repartition
                                  probs = seq(0, 1, by = 0.2), na.rm = TRUE)



updated_mids <- add_wealth_quintiles_to_mice(updated_mids, 
                                            quintile_breaks_log)


```

### Marital support/strain


    "sustdfe" = "Partner understands the way they feel",
    "srely" = "can rely on Partner for a serious problem", 
    "sopenup" = "can open up their worries to Partner",
    "scritze" = "Partner criticizes r",
    "sletdow" = "Partner let r down when counting on",
    "sgetnev" = "Partner gets on r's nerves"

the total score here is log1p() transformed to improve its skewed distribution. The total score is monotonic with greater scores meaning more strain the supportiveness of the relationship

### Lonelyness bin

```{r}
data_imp1 <- mice::complete(updated_mids, 1)
data_w4<-data_imp1%>% filter(wave == 4)
data_w5<-data_imp1%>% filter(wave == 5)
data_w6<-data_imp1%>% filter(wave == 6)


cluster_wave <- function(data, wave_num, n_clusters){
 # PAM clustering
  pam_result <- pam(data$lnlys, k = n_clusters, metric = "manhattan")
  
  # Medoids
  cat("\nCluster medoids:\n")
  print(pam_result$medoids)
  
  # Cluster sizes
  cat("\nCluster sizes:\n")
  print(table(pam_result$clustering))
  
  # Silhouette analysis
  sil <- silhouette(pam_result$clustering, dist(data$lnlys, method = "manhattan"))
  
  # Plot silhouette
  plot(sil, main = paste("Silhouette Plot - Wave", wave_num), 
       col = 1:n_clusters, border = NA)
  
  # Average silhouette width
  cat("\nAverage silhouette width:", mean(sil[, 3]), "\n")
  
  # Return results
  return(list(
    wave = wave_num,
    pam_result = pam_result,
    silhouette = sil,
    medoids = pam_result$medoids,
    cluster_sizes = table(pam_result$clustering),
    avg_sil_width = mean(sil[, 3])
  ))
}

# n_clusters = 2: frozen value confirmed by PAM bootstrap stored in
# Impact_clustStability_boot2000.rds (eval=FALSE chunk, lines 69–129)
results_w4 <- cluster_wave(data_w4, n_clusters = 2, wave_num = 4)
results_w5 <- cluster_wave(data_w5, n_clusters = 2, wave_num = 5)
results_w6 <- cluster_wave(data_w6, n_clusters = 2, wave_num = 6)

comparison <- data.frame(
  Wave = c(4, 5, 6),
  Medoid_1 = c(results_w4$medoids[1], results_w5$medoids[1], results_w6$medoids[1]),
  Medoid_2 = c(results_w4$medoids[2], results_w5$medoids[2], results_w6$medoids[2]),
  Avg_Silhouette = c(results_w4$avg_sil_width, results_w5$avg_sil_width, results_w6$avg_sil_width),
  Cluster1_Size = c(results_w4$cluster_sizes[1], results_w5$cluster_sizes[1], results_w6$cluster_sizes[1]),
  Cluster2_Size = c(results_w4$cluster_sizes[2], results_w5$cluster_sizes[2], results_w6$cluster_sizes[2])
)

### thresholds are stable, 

apply_loneliness_clustering <- function(mids_object, ref_medoids = c(1.00, 1.75)) {
  
  # Extract long format data
  long_data <- mice::complete(mids_object, action = "long", include = TRUE)
  
  # Apply clustering
  processed_long <- long_data %>%
    mutate(
      # Assign to nearest medoid
      lnlys_cat = sapply(lnlys, function(x) {
        if(is.na(x)) return(NA_integer_)
        which.min(abs(x - ref_medoids))
      }),
      
      # Create factor label
      loneliness_group = factor(lnlys_cat, 
                               levels = 1:2, 
                               labels = c("Low", "High"))
    )
  
  # Convert back to mids object
  new_mids <- as.mids(processed_long, .imp = ".imp", .id = ".id")
  
  return(new_mids)
}

# Apply to your mids object
# ref_medoids c(1.00, 1.75) are hardcoded as the reference clustering centres.
# Validated as stable across all three waves via PAM (cluster_wave()):
#   Wave 4: medoids = 1.00, 1.75 | sizes = 3782, 4232 | avg silhouette = 0.604
#   Wave 5: medoids = 1.00, 1.75 | sizes = 4224, 3790 | avg silhouette = 0.613
#   Wave 6: medoids = 1.00, 1.75 | sizes = 3823, 4192 | avg silhouette = 0.615
# Silhouette widths >0.60 indicate well-separated clusters. Wave 5 label order
# is reversed in raw output (1.75 first) but medoid values are identical.
# Wave 4 medoids are used as the cross-wave reference (most common convention).
updated_mids <- apply_loneliness_clustering(updated_mids, ref_medoids = c(1.00, 1.75))


# Verify the clustering worked across imputations
#check_clustering <- mice::complete(updated_mids_with_clusters, action = "long", include = F)

# Check distribution by wave and imputation
# check_clustering %>%
#   group_by(.imp, wave, loneliness_group) %>%
#   summarise(
#     n = n(),
#     mean_lnlys = mean(lnlys, na.rm = TRUE),
#     max_lnlys = max(lnlys, na.rm = TRUE),
#     min_lnlys = min(lnlys, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   print(n = 30)

# Verify no label switching across imputations
# check_clustering %>%
#   group_by(.imp, loneliness_group) %>%
#   summarise(
#     mean_lnlys = mean(lnlys, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   arrange(loneliness_group,mean_lnlys,.imp ) %>% print(n = 100)


```

```{r eval = T, echo=FALSE }
df_plot<-imputed.df %>% filter(wave == 4 & !is.na(ssupport6_log1p))


ggplot(df_plot, aes(x=ssupport6_log1p)) + 
    geom_histogram(aes(y=after_stat(density)),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white") +
    #geom_density(alpha=.2, fill="#FF6666")+  # Overlay with transparent density plot
    geom_vline(aes(xintercept=mean(ssupport6_log1p, na.rm=T)), color="red", linetype="dashed", size=1) +
  labs(
    title = paste("Distribution of ssupport6_log1p with Mean Line (N =", nrow(df_plot),")"),
    caption = "Data source: ELSA, wave 4, SSupport score for respondents with partner"
  ) +
  theme_minimal()

```


### predictors semplification

We create two baseline Variables(Wave 4):

We Transform categorical and ordinal variables into binary variables to ease LTA model complexity and prevent instability or separation. For the same reason we set the reference category (0) so that the most endorsed binary category is the refence category.

* MLTC_BIN: Dichotomizes long-term conditions (≤2 vs >2)
* Sex_BIN: Converts gender coding to "Female"/"Male"
* partner: having a partner is coded as zero 
* painchr_ext: Recodes pain scale to start from 0 (required for LME/ST modeling)


* lnlys3_bin: is not parametrised for its poor sensitivity and specificity when parametrised  using the common threshold [see this ](https://pmc.ncbi.nlm.nih.gov/articles/PMC11578901/)




```{r, eval=T, echo=FALSE}

# This function takes a mids (multiply imputed datasets) object and creates 
# binary/categorical variables are labeled plus the indicators for the latent MM 
# are set as integer as required by lmest()

updated_mids <- label_variables_for_LTA(updated_mids)

```


## Write mids to rds
```{r, eval=FALSE, echo=FALSE}

saveRDS(updated_mids, paste0(paper1_rds_path,"mice_updatedMIDS_VH.rds"))


```
