---
title: "Multiple imputation Mice"
author: "Diego Vitali"
output: 
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib

---
```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(lavaan)
library(tidyr)
library(dplyr)
library(ggplot2)
library(naniar)
library(NbClust)
library(readr)
library(polycor)
library(patchwork)
library(corrplot)
library(GGally)
source("c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/Sources/Paper1_func.R")

paper1_rds_path<-"c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/paper1/"

H_elsa_w4_6<-readRDS("c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/paper1/H_elsa_w4_6.rds")


# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

```

This documentation describes our R-based pipeline for handling missing data using 
Multiple Imputation by Chained Equations (MICE). 

This pipeline operates on data in the long data format (one row per participant per wave, with a time/wave indicator).

*Features:*

* Conditional imputation: The code addresses conditional imputation scenarios, such as imputing "Marital strain" only for partnered individuals. This is implemented via the where_matrix in MICE, ensuring that imputation is only performed where appropriate.
* Data types: Variables are explicitly cast to factor (with or without ordering) or numeric  types as required. This is essential for MICE to select the correct imputation model and for downstream analyses.
* Multicollinearity: The script includes steps for checking multicollinearity using heterogeneous correlation matrices (polycor::hetcor). This is a good practice, as highly collinear predictors can destabilize imputation models.
* Minimal Missingness handling: For variables with very few missing values, simple imputation (median for numeric, mode for categorical) is performed before running MICE. This is efficient and avoids unnecessary complexity for almost-complete variables.
* Missingness visualisation: The use of naniar::vis_miss to visualize missingness patterns before and after imputation is excellent for transparency and quality control.

https://arxiv.org/pdf/2404.06967



## Age and immortal cohort assumption

We assume no death event (WAITING FOR HSE death data) and we take the baseline age as a timeinvariant predictor 

```{r}

H_elsa_w4_6<- H_elsa_w4_6%>% mutate(
                            r5agey = case_when(
                                    is.na(r5agey) ~ r4agey + 2,
                                    TRUE ~ r5agey
                                    ),
                            r6agey = case_when(
                                    is.na(r6agey) ~ r5agey + 2,
                                    TRUE ~ r6agey
                                    )
                            )
                            

H_elsa_w4_6$raagey<- H_elsa_w4_6$r4agey
      

```


## defining the imputation dataset

Select relevant variables: 
(race is excluded from the auxillary variable because sample is 97% white)
```{r}
Missingdata_wide <- H_elsa_w4_6 %>%
  select(c("idauniq",
          #varnames                          # description    (type); whether (comp)lete or (inc)omplete data
           paste0("r",4:6,"painlv"),         #                (categorical); incomp 
           paste0("r",4:6,"painchr_ext"),    #                (categorical); incomp 
           paste0("r",4:6,"lvimp"),          #                (categorical); incomp 
           paste0("r",4:6,"shlt"),           # self reported health               (categorical); incomp 
           paste0("r",4:6,"adltot6"),        # ADL total score                    (numeric); incomp 
           paste0("r",4:6,"partner"),        # having a partner                   (binary); comp 
           paste0("r",4:6,"ssupport6"),      # Marital strain score             (numeric, skewed) incomplete;
           paste0("r",4:6,"lnlys"),          # UCLA - loneliness                (numeric, skewed) incomplete;
           "raagey",                         # Baseline age                     (numeric, normal); comp 
           "ragender",                       # Gender                           (binary); comp 
           #"raracem",                        # Race/ethnicity                   (binary); comp
           paste0("r",4:6,"nettotw_bu_s"),     # Net wealth                     (numeric, skewed); incomp
           "raeduc_e",                       # Education                        (categorical); incomp
           paste0("r",c(4:6),"smoken"),      # smoking                          (binary); incomp
           paste0("r",4:6,"work"),           # Working status                   (binary); incomp
           paste0("r",4:6,"drinkwn_e"),      # Drinking                         (binary); incomp
#           paste0("r",c(4,6),"mbmi"),        # BMI                              (numeric); incomp
#           paste0("r",c(4,6),"hscrp"),       # CRP                              (numeric); incomp
           paste0("r",4:6,"PA_cat"),         # Physical activity                (categorical); incomp
           paste0("r",4:6,"cesd_lat"),       # Depression latent score          (numeric, skewed); incomp
           paste0("r",4:6,"ssupport6"),      # Marital strain score             (numeric, skewed) incomplete;
           paste0("r",4:6,"MLTC")            # Multiple long-term condit. count (numeric/count, skewed); incomp
  ))

#str(Missingdata_wide)
```

### Transform skewed variables

Wealth: Shift to positive and log-transform.

Impact and drinking: Shift/log-transform as appropriate.

Create auxiliary variables for change (e.g., MLTC_change_w5, MLTC_change_w6) to 
capture longitudinal change and reduce collinearity.

```{r}
###### TRANSFORM ###################
### Transform wealth data

# Calculate global minimums once
global_min_wealth <- min(c(Missingdata_wide$r4nettotw_bu_s, 
                          Missingdata_wide$r5nettotw_bu_s, 
                          Missingdata_wide$r6nettotw_bu_s), na.rm = TRUE)

global_min_lvimp <- min(c(Missingdata_wide$r4lvimp, 
                          Missingdata_wide$r5lvimp, 
                          Missingdata_wide$r6lvimp), na.rm = TRUE)

global_min_cesd_lat <- min(c(Missingdata_wide$r4cesd_lat, 
                          Missingdata_wide$r5cesd_lat, 
                          Missingdata_wide$r6cesd_lat), na.rm = TRUE)



# Apply same shift to all waves
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4nettotw_bu_s_log = log(r4nettotw_bu_s - global_min_wealth + 1),
    r5nettotw_bu_s_log = log(r5nettotw_bu_s - global_min_wealth + 1),
    r6nettotw_bu_s_log = log(r6nettotw_bu_s - global_min_wealth + 1)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4nettotw_bu_s,-r5nettotw_bu_s,-r6nettotw_bu_s)


### shift imp_data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4lvimp_shifted = r4lvimp - global_min_lvimp,
    r5lvimp_shifted = r5lvimp - global_min_lvimp,
    r6lvimp_shifted = r6lvimp - global_min_lvimp,
  )
# Remove original impact variables

Missingdata_wide<- Missingdata_wide %>%
  select(-r4lvimp,-r5lvimp,-r6lvimp)

### shift depr_lat data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4cesd_lat_shifted = r4cesd_lat - global_min_cesd_lat,
    r5cesd_lat_shifted = r5cesd_lat - global_min_cesd_lat,
    r6cesd_lat_shifted = r6cesd_lat - global_min_cesd_lat,
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4cesd_lat,-r5cesd_lat,-r6cesd_lat)


### Transform drinkning data
#Missingdata_wide$r4drinkwn_e %>% log() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4drinkwn_e_log  = log(r4drinkwn_e),
    r5drinkwn_e_log  = log(r5drinkwn_e),
    r6drinkwn_e_log  = log(r6drinkwn_e)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4drinkwn_e,-r5drinkwn_e,-r6drinkwn_e)


### Transform support data
#Missingdata_wide$r4ssupport6 %>% log1p() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4ssupport6_log1p  = log1p(r4ssupport6),
    r5ssupport6_log1p  = log1p(r5ssupport6),
    r6ssupport6_log1p  = log1p(r6ssupport6)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4ssupport6,-r5ssupport6,-r6ssupport6)


#str(Missingdata_wide)

# wave 4 ssupport missingness
ss4<-vis_miss(Missingdata_wide[which(Missingdata_wide$r4partner == 1),] %>% select(r4ssupport6_log1p))
# wave 5 ssupport missingness
ss5<-vis_miss(Missingdata_wide[which(Missingdata_wide$r5partner == 1),] %>% select(r5ssupport6_log1p))
# wave 5 ssupport missingness
ss6<-vis_miss(Missingdata_wide[which(Missingdata_wide$r6partner == 1),] %>% select(r6ssupport6_log1p))

(ss4 + ss5 + ss6)+
  plot_annotation(
    title = "Missingness in conditional variable `Marital strain` total score"
  )


Missingdata_wide %>% select(-r4ssupport6_log1p,-r5ssupport6_log1p,-r6ssupport6_log1p) %>% vis_miss()

```

## Defining variables type
```{r}

bin.vars<-  c(paste0("r",4:6,"work"),         # currently working (bl only)       (binary); incomp 
              paste0("r",4:6,"partner"),      # having a partner (imp condition)  (binary); comp 
              paste0("r",4:6,"smoken"),       # smoking (bl only)                 (binary); incomp
              "ragender"                     # gender                            (binary); incomp
        #      "raracem"                       # race                              (binary); comp
              )

ord.vars<-c(paste0("r",4:6,"PA_cat"),          # Physical Activity                (cat-ordinal); incomp
#            paste0("r",4:6,"shlt"),           # self-reported health             (cat-ordinal); incomp
            "raeduc_e",                        # education                        (cat-ordinal); incomp
            paste0("r",4:6,"painchr_ext")
            )        

cont.vars<-c(#paste0("r",4:6,"agey"),            # age                            (continuous); incomp
            "raagey",                          # age  (bl only)                   (continuous); incomp
             paste0("r",4:6,"shlt"),           # self-reported health             (cat-ordinal); incomp
             paste0("r",4:6,"drinkwn_e_log"),# units/week (bl only)             (continuous); incomp
             paste0("r",4:6,"lvimp_shifted"),  # units/week (bl only)             (continuous); incomp
             paste0("r",4:6,"nettotw_bu_s_log"),# net wealth (bl only)            (continuous); incomp
             paste0("r",4:6,"MLTC"),           # N. MLTC (bl only)                (cat-ordinal); incomp
             paste0("r",4:6,"lnlys"),          # Lonelyness
#             paste0("r",c(4,6),"mbmi"),        # MBMI                             (continuous); incomp
#             paste0("r",c(4,6),"hscrp"),       # CRP                              (continuous); incomp
             paste0("r",4:6,"adltot6"),        # count of 6 ADL items             (continuous); incomp
             paste0("r",4:6,"painlv"),         # Pain level (0-3)                 (continuous); incomp
             paste0("r",4:6,"ssupport6_log1p"),      # Marital strain                   (continuous); incomp 
             paste0("r",4:6,"cesd_lat")        # cesd                             (continuous); incomp 
            )

Missingdata_wide<-Missingdata_wide %>% mutate(
                                ragender = case_when(
                                          ragender == 1 ~ 0, # male
                                          ragender == 2 ~ 1, # female
                                          TRUE ~ NA_real_
                                          )#,
#                                raracem = case_when(
#                                          raracem == 1 ~ 1, # white
#                                          raracem == 4 ~ 0, #non-white
#                                          TRUE ~ NA_real_
#                                          ),
                                )

```
## Variable type definition

```{r}


# categorical variables,# The first step is that all binary and categorical incomplete variables need to be stored as type “factor”;  @wijesuriya2025multiple


Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(ord.vars), ~factor(., ordered = T)))

Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(bin.vars), ~factor(.)))

Missingdata_wide<-as.data.frame(Missingdata_wide)

str(Missingdata_wide)

```


# Minimal Missingmess

For variables with ≤5 missing values (0.06% of total sample N = 8014), impute using median (numeric) or mode (categorical).

While efficient, this approach doesn't propagate imputation uncertainty for these variables and should perhaps be reconsidered (?)
https://pmc.ncbi.nlm.nih.gov/articles/PMC3074241/
[@azur2011multiple]

```{r eval = T}
# Identify variables with minimal missingness
minimal_missing <- names(which(colSums(is.na(Missingdata_wide)) > 0 & 
                              colSums(is.na(Missingdata_wide)) <= 5))


# Simple imputation for these variables
for(var in minimal_missing) {
  if(is.numeric(Missingdata_wide[[var]])) {
    # For numeric variables: median imputation
    Missingdata_wide[[var]][is.na(Missingdata_wide[[var]])] <- 
      median(Missingdata_wide[[var]], na.rm = TRUE)
  } else if(is.factor(Missingdata_wide[[var]]) || is.character(Missingdata_wide[[var]])) {
    # For categorical variables: mode imputation
    mode_val <- names(sort(table(Missingdata_wide[[var]]), decreasing = TRUE))[1]
    Missingdata_wide[[var]][is.na(Missingdata_wide[[var]])] <- mode_val
  }
}


# Use r6smoken to fill missing values in earlier waves
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Fill r4smoken missing values with r6smoken
    r4smoken = case_when(
      !is.na(r4smoken) ~ r4smoken,  # Keep existing values
      is.na(r4smoken) ~ r6smoken    # Fill missing with wave 6
    ),
    
    # Fill r5smoken missing values with r6smoken  
    r5smoken = case_when(
      !is.na(r5smoken) ~ r5smoken,  # Keep existing values
      is.na(r5smoken) ~ r6smoken    # Fill missing with wave 6
    )
  )

#str(Missingdata_wide)
#summary(Missingdata_wide)
```


## wide to long

Define time-invariant variables (ID, gender, race, education).

Use pivot_longer() to convert repeated measures to long format, extracting the wave number and aligning variables accordingly.

Visualize missingness patterns overall and for conditional variables (e.g., Marital Strain for partnered individuals) using naniar::vis_miss.
```{r}

vars_time_invariant <- c("idauniq", "ragender", 
                         #"raracem", 
                         "raeduc_e","raagey")

Missingdata_long <- Missingdata_wide %>%
  pivot_longer(
    cols = -all_of(vars_time_invariant),
    names_to = c("wave", ".value"),
     # Improved pattern that ensures we match the full variable name
    names_pattern = "^r([456])(.+)$") %>%
      # Convert wave to numeric if needed
      mutate(wave = as.numeric(wave))


Missingdata_long  %>% vis_miss()


```



|Variable          |Type                      |Time Status     |Waves Available  |Missing % (Long Format) |
|------------------|--------------------------|----------------|-----------------|---------|
|idauniq           |Integer                   |Structural      |All waves        |Complete |
|ragender          |Binary                    |Time-invariant  |Fixed            |Complete |
|raeduc_e          |Categorical (5 levels)    |Time-invariant  |Fixed            |Variable |
|wave              |Numeric                   |Structural      |4, 5, 6          |Complete |
|painlv            |Continuous                |Time-varying    |4, 5, 6          |Variable |
|painchr_ext       |Ordinal (4 levels)        |Time-varying    |4, 5, 6          |Variable |
|shlt              |Ordinal (5 levels)        |Time-varying    |4, 5, 6          |Variable |
|adltot6           |Count/Continuous          |Time-varying    |4, 5, 6          |Variable |
|partner           |Binary                    |Time-varying    |4, 5, 6          |Complete |
|ssupport6_log1p   |Continuous                |Time-varying    |4, 5, 6          |Variable (conditional) |
|raagey            |Continuous                |Time-invariant  |Fixed            |Complete |
|smoken            |Binary                    |Time-varying    |4, 5, 6          |Variable |
|work              |Binary                    |Time-varying    |4, 5, 6          |Variable |
|PA_cat            |Ordinal (4 levels)        |Time-varying    |4, 5, 6          |Variable |
|cesd_lat          |Continuous                |Time-varying    |4, 5, 6          |Variable |
|MLTC              |Count/Continuous          |Time-varying    |4, 5, 6          |Variable |
|nettotw_bu_s_log  |Continuous (log-trans.)   |Time-varying    |4, 5, 6          |Variable |
|lvimp_shifted     |Continuous (shifted)      |Time-varying    |4, 5, 6          |Variable |
|drinkwn_e_log   |Continuous (log1p-trans.) |Time-varying    |4, 5, 6          |Variable |


### Data Structure

* Level 1: Observations (waves) nested within individuals
* Level 2: Individuals (participants)

**Level 1 Predictors** (Time-Varying): These are variables that change over time within the same individual across waves 4, 5, and 6. level 1 structural variable is therefore "wave"


```{r}
############################
#### **LEVEL 1 Predictors** 
###########################
# Binary time-varying
bin_TimeVar.long <- c("work","smoken","partner")

# Continuous time-varying
cont_TimeVar.long <- c("painlv","shlt","drinkwn_e_log","nettotw_bu_s_log",
                       "ssupport6_log1p","lvimp_shifted","cesd_lat_shifted",
                       "lnlys",
                       # the following are integer ordinal/count variables, we re ordered but treating as integer here to facilitate mice predictions
                       "adltot6","MLTC",
                       "PA_cat","painchr_ext") 

lv1_vars<-c(cont_TimeVar.long,bin_TimeVar.long)
```


**Level 2 Predictors** (Time-Invariant): These are variables that remain constant for each individual across all waves (or are measured once at baseline). Level 2 structural variable is therefore "idauniq"


```{r}
############################
#### **LEVEL 2 Predictors** 
###########################

# Binary time-invariant
bin_TimeInvar.long <- c("ragender") ## taking race out because it sample is 97% white

# Continuous time-invariant
cont_TimeInvar.long<- c("raagey", # Age at baseline (though technically increases, treated as baseline)
                        "raeduc_e") # this is ordered but treating as integer here to facilitate mice predictions

lv2_vars<-c(bin_TimeInvar.long,cont_TimeInvar.long)


```


Complete and incomplete vars
```{r}

complete.vars<-names(Missingdata_long)[colSums(is.na(Missingdata_long)) == 0]
incomplete.vars<-names(Missingdata_long)[colSums(is.na(Missingdata_long)) != 0]

# in mice the id variable is required to be integer format
Missingdata_long$idauniq <- as.integer(Missingdata_long$idauniq)  # Convert idauniq to integer

# converting ordered factors to integers
Missingdata_long<-Missingdata_long %>%
                    mutate(
                      adltot6 = as.integer(adltot6),
                      raeduc_e = as.integer(raeduc_e),
                      PA_cat = as.integer(PA_cat),
                      MLTC = as.integer(MLTC),
                      painchr_ext = as.integer(painchr_ext)
                    )

```

## MICE imputation

Linear MixedModel-based FCS on https://arxiv.org/pdf/2404.06967

The FCS-2L approach imputes missing values using a series of conditional LMMs specified for each incomplete variable, tailored according to the distribution of the variable. 

The mice, micemd and miceadds R packages contain functions for implementing FCS-2L, identified by names starting with "2l" (2-level imputation).



In addition in our case we have a conditional variable, that is a variable having structural 
missingness that depend on the value of another variable: 

* Marital Strain (conditional on having a partner)


As with FCS-1L-wide method, the variables first need to be stored as correct type before 
generating the imputations, as described earlier. *The cluster variable must be of type integer*.

Therefore we configure the variable types so that each variable is imputed using an appropriate method: logistic regression for binary variables, ordered logistic regression for ordinal variables, and polytomous regression for unordered categorical data.

```{r}
# Check missingness by variable
vars_struct_missing<-c("ssupport6_log1p")
incomplete.vars_noStruct<-incomplete.vars[!incomplete.vars %in% vars_struct_missing]


missingness_by_var <- Missingdata_long %>%
  mutate(
    # Mark structural NAs that shouldn't count
    ssupport6_log1p_missing = ifelse(partner == 0, FALSE, is.na(ssupport6_log1p)),
#    mbmi_missing = ifelse(wave == 5, FALSE, is.na(mbmi)),
#    hscrp_missing = ifelse(wave == 5, FALSE, is.na(hscrp))
  ) %>%
  summarise(
    # For conditional variables - only count eligible cases
    ssupport6_log1p_pct = mean(ssupport6_log1p_missing[partner == 1], na.rm = TRUE) * 100,
#    mbmi_pct = mean(mbmi_missing[wave != 5], na.rm = TRUE) * 100,
#    hscrp_pct = mean(hscrp_missing[wave != 5], na.rm = TRUE) * 100,
    
    # For other variables - standard calculation
    across(all_of(incomplete.vars_noStruct), 
           ~mean(is.na(.)) * 100, .names = "{.col}_pct")
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missingness_by_var)

# Use the maximum missingness rate
max_missing <- max(missingness_by_var$pct_missing)
m_imputations <- ceiling(max_missing)

print(paste("Number of imputations:", m_imputations))




library(mice)
library(miceadds)
library(micemd)
# Define imputation methods
meth1 <- make.method(Missingdata_long)

meth1[complete.vars] <- ""  # No imputation for complete variables

## Time variant methods
Lv1_incomplete.vars<-incomplete.vars[incomplete.vars %in% lv1_vars]

incomplete_continuous<-cont_TimeVar.long[cont_TimeVar.long %in% Lv1_incomplete.vars]
meth1[incomplete_continuous] <- "2l.pmm" # Two-level predictive mean matching for incomplete continuous variables

incomplete_binary<-bin_TimeVar.long[bin_TimeVar.long %in% Lv1_incomplete.vars]

meth1[incomplete_binary] <- "2l.bin"  # Two-level logistic regression

## structural missingness will fail with 2l.xxx methods 
#meth1["mbmi"] <- "pmm"        # Single-level instead of 2l.pmm (wave 5 structural missing)
#meth1["hscrp"] <- "pmm"       # Single-level instead of 2l.pmm (wave 5 structural missing)
meth1["ssupport6_log1p"] <- "pmm"   # Single-level instead of 2l.pmm (conditional on partner)



## time invariant methods
Lv2_incomplete.vars<-incomplete.vars[incomplete.vars %in% lv2_vars]
meth1["raeduc_e"] <- "2lonly.pmm"   # Safer as raeduc_e has specific values 


# Define predictor Matrix
pred <- make.predictorMatrix(Missingdata_long)

# with "0" we tell which variables are not predictors
# leaving blank the row means all rows
#pred["predicted","predictor"]
pred[,c("idauniq")] <- 0

# time-fixed variable should not be imputed using time
# all time fixed variables are complete in this case
pred[lv2_vars, c("wave")] <- 0

# Specify alias as cluster variable
pred[, "idauniq"] <- -2  # Cluster grouping variable


# To specify random slopes for a predictor in a univariate imputation model for a certain variable, 
# number “2” should be included in the corresponding column variable for the row variable that 
# is being imputed. 

# Only use random slopes for variables with sufficient complete cases (20,000+)
# However, we need to prevent overfitting because with random slops we fit 2 parameters 
# per person (random intercept + random slope). With 8014 individuals we need to fit 16,028 random effect parameters, 
# but we only have 23,808 observations. This creates an overfitting situation.

# Apply random slopes selectively
#pred[c("ssupport6_log1p"),"wave"] <- 2 

### Special Exclusions ###
# Prevent painchr_ext from predicting painlv
pred["painchr_ext", "painlv"] <- 0  
pred["painlv", "painchr_ext"] <- 0  
# we do not use partner as predictor of spouse support because it is conditional on partner:
# we do not want to impute ssupport when it should be NA
pred["ssupport6_log1p", "partner"] <- 0


## this was added after I noticed that pmm struggled (numerical instability) with conditional variables as predictors given that these have systematic missingness
conditional_vars <- c("ssupport6_log1p")
pred[, conditional_vars] <- 0




#pred[ord_TimeVar.long, ] <- 0



################################################################################
##### CONDITIONAL IMPUTATION STEP 2 - "WHERE" MATRIX
#################################################################################


where_matrix <- make.where(Missingdata_long, "missing")

Missingdata_long <- Missingdata_long %>%
  mutate(
    # Set support to NA for non-partnered individuals
    ssupport6_log1p = ifelse(partner == 0, NA, ssupport6_log1p),
    # Set BMI and SCRP to NA for wave 5 because there was no nurse visit at that wave
#    mbmi  = ifelse(wave == 5, NA, mbmi),
#    hscrp = ifelse(wave == 5, NA, hscrp)
  )




# Modify where matrix to prevent imputation when partner=0
where_matrix[Missingdata_long$partner == 0 & is.na(Missingdata_long$ssupport6_log1p), "ssupport6_log1p"] <- FALSE
# Modify where matrix to prevent imputation when wave=5 from bmi and HSCRP
#where_matrix[Missingdata_long$wave == 5 & is.na(Missingdata_long$mbmi), "mbmi"] <- FALSE
#where_matrix[Missingdata_long$wave == 5 & is.na(Missingdata_long$hscrp), "hscrp"] <- FALSE



```

```{r, eval=FALSE}

set.seed(200720)
imp_1 = mice(data = Missingdata_long, 
            m = 12, 
            predictorMatrix = pred, 
            method = meth1, 
            where = where_matrix,
            diagnostics = TRUE,
            printFlag = TRUE,
            maxit = 10)
saveRDS(imp_1, paste0(paper1_rds_path,"mice_imputationlong_rev4.rds"))

```


```{r eval=T}
imp_1<-readRDS(paste0(paper1_rds_path,"mice_imputationlong_rev4.rds"))
summary(imp_1)
#saveRDS(imp, paste0(paper1_rds_path,"mice_imputation.rds"))
plot(imp_1)



# Extract one completed dataset
imputed_result <- complete(imp_1, 1)

Missingdata_long %>% vis_miss()
imputed_result %>% vis_miss()
# wave 4 ssupport missingness
imputed_result %>% filter(partner == 1) %>% select(ssupport6_log1p) %>% vis_miss
```
