---
title: "Change of Pain classes"
author: "Diego Vitali"
output: 
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
library(patchwork)
library(GGally)
library(ggdag)
library(mice)
library(kableExtra)


library(parallel)
library(future)
library(future.apply)
library(LMest)

source("Sources/Paper1_func.R")

paper1_rds_path<- "~/private_WP5/WP5_data/rds/"
paper1_MODELS_rds_path<-"~/private_WP5/WP5_data/model_fits/"

# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

#H_elsa_w4_6<-readRDS("c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/paper1/H_elsa_w4_6.rds")
#H_elsa_w1<-readRDS("c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/paper1/H_elsa_w1.rds")

## imputed dataset
updated_mids<-readRDS(paste0(paper1_rds_path,"mice_updatedMIDS_VH.rds"))

```

# MAIN EFFECT
##Predictors list

```{r echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

LTAdata_long <- mice::complete(updated_mids,1)

LTAdata_long<- LTAdata_long %>% mutate(
                                 ssupport6 = expm1(ssupport6_log1p),
                                 ssupport6_cat= case_when(
                                   partner == "No-partner" ~ "No-partner",     # smaller partner support scores = better support
                                   # "supportive partner" in the label means that partner support level was the same or better than the median
                                   ssupport6 <= 1.5 ~ "supportive partner",  
                                   # "less than supportive partner" in the label means that partner support level was worse than the median
                                   ssupport6 > 1.5 ~ "less than supportive partner",
                                   TRUE ~ NA_character_
                                 ) %>% factor(levels = c("No-partner", 
                                                         "less than supportive partner",
                                                         "supportive partner"))
                                 )



Variables<-c(
            # ID variable
               "idauniq",                # integer
            # TIME variable
               "time",                   # Integer: 1 = Wave 4, 2 = Wave 5, 3 = Wave 6
            # TIME INVARIANT PREDICTORS
               "raagey_z",               # Age stnd 
               "Sex_BIN",                # binary: Female, male
               "raeduc_e",
            # TIME-VARYING PREDICTORS
               "wealthQ_bin",           # binary: "Least deprived Qs"; "Most Deprived Q"
               "MLTC",                   # numeric (count)
               "cesd_lat_shifted",       # numeric Depression score
               "loneliness_group",              # binary: "Low, high loneliness"
               "partner",
               "ssupport6_cat",          # categorical (no partner, high support, low support)
            # OUTCOME'S INDICATORS
               "painchr_ext",            # Integer (ordered)
               "lvimp_shifted_cat3"#,     # Integer (ordered) This is 3 level categorical impact 0=func, 1=!HI, 2= HI 
            )

LTAdata_long<-LTAdata_long %>% select(all_of(Variables))

```

## Binary predictor imbalance check

Statistical software often selects the reference category alphabetically. This is not a problem when categories are balanced or we do not have information that identifies a group of answers as normative. The latter is the case for Wealth Quintile — the most logical comparison is against the least deprived group. Extreme imbalance in binary predictors (e.g. RACE_BIN ~97%/3%) can cause model separation; these variables should be flagged for possible instability.

```{r echo=FALSE}
binary_factors <- names(LTAdata_long)[sapply(LTAdata_long, function(x) is.factor(x) && nlevels(x) == 2)]

imbalance_data <- data.frame(
  variable    = character(),
  level1      = character(),
  level2      = character(),
  prop_level1 = numeric(),
  prop_level2 = numeric(),
  imbalance   = numeric(),
  stringsAsFactors = FALSE
)

for(var in binary_factors) {
  props      <- prop.table(table(LTAdata_long[[var]]))
  levels_var <- names(props)
  imbalance_data <- rbind(imbalance_data, data.frame(
    variable    = var,
    level1      = levels_var[1],
    level2      = levels_var[2],
    prop_level1 = round(props[1], 2),
    prop_level2 = round(props[2], 2),
    imbalance   = round(abs(props[1] - props[2]), 2),
    stringsAsFactors = FALSE
  ))
}
imbalance_data <- imbalance_data[order(imbalance_data$imbalance, decreasing = TRUE), ]
kable(imbalance_data, row.names = FALSE,
      caption = "Binary variables ordered by imbalance (most unbalanced first)")
```

# Partnership quality — LTA Model Specification

In this Latent Transition Analysis (LTA) we apply a Latent Markov model with time-varying categorical indicators to empirically determine latent Markov states, together with the probability of subjects to move between these unobserved states over time. The model captures both the initial state probability distribution and the probability of temporal transitions between latent states. Both components are parametrised with multinomial logistic regression via which we determine the effects of our covariates of interest.

## Confounding structure

### Initial state probabilities

**Exposure:** Partnership status — three levels: no partner; less than supportive partner; supportive partner.

**Confounders (TE model):** Sex, Age, Education level, Wealth (most deprived quintile), MLTC.

**Additional covariates (DE model):** Depression (cesd_lat_shifted) and Loneliness — included to separate the direct effect of partnership from the portion operating through emotional wellbeing (see DE model rationale below).

### Transition probabilities

**Time-invariant confounders:** Sex, Age, Education.

**Time-varying confounders:** Wealth, MLTC.

**Time-varying confounders/mediators (DE only):** Depression, Loneliness.

## TE DE config

```{r}
### LTA Formulas

latent_state_vars<-c("lvimp_shifted_cat3", #Y1
                     "painchr_ext"         #Y2
)


# TE_initial_prob_vars
TE_initial_prob_vars<- c(
                  "Sex_BIN",
                  "raeduc_e",
                  "wealthQ_bin",
                  "ssupport6_cat",
                  "MLTC",
                  "raagey_z"
                  )

TE_trans_prob_vars<- c(TE_initial_prob_vars)

# DE_initial_prob_vars: 
DE_vars<-c(TE_initial_prob_vars,"cesd_lat_shifted","loneliness_group")
DE_vars_sub<-c(TE_initial_prob_vars,"cesd_lat_shifted")


###  TE formula
# Total Effects (TE) model: estimates the combined effects of partnership status
# and key psychosocial/demographic variables (sex, education, wealth, MLTC, age)
# on both initial state membership and transitions between latent states.
# Treats included predictors as potential total effects — pathways via depression
# and loneliness are absorbed into the partnership status coefficient.
fmLatent_TE <- lmestFormula(data = LTAdata_long, response = latent_state_vars,
                                LatentInitial = TE_initial_prob_vars,
                                LatentTransition =TE_trans_prob_vars
)


###  DE formula
# Direct Effects (DE) model: extends TE by additionally including depression
# (cesd_lat_shifted) and loneliness as covariates. This adjustment reflects the
# hypothesis that partnership status contributes indirectly to health transitions
# through downstream effects on emotional wellbeing. By conditioning on these
# pathways, DE disaggregates direct effects from those potentially mediated by
# depression and loneliness.
#
# DE model rationale: partnership quality and emotional wellbeing are persistently
# co-occurring throughout waves 4-6 (loneliness: Cramér's V = 0.291; depression:
# η² = 0.074, stable gradient across all three waves). This justifies asking
# "how much of the partnership → pain state association operates through, or
# alongside, depression and loneliness?"
fmLatent_DE <- lmestFormula(data = LTAdata_long, response = latent_state_vars,
                                LatentInitial    = DE_vars,
                                LatentTransition = DE_vars
)


```

### Sensitivity analysis: partnership quality × depression and loneliness

The DE model rationale requires that partnership quality and emotional wellbeing are
meaningfully associated. The analysis below quantifies this at baseline and checks
whether the gradient is stable across all three waves (justifying treatment as
time-varying confounders/co-occurring variables rather than a baseline-only artefact).

```{r de-sensitivity-partnership-wellbeing}
library(vcd)      # assocstats() for Cramér's V
library(rstatix)  # eta_squared()

baseline <- LTAdata_long %>% filter(time == 1)

# ── 1. Loneliness × partnership status (both categorical) ─────────────────
tab_lon <- table(baseline$loneliness_group, baseline$ssupport6_cat)

cat("=== Loneliness × partnership: column proportions ===\n")
print(round(prop.table(tab_lon, margin = 2), 3))

cat("\nCramér's V:\n")
print(assocstats(tab_lon))

# ── 2. Depression × partnership status (continuous × 3-level factor) ──────
cat("\n=== Depression × partnership: means by group ===\n")
baseline %>%
  group_by(ssupport6_cat) %>%
  summarise(
    n         = n(),
    mean_cesd = round(mean(cesd_lat_shifted, na.rm = TRUE), 3),
    sd_cesd   = round(sd(cesd_lat_shifted, na.rm = TRUE), 3),
    .groups   = "drop"
  ) %>%
  print()

aov_dep <- aov(cesd_lat_shifted ~ ssupport6_cat, data = baseline)
cat("\nOne-way ANOVA:\n")
print(summary(aov_dep))
cat("\nη²:\n")
print(eta_squared(aov_dep))

# ── 3. Longitudinal stability check: same gradient across waves 4–6 ───────
cat("\n=== Wave-by-wave stability ===\n")
LTAdata_long %>%
  group_by(time, ssupport6_cat) %>%
  summarise(
    mean_cesd        = round(mean(cesd_lat_shifted, na.rm = TRUE), 3),
    prop_high_lonely = round(mean(loneliness_group == "High", na.rm = TRUE), 3),
    .groups          = "drop"
  ) %>%
  arrange(ssupport6_cat, time) %>%
  print()
```

**Expected results (pre-computed in archive/LTA_pull_res_tables.r):**

- Loneliness × partnership: Cramér's V = 0.291 — moderate, monotonic gradient
  (supportive: ~36% high loneliness; less-than-supportive: ~61%; no-partner: ~69%)
- Depression × partnership: η² = 0.074 (above Cohen's medium threshold of 0.06),
  F(2, 8011) = 320.2, p < 2×10⁻¹⁶ — monotonic gradient
  (supportive mean CES-D ≈ 0.64; less-than-supportive ≈ 0.94; no-partner ≈ 1.13)
- Both gradients are stable across waves 4–6 (supportive range 0.61–0.64;
  no-partner range 1.13–1.14), confirming this is not a cross-sectional artefact.

Both associations are moderate in magnitude but not collinear. The consistent
monotonic gradient in the expected direction — better partnership quality → less
loneliness, less depression — supports including these variables in the DE model to
isolate the direct partnership effect from the portion operating through emotional
wellbeing.



```{r}

model_configs <- list(
     Total_Effects_fit = list(
       name = "TE",
       description = "TE model",
       responsesFormula = fmLatent_TE$responsesFormula,
       latentFormula = fmLatent_TE$latentFormula
       ),
       Direct_Effects_fit = list(
           name = "DE",
           description = "DE model",
           responsesFormula = fmLatent_DE$responsesFormula,
           latentFormula = fmLatent_DE$latentFormula
       )
     )


```

# Latent states determination (k = 4)

We expect to find respondents with significantly impaired functioning (high health impact) who have chronic pain, but also respondents with similar impairment without chronic pain. We expect respondents without physical functioning issues or pain, and respondents who may report acute pain without high levels of health impact. This should amount to k = 4 states.

As per Moore et al. (2025), we run all possible k+1 models until reaching consistent nonconvergence and compare how well each solution represents the data. `lmestSearch()` addresses both model selection and the multimodality of the likelihood function by employing different initializations to search for the global maximum.

Two main criteria guide selection of k: AIC and BIC. BIC is generally preferred in longitudinal latent class modeling for its stronger penalty on model complexity.

```{r eval=FALSE}
Lmest_search <- lmestSearch(responsesFormula = fmLatent_TE$responsesFormula,
                   index = c("idauniq", "time"),
                   data  = LTAdata_long,
                   version = "categorical",
                   k = 1:5,
                   modBasic = 1,
                   nrep = 400,
                   seed = 200720)
timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")
saveRDS(Lmest_search, paste0(paper1_MODELS_rds_path, "K-State_parametrisation_n400_", timestamp_suffix, ".rds"))
saveRDS(Lmest_search, paste0(paper1_MODELS_rds_path, "K-State_parametrisation_n400.rds"))
```

```{r echo= FALSE}
Lmest_search <- readRDS(paste0(paper1_MODELS_rds_path, "K-State_parametrisation_n400.rds"))
Lmest_search %>% summary()
Lmest_search$out.single[[4]] %>% plot(what = "CondProb")

state5_llDiffs <- lmestSearch_plot(Lmest_search$out.single, k = 5, k_multiplier = 3, plot_hist = TRUE)
state4_llDiffs <- lmestSearch_plot(Lmest_search$out.single, k = 4, k_multiplier = 3, plot_hist = TRUE)
state3_llDiffs <- lmestSearch_plot(Lmest_search$out.single, k = 3, k_multiplier = 3, plot_hist = TRUE)
state2_llDiffs <- lmestSearch_plot(Lmest_search$out.single, k = 2, k_multiplier = 3, plot_hist = TRUE)

cat("Median absolute diff near max for k=3:", abs(state3_llDiffs$median_diff), "\n")
cat("Median absolute diff near max for k=4:", abs(state4_llDiffs$median_diff), "\n")
cat("Median absolute diff near max for k=5:", abs(state5_llDiffs$median_diff), "\n")
```

The log-likelihood difference histograms describe convergence behavior across candidate k values:

- **k = 3**: Most runs find the global maximum or group in widely separated local optima (high median diff, multimodal plots). Poor stability.
- **k = 5**: Substantial variability and multiple peaks. High median absolute difference. Less reliable estimation.
- **k = 4**: Tightly clustered histogram near the global maximum with very low median absolute difference (~1.30). ~72% of starts converge near the global maximum (within tol = median +3 MAD). Strong stability for a complex model.

**WE ACCEPT k = 4.**

# LMest configuration

Key parameters for `lmest()`:

- **`modBasic = 1`**: Assumes the transition process is stationary across the two inter-wave intervals — i.e. the probabilities governing state-to-state moves are the same at waves 4→5 and 5→6. This is justified by the structural equivalence of the two intervals (identical measurement instrument, ~2 year gap each) and the absence of any known mechanism that would alter the pain transition process within this narrow observation window. `modBasic = 0` would allow time-heterogeneous transitions but adds 16 free parameters (a second full transition matrix) without substantive basis.
- **`start = 1`**: Random starting values. `start = 0` = deterministic only; `start = 2` = user-supplied initial values (used for rebasing).
- **`ntry`**: Number of random initializations. With `k = 4` and `ntry = 5`, each `lmest()` call runs 1 deterministic + 5×4 = 20 random starts.
- **`fort = TRUE`**: Use Fortran routines where available (faster).
- **`maxit = 10000`, `tol = 1e-8`**: Convergence criteria.

As we include covariates we have both a response formula and a latent formula: our model estimates conditional transitions, i.e. transition probabilities that depend on covariate values and therefore allow different individual characteristics to yield different transition probabilities.

## Multi-start methodology

*"A unique aspect of all finite mixture models is that although there is a global, best-fitting solution at the global maximum log likelihood (LL), there are also multiple less well-fitting solutions at different local maxima LL. Part of the exploratory nature of finite mixture models is running each K+1 class model with multiple random starts so that each model run includes multiple model solutions with their LL reported in the output."* [@moore2025introduction]

The estimation procedure works in three steps:

1. **Deterministic start** (`start = 0`): A single run iterates to a stable LL, serving as a baseline.
2. **Random starts** (`start = 1`): LMest scales the number of random starts by k (ntry × k). Each start prints its LL trajectory and convergence metrics.
3. **Selection**: After all starts, LMest takes the converged solution with the highest final log-likelihood as model output.

Using `run_lmest_parallel_seeds()` we parallelise across `n_reps` workers, each running a full `lmest()` call. Total random starts = n_reps × ntry × k. The function returns the best model across all workers plus a convergence histogram for visual assessment of LL replication.

**Convergence criterion:** If the top LL is replicated multiple times within a small tolerance (≤ 1e-4), we have strong evidence of proximity to the global maximum.

- TE model: n_reps = 36, ntry = 5, k = 4 → 720 total random starts
- DE model: n_reps = 24, ntry = 5, k = 4 → 480 total random starts (pilot confirmed 100% stability at 400)

```{r}
############################################################################
##  PARALLELISATION FOR LMEST — run_lmest_parallel_seeds() defaults
############################################################################
# modBasic = 1, start = 1, maxit = 10000, k = 4, tol = 1e-8
# fort = TRUE, out_SE = FALSE, base_seed = 200720
# exclude_deterministic = TRUE (removes the 1 deterministic start per rep
#   from convergence diagnostics so histogram reflects only random starts)
```

# Run Total Effects
```{r eval=FALSE}
TE_results_parallel_output <- run_lmest_parallel_seeds(
  config = model_configs[[1]], # first config in list does Total effects
  data = LTAdata_long,
  n_reps = 36,      # 720 reps
  ntry = 5, # random starts are k * ntry
  n_cores = 24      
)

TE_results_parallel_output$plot
TotalEffects<-TE_results_parallel_output$best_model

timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")

saveRDS(TotalEffects,paste0(paper1_MODELS_rds_path,"TotalEffects_",timestamp_suffix,".rds"))
saveRDS(TotalEffects,paste0(paper1_MODELS_rds_path,"TotalEffects.rds"))

saveRDS(TE_results_parallel_output,paste0(paper1_MODELS_rds_path,"TE_results_parallel_output_",timestamp_suffix,".rds"))
saveRDS(TE_results_parallel_output,paste0(paper1_MODELS_rds_path,"TE_results_parallel_output.rds"))




```

# Run Direct Effects
```{r eval=FALSE}
DE_results_parallel_output <- run_lmest_parallel_seeds(
   config = model_configs[[2]],
   data = LTAdata_long,
   n_reps = 24,      
   ntry = 5,
   n_cores = 24      
)

DirectEffects<-DE_results_parallel_output$best_model

timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")

saveRDS(DirectEffects,paste0(paper1_MODELS_rds_path,"DirectEffects_",timestamp_suffix,".rds"))
saveRDS(DirectEffects,paste0(paper1_MODELS_rds_path,"DirectEffects.rds"))

saveRDS(DE_results_parallel_output,paste0(paper1_MODELS_rds_path,"DE_results_parallel_output_",timestamp_suffix,".rds"))
saveRDS(DE_results_parallel_output,paste0(paper1_MODELS_rds_path,"DE_results_parallel_output.rds"))


```


## compare DE TE models (lk)

```{r}
# Load fitted models from disk
TotalEffects  <- readRDS(paste0(paper1_MODELS_rds_path, "TotalEffects.rds"))
DirectEffects <- readRDS(paste0(paper1_MODELS_rds_path, "DirectEffects.rds"))
```

## State ordering and rebasing

The order of latent states produced by the LM model is arbitrary. What is labeled State 1 in the TE model may be identified as State 3 in the DE model. However, TE and DE models are expected to produce very similar conditional response probability profiles across states — the configuration of observed indicators within each state should be nearly identical between models.

We reorder states using a **weight-based composite score** derived from the two indicator expected values. For each state $s$ and indicator $i$, we compute the expected value as a weighted sum of category labels by their conditional response probabilities ($\Psi$):

$$\text{composite}_s = 0.6 \times E[\text{lvimp\_shifted\_cat3} \mid s] + 0.4 \times E[\text{painchr\_ext} \mid s]$$

States are then ranked `order(composite_score)`: rank 1 = healthiest, rank 4 = HICP. The 0.6/0.4 split reflects that `lvimp` has a wider range (0–2) than the binary `painchr_ext` (0–1); equal weights would let the wider-range indicator dominate. Weights need only be consistent across imputations, not precisely calibrated.

To re-parameterize after reordering, we use `start = 2` with the reordered Be/Ga/Psi matrices as initial values. This is preferable to purely algebraic rebasing because it lets lmest confirm convergence and compute standard errors on the reordered solution in the usual way.

**Be rebasing (initial probabilities):** lmest models log-odds of each state relative to State 1 (reference). To shift the reference to the new State 1 (previously State $r$):

$$\log\frac{P(\text{State}_j)}{P(\text{State}_r)} = \log\frac{P(\text{State}_j)}{P(\text{State}_1)} - \log\frac{P(\text{State}_r)}{P(\text{State}_1)} = \beta_j - \beta_r$$

**Ga rebasing (transitions):** Ga has dimensions covariates × (k-1) contrasts × k starting states. For each starting state, we expand to a full $k \times k$ array (inserting a zero column for the same-state reference), reorder both the contrast dimension and the starting-state dimension by `state_order`, then subtract the new baseline column.

```{r}
ordering_indicators <- c("lvimp_shifted_cat3", "painchr_ext")
ordering_weights    <- c(0.6, 0.4)

state_order_TE <- Determine_state_ordering(
  TotalEffects,
  indicator_names   = ordering_indicators,
  indicator_weights = ordering_weights
)$severity_order

state_order_DE <- Determine_state_ordering(
  DirectEffects,
  indicator_names   = ordering_indicators,
  indicator_weights = ordering_weights
)$severity_order

cat("TE state order:", state_order_TE, "\n")
cat("DE state order:", state_order_DE, "\n")

# Sanity check: conditional probability plots before rebasing
plot(TotalEffects,  what = "CondProb")
mtext(paste("TE — raw order; severity order:", paste(state_order_TE, collapse = " ")),
      side = 3, line = -8, adj = 1)

plot(DirectEffects, what = "CondProb")
mtext(paste("DE — raw order; severity order:", paste(state_order_DE, collapse = " ")),
      side = 3, line = -8, adj = 1)
```

```{r}
# Rebase: refit with start=2 using reordered Be/Ga/Psi as initial values
# out_SE=TRUE computes standard errors on the rebased solution
TotalEffects_rebased <- rebase_and_refit_lmest(
  fit_model        = TotalEffects,
  state_order      = state_order_TE,
  latent_formula   = fmLatent_TE$latentFormula,
  response_formula = fmLatent_TE$responsesFormula,
  out_SE           = TRUE,
  plot             = TRUE
)

DirectEffects_rebased <- rebase_and_refit_lmest(
  fit_model        = DirectEffects,
  state_order      = state_order_DE,
  latent_formula   = fmLatent_DE$latentFormula,
  response_formula = fmLatent_DE$responsesFormula,
  out_SE           = TRUE,
  plot             = TRUE
)
```

## Model diagnostics

```{r model-diagnostics}
# compute_diagnostics() is defined in Sources/Paper1_func.R.
# Implements all four Nagin (2010) criteria plus entropy (Clark & Muthén 2009).
# Returns a named list: $entropy, $app, $class_err, $occ — one tibble each.
# Pass metrics = c("entropy","app") etc. to print a subset only.

diag_TE <- compute_diagnostics(TotalEffects_rebased, "TE")
diag_DE <- compute_diagnostics(DirectEffects_rebased, "DE")
```

## Results extraction

```{r}
TE_summaries <- regression_summaries(Model_fit = TotalEffects_rebased)
DE_summaries <- regression_summaries(Model_fit = DirectEffects_rebased)

# Initial state probabilities
kable(TE_summaries$initial_states %>% arrange(State, varname),
      caption = "TE model — initial state probability coefficients (multinomial logit, ref = State 1)")

kable(DE_summaries$initial_states %>% arrange(State, varname),
      caption = "DE model — initial state probability coefficients")

# Transition probabilities
kable(TE_summaries$transitions %>% arrange(Transition, varname),
      caption = "TE model — transition probability coefficients")

kable(DE_summaries$transitions %>% arrange(Transition, varname),
      caption = "DE model — transition probability coefficients")
```

```{r}
# DE is the more complex (full) model; TE is the restricted model
# LRT tests whether adding depression + loneliness to TE significantly improves fit
LR_TE_DE <- 2 * (DirectEffects$lk - TotalEffects$lk)
df_TE_DE <- DirectEffects$np - TotalEffects$np
p_TE_DE  <- pchisq(LR_TE_DE, df = df_TE_DE, lower.tail = FALSE)

cat(sprintf("DE vs TE: χ²(df=%d) = %.2f, p = %.4g\n", df_TE_DE, LR_TE_DE, p_TE_DE))

TE_DE_vals <- data.frame(
  Model = c("TE model", "DE model"),
  lK    = c(TotalEffects$lk,  DirectEffects$lk),
  AIC   = c(TotalEffects$aic, DirectEffects$aic),
  BIC   = c(TotalEffects$bic, DirectEffects$bic)
)
TE_DE_vals$DeltaLk  <- 2 * (TE_DE_vals$lK  - min(TE_DE_vals$lK,  na.rm = TRUE))
TE_DE_vals$DeltaAIC <- TE_DE_vals$AIC - min(TE_DE_vals$AIC, na.rm = TRUE)
TE_DE_vals$DeltaBIC <- TE_DE_vals$BIC - min(TE_DE_vals$BIC, na.rm = TRUE)

kable(TE_DE_vals[order(TE_DE_vals$AIC), ], row.names = FALSE,
      caption = "TE vs DE model comparison — DE adds depression and loneliness to TE specification")
```

TE estimates total effects of partnership status (including pathways via depression and loneliness). DE estimates direct effects (conditional on depression and loneliness), so we expect partnership status coefficients to attenuate relative to TE.

A large positive LRT statistic (DE vs TE) indicates the DE model significantly improves fit over TE. Improved AIC/BIC under DE supports depression and loneliness being strongly related to the latent state process as confounders and/or mediators. This does not invalidate the TE model — it confirms that part of the TE association operates through depression and loneliness, consistent with theory.

# Sensitivity analysis: nested models without exposure block

The noExp models remove ssupport6_cat from the TE specification to test whether
the exposure block adds significant explanatory power. Two nested models:
   - TE_noEXP_init:  ssupport6_cat removed from INITIAL probabilities only
   - TE_noEXP_trans: ssupport6_cat removed from TRANSITION probabilities only

LRT (TE vs TE_noEXP) tests the null that partnership status has no effect on
initial state membership / transitions respectively.

Run on a single imputation (imp=1, same LTAdata_long as main models).
Single-imputation LRT is sufficient for sensitivity purposes — the exposure
block significance is expected to be robust across imputations.

```{r}
################################################################################
##################### REDUCED MODELS excluding the exposure
################################################################################


# 1 # LTA TE NO-EXP FORMULAS
fmLatent_TE_noExpBlock_init <- lmestFormula(data = LTAdata_long, response = latent_state_vars, 
                          LatentInitial   =  TE_initial_prob_vars %>% setdiff(c("ssupport6_cat")),
                          LatentTransition = TE_trans_prob_vars   # we keep transition formula unchanged so that this model is nested in the previous.
                         )

fmLatent_TE_noExpBlock_trans <- lmestFormula(data = LTAdata_long, response = latent_state_vars, 
                          LatentInitial    = TE_initial_prob_vars, # we keep Initial prob formula unchanged so that this model is nested in the previous. 
                          LatentTransition = TE_trans_prob_vars %>% setdiff(c("ssupport6_cat"))
                         )

# 2 # LTA TE NO-EXP CONFIG
model_configs_noEXP <- list(
     NO_Effects_init_fit = list(
       name = "TE-noEXPinit",
       description = "TE-noEXP init model",
       responsesFormula = fmLatent_TE_noExpBlock_init$responsesFormula,
       latentFormula = fmLatent_TE_noExpBlock_init$latentFormula
       ),
       NO_Effects_trans_fit = list(
           name = "TE-noEXPtrans",
           description = "TE-noEXP trans model",
           responsesFormula = fmLatent_TE_noExpBlock_trans$responsesFormula,
           latentFormula = fmLatent_TE_noExpBlock_trans$latentFormula
       )
     )
```



```{r eval=FALSE}
# 3 # LTA TE NO-EXP INIT parallel run
TE_init_noEXP_results_parallel_output <- run_lmest_parallel_seeds(
  config  = model_configs_noEXP[[1]],
  data    = LTAdata_long,
  n_reps  = 20,
  ntry    = 5,
  n_cores = 20,
  k = 4
)

TE_init_noEXP_results_parallel_output$plot
TotalEffects_init_noEXP <- TE_init_noEXP_results_parallel_output$best_model

timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")
saveRDS(TotalEffects_init_noEXP, paste0(paper1_MODELS_rds_path, "TotalEffects_init_noEXP_", timestamp_suffix, ".rds"))
saveRDS(TotalEffects_init_noEXP, paste0(paper1_MODELS_rds_path, "TotalEffects_init_noEXP.rds"))
saveRDS(TE_init_noEXP_results_parallel_output, paste0(paper1_MODELS_rds_path, "TE_init_noEXP_results_parallel_output.rds"))

# 4 # LTA TE NO-EXP TRANS parallel run
TE_trans_noEXP_results_parallel_output <- run_lmest_parallel_seeds(
  config  = model_configs_noEXP[[2]],
  data    = LTAdata_long,
  n_reps  = 20,
  ntry    = 5,
  n_cores = 20
)

TE_trans_noEXP_results_parallel_output$plot
TotalEffects_trans_noEXP <- TE_trans_noEXP_results_parallel_output$best_model

timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")
saveRDS(TotalEffects_trans_noEXP, paste0(paper1_MODELS_rds_path, "TotalEffects_trans_noEXP_", timestamp_suffix, ".rds"))
saveRDS(TotalEffects_trans_noEXP, paste0(paper1_MODELS_rds_path, "TotalEffects_trans_noEXP.rds"))
saveRDS(TE_trans_noEXP_results_parallel_output, paste0(paper1_MODELS_rds_path, "TE_trans_noEXP_results_parallel_output.rds"))
```

# 5 # LRT: TE vs noEXP models
```{r}
TotalEffects_init_noEXP  <- readRDS(paste0(paper1_MODELS_rds_path, "TotalEffects_init_noEXP.rds"))
TotalEffects_trans_noEXP <- readRDS(paste0(paper1_MODELS_rds_path, "TotalEffects_trans_noEXP.rds"))

# LRT: TE vs TE_noEXP_init — tests whether ssupport6_cat affects INITIAL state membership
LR_init  <- 2 * (TotalEffects$lk - TotalEffects_init_noEXP$lk)
df_init  <- TotalEffects$np - TotalEffects_init_noEXP$np
p_init   <- pchisq(LR_init, df = df_init, lower.tail = FALSE)

# LRT: TE vs TE_noEXP_trans — tests whether ssupport6_cat affects TRANSITIONS
LR_trans <- 2 * (TotalEffects$lk - TotalEffects_trans_noEXP$lk)
df_trans <- TotalEffects$np - TotalEffects_trans_noEXP$np
p_trans  <- pchisq(LR_trans, df = df_trans, lower.tail = FALSE)

cat(sprintf("TE vs TE_noEXP_init:  χ²(df=%d) = %.2f, p = %.4g\n", df_init,  LR_init,  p_init))
cat(sprintf("TE vs TE_noEXP_trans: χ²(df=%d) = %.2f, p = %.4g\n", df_trans, LR_trans, p_trans))

noexp_vals <- data.frame(
  Model    = c("TE (full)", "TE — no exposure (initial)", "TE — no exposure (transition)"),
  lK       = c(TotalEffects$lk,  TotalEffects_init_noEXP$lk,  TotalEffects_trans_noEXP$lk),
  AIC      = c(TotalEffects$aic, TotalEffects_init_noEXP$aic, TotalEffects_trans_noEXP$aic),
  BIC      = c(TotalEffects$bic, TotalEffects_init_noEXP$bic, TotalEffects_trans_noEXP$bic),
  LRT_chi2 = c(NA, round(LR_init, 2),  round(LR_trans, 2)),
  df       = c(NA, df_init,  df_trans),
  p        = c(NA, round(p_init, 4), round(p_trans, 4))
)

kable(noexp_vals, row.names = FALSE,
      caption = "Sensitivity analysis: LRT testing significance of partnership status exposure block (single imputation)")
```


