---
title: "Data preparation"
author: "Diego Vitali"
output: 
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib

---
```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(lavaan)
library(semTools)
library(tidyr)
library(dplyr)
library(ggplot2)
library(naniar)
library(NbClust)
library(readr)
library(polycor)
library(patchwork)
library(corrplot)
library(GGally)
source("Sources/Paper1_func.R")

#ELSA_tab_path<-"c:/Users/d.vitali/OneDrive - University College London/CRIISP/00_WP5/00_WP5_data/5050tab/UKDA-5050-tab/tab/"
ELSA_tab_path<-"~/private/WP5_data/tab/"

#clean_tab_path<-"c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/"
clean_tab_path<-"~/private/WP5_data/"

#paper1_rds_path<-"c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/data/clean/paper1/"
paper1_rds_path<-"~/private/WP5_data/rds/"

#paper1_MODELS_rds_path<-"c:/Users/d.vitali/Desktop/Github/CRIISP-WP5/LatGrCrv_mod/Pain - trajectory profiling/model_fits/"
paper1_MODELS_rds_path<-"~/private/WP5_data/model_fits/"


# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

```


# IMPORT impact IRT data
```{r eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
H_elsa_w1_9<-read.csv(paste0(clean_tab_path,"Hg3_w1_9_ALL.tab"), header=T)

################################################################################
## IMPACT ESTIMATES  (Fscores from IRT)
################################################################################
impactw2_8<-readRDS(paste0(paper1_rds_path,"impactEstimates2-8_partialScalar.rds")) 

## merging impact estimates DF with main DF
H_elsa_w1_9<-merge(impactw2_8,H_elsa_w1_9, by = "idauniq", all = TRUE)
```

# IMPORT ELSA data
```{r eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
################################################################################
## PAIN DURATION VARIABLE (hepag)
################################################################################
# HEPAG
# How long has the pain been bothering you? ((if pain AT LEAST moderate or severe))
#	Value = 1.0	Label = ...less than 3 months,
              #	Value = 2.0	Label = more than 3 months, but less than 6 months,
              #	Value = 3.0	Label = more than 6 months, but less than 12 months,
              #	Value = 4.0	Label = or more than 12 months?
              #	Value = -1.0	Label = Item not applicable
              #	Value = -9.0	Label = Refusal
              #	Value = -8.0	Label = Don't Know
              #	Value = -2.0	Label = Schedule not applicable
    #        "hepaj", # Are you currently receiving any treatment for your pain? 
    #         1:yes;2:no
    
    #       "hepak3" # How well does the treatment control your pain? 
    #         1:very well; 2:fairly well; 3:not very well; 4:not at all;

loc_varnames<-c("hepawba", # *Whether feel pain in back*
                "hepawhi", # *Whether feel pain in hips*
                "hepawkn", # *Whether feel pain in knees*
                "hepawfe", # *Whether feel pain in feet*
                "hepawmo", # *Whether feel pain in mouth or teeth*
                "hepawot", # *Whether feel pain elsewhere*
                "hepawal" # *Whether feel pain all over*
                )



# importing the wave 4 variables that are not in the harmonised Dataset
elsa_w4_core <- read_delim(paste0(ELSA_tab_path,"wave_4_elsa_data_v3.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", "hepag","hepaj",loc_varnames)))

elsa_w5_core <- read_delim(paste0(ELSA_tab_path,"wave_5_elsa_data_v4.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", "hepag","hepaj",loc_varnames)))

elsa_w6_core <- read_delim(paste0(ELSA_tab_path,"wave_6_elsa_data_v2.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", "HePag","HePaj",loc_varnames)))


#renaming pain duration variables to suit Harmonized data

colnames(elsa_w4_core)<-c("idauniq", "r4hepag","r4hepaj",paste0("r4",loc_varnames))
colnames(elsa_w5_core)<-c("idauniq", "r5hepag","r5hepaj",paste0("r5",loc_varnames))
colnames(elsa_w6_core)<-c("idauniq", "r6hepag","r6hepaj",paste0("r6",loc_varnames))

H_elsa_w1_9<-merge(elsa_w4_core,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w5_core,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w6_core,H_elsa_w1_9, by = "idauniq", all = TRUE)


################################################################################
## C-Reactive Proteine var from Nurse data
################################################################################

## CRP
#var: hscrp
# Value = -11.0	Label = Blood sample not taken
#	Value = -1.0	Label = Sample not received
#	Value = 99998.0	Label = Refusal
#	Value = -9.0	Label = TEXT
#	Value = -8.0	Label = Possible presence of variant haemoglobin
#	Value = -7.0	Label = Sample unsuitable for analysis (variety of causes)
#	Value = -6.0	Label = Sample took more than five days to reach the lab so is unsuitable for analysis
#	Value = -5.0	Label = Indicates that triglyceride measurement is unreliable
#	Value = -4.0	Label = Indicates that total cholesterol measurement is unreliable
#	Value = -3.0	Label = Sample received but was haemolysed and so not suitable for analysis
#	Value = -2.0	Label = Sample received but insufficient blood for analysis
#	Value = 99999.0	Label = Don't Know

elsa_w4_nurse <- read_delim(paste0(ELSA_tab_path,"wave_4_nurse_data.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", "hscrp")))
elsa_w6_nurse <- read_delim(paste0(ELSA_tab_path,"wave_6_elsa_nurse_data_v2.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", "hscrp")))

#renaming pain duration variables to suit Harmonized data

colnames(elsa_w4_nurse)<-c("idauniq", "r4hscrp")
colnames(elsa_w6_nurse)<-c("idauniq", "r6hscrp")

H_elsa_w1_9<-merge(elsa_w4_nurse,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w6_nurse,H_elsa_w1_9, by = "idauniq", all = TRUE)


################################################################################
## FINANCIALLY DERIVED VARIABLES
################################################################################
elsa_w1_fin <- read_delim(paste0(ELSA_tab_path,"wave_1_financial_derived_variables.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", 
                                         "nettotw_bu_s",  # net total wealth
                                         "totwq5_bu_s" )))%>%# net total wealth quintiles
                                                  mutate(
                                                    nettotw_bu_s = as.integer(nettotw_bu_s),
                                                    totwq5_bu_s = as.integer(totwq5_bu_s))

elsa_w4_fin <- read_delim(paste0(ELSA_tab_path,"wave_4_financial_derived_variables.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", 
                                         "nettotw_bu_s",  # net total wealth
                                         "totwq5_bu_s" ))) %>%# net total wealth quintiles
                                                  mutate(
                                                    nettotw_bu_s = as.integer(nettotw_bu_s),
                                                    totwq5_bu_s = as.integer(totwq5_bu_s))

elsa_w5_fin <- read_delim(paste0(ELSA_tab_path,"wave_5_financial_derived_variables.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", 
                                         "nettotw_bu_s",  # net total wealth
                                         "totwq5_bu_s")))%>% # net total wealth quintiles
                                                  mutate(
                                                    nettotw_bu_s = as.integer(nettotw_bu_s),
                                                    totwq5_bu_s = as.integer(totwq5_bu_s))

elsa_w6_fin<- read_delim(paste0(ELSA_tab_path,"wave_6_financial_derived_variables.tab"), 
                          delim = "\t",
                          col_select = all_of(c("idauniq", 
                                         "nettotw_bu_s",  # net total wealth
                                         "totwq5_bu_s"))) %>% # net total wealth quintiles
                                                  mutate(
                                                    nettotw_bu_s = as.integer(nettotw_bu_s),
                                                    totwq5_bu_s = as.integer(totwq5_bu_s))


#renaming variables to suit Harmonized data



colnames(elsa_w1_fin)<-c("idauniq", "r1nettotw_bu_s","r1totwq5_bu_s")
colnames(elsa_w4_fin)<-c("idauniq", "r4nettotw_bu_s","r4totwq5_bu_s")
colnames(elsa_w5_fin)<-c("idauniq", "r5nettotw_bu_s","r5totwq5_bu_s")
colnames(elsa_w6_fin)<-c("idauniq", "r6nettotw_bu_s","r6totwq5_bu_s")


## merging
H_elsa_w1_9<-merge(elsa_w1_fin,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w4_fin,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w5_fin,H_elsa_w1_9, by = "idauniq", all = TRUE)
H_elsa_w1_9<-merge(elsa_w6_fin,H_elsa_w1_9, by = "idauniq", all = TRUE)

```



```{r eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

### list longitudinal variables that will be used

inwVars   <- paste0("inw",4:6)        # Whether in current wave

proxyVars <- paste0("r",4:6,"proxy")  # Whether responses are proxy

iwstatVars<- paste0("r",4:6,"iwstat") # Respondet status (alive..)

ageVars   <- paste0("r",4:6,"agey")   # Age variables

painVars  <- paste0("r",4:6,"painlv") # Pain severity variables

painBin   <- paste0("r",4:6,"painfr")  # Whether often bother by pain

hepagVars <- paste0("r",4:6,"hepag") # How long the pain has been

loc_vars  <- outer(paste0("r", 4:6),  # Prefixes
                  loc_varnames,      # Base names
                  paste0            # Function to combine
                  ) %>% as.vector()

impactVars<- paste0("r",4:6,"lvimp")  # health limits self-care and work activities

deprVars<- c(paste0("r",3:6,"cesd"),  # cesd total score
             paste0("r",3:6,"depres"),
             paste0("r",3:6,"effort"),
             paste0("r",3:6,"sleepr"),
             paste0("r",3:6,"whappy"),
             paste0("r",3:6,"flone"),
             paste0("r",3:6,"fsad"),
             paste0("r",3:6,"going"),
             paste0("r",3:6,"enlife"))

lonelyness_var<-c(paste0("r",3:6,"lnlys"))

CRPVars   <-c("r4hscrp","r6hscrp")    # C-reactive Proteine measure (only w4 & w6)

finVARS   <- c(paste0("r",4:6,"nettotw_bu_s"),
               paste0("r",4:6,"totwq5_bu_s"))

SSuppVARS<- c(paste0("r",4:6,"ssupport6"),  # Marital strain items
                paste0("r",4:6,"sustdfe"), 
                paste0("r",4:6,"srely"), 
                paste0("r",4:6,"sopenup"), 
                paste0("r",4:6,"scritze"), 
                paste0("r",4:6,"sletdow"), 
                paste0("r",4:6,"sgetnev")
              )

PActVars<-c(paste0("r",4:6,"vgactx_e"),# - Vigorous Act
           paste0("r",4:6,"mdactx_e"),# - Moderately energetic Act
           paste0("r",4:6,"ltactx_e"))# - Mildly energetic Act)

DemoVars<-c("raracem",
            "raeduc_e",
            paste0("r",c(1,4:6),"work"),  # whether does paid work
            paste0("r",4:6,"mpart"),      # implied partnership
            paste0("r",4:6,"mstat"),      # partnership status (married)
            paste0("r",c(1,4:6),"smoken"),# smoking
            paste0("r",4:6,"drinkwn_e"),  # Drinking
            paste0("r",c(4,6),"mbmi"),    # measured BMI
            ###### MLTC
            paste0("r",c(1,4:6),"angine"), # angina pectoris
            paste0("r",c(1,4:6),"hrtatte"),# myocardial infarction
            paste0("r",c(1,4:6),"stroke"), # stroke
            paste0("r",c(1,4:6),"diabe"),  # diabetes
            paste0("r",c(1,4:6),"lunge"),  # pulmonary disease (chronic bronchitis, emphysema)
            paste0("r",c(1,4:6),"arthre"), # arthritis (osteoarthritis | rheumatoid | arthritis type not specified)
            paste0("r",c(1,4:6),"cancre"), # cancer
            paste0("r",c(1,4:6),"alzhe"),  # Alzheimer's disease
            paste0("r",c(1,4:6),"demene"), # dementiar4psyche
            paste0("r",c(1,4:6),"psyche"), # dementia


            paste0("r",c(1,4:6),"parkine"), # Parkinson
            paste0("r",c(4:6),"adltot6"),   # adl total 6 item count (for inputation) 
            paste0("r",c(4:6),"shlt")      # self reported health (for inputation) 
)


```





# SELECTING WAVE 4-6

1. select variables of interest
2. select sample 50yo or older
3. exclude responded with unknown live-status (potentially dead at wave 4,5,6)
4. 
```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
#select wave 4,5,6
H_elsa_w4_6<-H_elsa_w1_9 %>% dplyr::select(all_of(c("idauniq","ragender", 
                                                    inwVars,
                                                    iwstatVars,
                                                    proxyVars,
                                                    ageVars,
                                                    DemoVars,
                                                    deprVars,
                                                    lonelyness_var,
                                                    painVars,
                                                    painBin,
                                                    hepagVars,
                                                    loc_vars,
                                                    impactVars,
                                                    finVARS,
                                                    SSuppVARS,
                                                    PActVars,
                                                    CRPVars))) %>% filter(inw4 == 1) # n = 11050


H_elsa_w4_6<-H_elsa_w4_6 %>% filter(r4agey>=50) # n = 10749

########################################
############ filtering out "CAT" states
#######################################
#  rXiwstat == 1 ~ "responded, alive",
#  rXiwstat == 4 ~ "no-response but alive",
#  rXiwstat == 5 ~ "died between the last interview and the current one",
#  rXiwstat == 6 ~ "died in a previous wave",
#  rXiwstat == 7 ~ "dropped out from sample",
#  rxiwstat == 9 ~ "cat state"
####### THIS WONT BE NECESSARY WITH HSE data

H_elsa_w4_6<-H_elsa_w4_6 %>% filter(r4iwstat != 9) # n = 10749
H_elsa_w4_6<-H_elsa_w4_6 %>% filter(r5iwstat != 9) # n = 10071
H_elsa_w4_6<-H_elsa_w4_6 %>% filter(r6iwstat != 9) # n = 9452


########################################
############ filtering out responded that died at wave 6 or previously
#######################################
H_elsa_w4_6<-H_elsa_w4_6 %>% filter(r6iwstat != 6, r6iwstat != 5 ) # n = 8726


H_elsa_w1<-H_elsa_w1_9 %>% filter(r1cohort_e == 1, 
#                                  r1proxy != 1,
                                  r1agey>=50
                                  )




```


# Clearing missingness codes

all negative value codes that correspond to an NA reason are substituted with NA
```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

##### Preserving NA-code "-15" as EDUCATION VARIABLE CATEGORY "OTHER"

## preserving category "-15" as "other" education, 
# we recode value -15 to value "2" (unused) to prevent treating -15 as missing value further down in the pipeline

H_elsa_w1$raeduc_e <- ifelse(H_elsa_w1$raeduc_e == -15, 2,H_elsa_w1$raeduc_e)
H_elsa_w4_6$raeduc_e <- ifelse(H_elsa_w4_6$raeduc_e == -15, 2,H_elsa_w4_6$raeduc_e)


###### clearing missingness codes (values less than 0) from pain vars:
ssupport6_items<- c(paste0("r",4:6,"sustdfe"), 
            paste0("r",4:6,"srely"), 
            paste0("r",4:6,"sopenup"), 
            paste0("r",4:6,"scritze"), 
            paste0("r",4:6,"sletdow"), 
            paste0("r",4:6,"sgetnev"))

col_with_NA<-c(painVars, painBin,hepagVars,CRPVars,ssupport6_items,DemoVars,deprVars,PActVars)

H_elsa_w4_6<-H_elsa_w4_6 %>%
  mutate(across(all_of(col_with_NA), ~if_else(. < 0, NA_real_, .)))

H_elsa_w1<-H_elsa_w1 %>%
  mutate(across(all_of(col_with_NA), ~if_else(. < 0, NA_real_, .)))

```


# DERIVED VARIABLES

## Chronic pain Classification
```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
######################################################################
## CHRONIC PAIN CLASSIFICATION
######################################################################

H_elsa_w4_6$r4painchr <- chronic_pain_var(H_elsa_w4_6, "r4painfr", "r4hepag", "r4painlv")
H_elsa_w4_6$r5painchr <- chronic_pain_var(H_elsa_w4_6, "r5painfr", "r5hepag", "r5painlv")
H_elsa_w4_6$r6painchr <- chronic_pain_var(H_elsa_w4_6, "r6painfr", "r6hepag", "r6painlv")
  

#table(H_elsa_w4_6$r4painchr, useNA = "ifany") %>% addmargins()
#table(H_elsa_w4_6$r5painchr, useNA = "ifany") %>% addmargins()
#table(H_elsa_w4_6$r5painfr,H_elsa_w4_6$r5painlv, useNA = "ifany") %>% addmargins()

# Function to create summary table for a wave
create_CPsummary(H_elsa_w4_6, "r4painchr", "Wave 4")
create_CPsummary(H_elsa_w4_6, "r5painchr", "Wave 5")
create_CPsummary(H_elsa_w4_6, "r6painchr", "Wave 6")

chronicPAINvars<-c("r4painchr","r5painchr","r6painchr")

### creating a binary variable
# pain severity (for those with pain):
# I am collapsing "moderate" and "severe" because of clinical importance and
# to lower the number of categories that the LTA will have to handle.
# I noticed that too many categories could lead to:
# 1. sparse contingency tables, 
# 2. difficult interpretation of models
# 3. potentially unstable solutions with large SE

# For those without pain, set to two levels: 
# "No pain or mild pain" (0~ 0|1); "moderate or severe pain" (1~2|3)
H_elsa_w4_6<-H_elsa_w4_6 %>% mutate(r4painsev = case_when(r4painfr <  0 ~ NA_real_,
                                                          r4painfr == 0 ~ 0,
                                                          r4painlv >= 2 ~ 1,
                                                          r4painlv >  0 ~ 0,
                                                          TRUE ~ NA_real_),
                                    r5painsev = case_when(r5painfr <  0 ~ NA_real_,
                                                          r5painfr == 0 ~ 0,
                                                          r5painlv >= 2 ~ 1,
                                                          r5painlv >  0 ~ 0,
                                                          TRUE ~ NA_real_),
                                    r6painsev = case_when(r6painfr <  0 ~ NA_real_,
                                                          r6painfr == 0 ~ 0,
                                                          r6painlv >= 2 ~ 1,
                                                          r6painlv >  0 ~ 0,
                                                          TRUE ~ NA_real_),
  
)

#table(H_elsa_w4_6$r4painchr,H_elsa_w4_6$r4painsev, useNA = "ifany") %>% addmargins()
#table(H_elsa_w4_6$r5painchr,H_elsa_w4_6$r5painsev, useNA = "ifany") %>% addmargins()
#table(H_elsa_w4_6$r6painchr,H_elsa_w4_6$r6painsev, useNA = "ifany") %>% addmargins()


```


## PHYSICAL ACTIVITY
Participants were asked to self-report the frequency of their participation in sports or activities that were vigorous, moderately energetic and mildly energetic (more than once a week, once a week, one to three times a month, hardly ever or never). Physical activity was then categorised into four groups in accordance with previous literature [@dhalwani2016long,malkowski2023socioeconomic]: 
(1) inactive (no activity on a weekly basis); 
(2) only mild activity at least once per week; 
(3) at least moderate but no vigorous activity at least once per week; and 
(4) vigorous activity at least once per week.
```{r}
# rwvgactx_e - Vigorous Act
# rwmdactx_e - Moderately energetic Act
# rwltactx_e - Mildly energetic Act
#
# The above indicate the frequency of vigorous, moderately energetic, or mildly energetic physical activity, respectively. coded as:

# -18 : Refused item 
# -4 Don't know
# 2 more than once a week
# 3 once a week
# 4 one to three times a month
# 5 hardly ever or never

# Create new physical activity category variable
H_elsa_w4_6<-create_PA_categories(H_elsa_w4_6,wave = 4)
H_elsa_w4_6<-create_PA_categories(H_elsa_w4_6,wave = 5) 
H_elsa_w4_6<-create_PA_categories(H_elsa_w4_6,wave = 6)

```



## Depression 8 item
Depression is strongly linked with chronic pain [@mills2019chronic]

The CES-D-8 is a valid and reliable measure of depressive symptoms in community-dwelling older people, [@briggs2018validation] and a score of >=3 is generally accepted as threshold for caseness[@turvey1999revised,white2016duration].

### Depression latent score
```{r,echo=TRUE,eval=FALSE, warning=FALSE,message=FALSE,error=FALSE,results='hide'}
library(lavaan)
library(semTools)


######################################################
##### CONFIRMATORY FACTOR ANALYSIS: wave 4 
#######################################################

CESD_8itm.W4<-'Lv1   =~ r4depres + r4effort + r4sleepr + r4whappy + r4flone + r4fsad + r4going + r4enlife'

CESD_8itm.W4fit<- cfa(model = CESD_8itm.W4, 
                     data = H_elsa_w4_6,
                     ordered = TRUE, # polychoric correlations for the binary variables
                     std.lv = TRUE,
                     auto.fix.first = FALSE,
                     estimator = "WLSM", #  WLSM or alternatively WLSMV
                     missing = "pairwise" # this way I mainteing the NA where appropriate in the lavPredict output
                     )
                                 


r<-residuals(CESD_8itm.W4fit, type = "cor")$cov %>% as.table %>% round(2)
r[abs(r) < 0.1] <- "-"

#r

## Diagnostic tool to explore residual correlations or other potential sources of bias in the model
#m<-modificationindices(ALL.one.factor.Fit)

#summary(CESD_8itm.W4fit, fit.measures = T, rsquare=T)

#######################################################
##### END CONFIRMATORY FACTOR ANALYSIS: wave 4
#######################################################

#######################################################
###### MEASUREMENT INVARIANCE TESTING
######################################################

## define the 3 latent factors that are actually the same factor (FU) measured overtime
longFacNames <- list(FU = c(#"CESD_w3",
                            "CESD_w4","CESD_w5","CESD_w6"))


# Model specification with metric invariance constraints
free_model <- '
  # Define latent factors for each wave
 # CESD_w3 =~ r3depres + r3effort + r3sleepr + r3whappy + r3flone + r3fsad + r3going + r3enlife
  CESD_w4 =~ r4depres + r4effort + r4sleepr + r4whappy + r4flone + r4fsad + r4going + r4enlife
  CESD_w5 =~ r5depres + r5effort + r5sleepr + r5whappy + r5flone + r5fsad + r5going + r5enlife  
  CESD_w6 =~ r6depres + r6effort + r6sleepr + r6whappy + r6flone + r6fsad + r6going + r6enlife

  # Allow latent factor residuals to correlate across waves
#  CESD_w3 ~~ CESD_w4
#  CESD_w3 ~~ CESD_w5
#  CESD_w3 ~~ CESD_w6
  CESD_w4 ~~ CESD_w5
  CESD_w4 ~~ CESD_w6
  CESD_w5 ~~ CESD_w6
'

fit_free <- measEq.syntax(configural.model = free_model,
                               data = H_elsa_w4_6,
                               ordered = TRUE,
                               parameterization = "theta", # theta parameterization (for ordinal data)
                               ID.fac = "std.lv", 
                               longFacNames = longFacNames,
                               fixed.x = TRUE,
                               return.fit = TRUE,
                               missing = "pairwise" # this way I mainteing the NA where appropriate in the lavPredict output
                                                    # allows the model fitting to retain the maximum amount of observed data 
                                                    # per covariance element, rather than dropping entire rows if any wave/item 
                                                    # has missingness, while still accounting for NA in predictions
                          )
summary(fit_free, standardized = T, fit.measures = T)


fit_metric <- measEq.syntax(configural.model = free_model,
                               data = H_elsa_w4_6,
                               ordered = TRUE,
                               parameterization = "theta",
                               ID.fac = "std.lv", 
                               longFacNames = longFacNames,
                               long.equal = c("loadings"),
                               fixed.x = TRUE,
                               return.fit = TRUE,
                               missing = "pairwise" # this way I mainteing the NA where appropriate in the lavPredict output
                          )
summary(fit_metric, standardized = T, fit.measures = T)

anova(fit_free, fit_metric)

#summary(compareFit(fit_free, fit_metric))
#######################################################
###### END OF MEASUREMENT INVARIANCE TESTING
######################################################
timestamp_suffix <- format(Sys.time(), "%Y%m%d_%H%M%S")
#saveRDS(fit_free,paste0(paper1_MODELS_rds_path,"cesd_lat_free_fit",timestamp_suffix,".rds"))
#saveRDS(fit_metric,paste0(paper1_MODELS_rds_path,"cesd_lat_metricInv_fit",timestamp_suffix,".rds"))
```


```{r}

#fit_free<-readRDS(paste0(paper1_MODELS_rds_path,"cesd_lat_free_fit20260313_160306.rds"))
fit_metric<-readRDS(paste0(paper1_MODELS_rds_path,"cesd_lat_metricInv_fit20260313_160306.rds"))

cesd_scores <- lavPredict(fit_metric, 
                        newdata = H_elsa_w4_6, 
                        type = "lv")

#H_elsa_w4_6$r3cesd_lat<-cesd_scores[,1]
H_elsa_w4_6$r4cesd_lat<-cesd_scores[,1]
H_elsa_w4_6$r4cesd_lat %>% hist
H_elsa_w4_6$r5cesd_lat<-cesd_scores[,2]
H_elsa_w4_6$r5cesd_lat %>% hist
H_elsa_w4_6$r6cesd_lat<-cesd_scores[,3]
H_elsa_w4_6$r6cesd_lat %>% hist
```



### Depression categorical scale
```{r}
###########################################
# CREATE A categorical variable for Depression

H_elsa_w4_6 <- H_elsa_w4_6 %>% 
                    mutate(r4cesd8_cat = case_when(
                                      r4cesd == 0 ~ "0 depr. signs",
                                      r4cesd == 1 ~ "1 depr. signs",
                                      r4cesd == 2 ~ "2 depr. signs",
                                      r4cesd >= 3 ~ "3 or more",
                                      TRUE ~ NA_character_),
                           r5cesd8_cat = case_when(
                                      r5cesd == 0 ~ "0 depr. signs",
                                      r5cesd == 1 ~ "1 depr. signs",
                                      r5cesd == 2 ~ "2 depr. signs",
                                      r5cesd >= 3 ~ "3 or more",
                                      TRUE ~ NA_character_),
                           r6cesd8_cat = case_when(
                                      r6cesd == 0 ~ "0 depr. signs",
                                      r6cesd == 1 ~ "1 depr. signs",
                                      r6cesd == 2 ~ "2 depr. signs",
                                      r6cesd >= 3 ~ "3 or more",
                                      TRUE ~ NA_character_)
                           )
```

## MLTC
```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

###############################################################
###############################################################
#  CREATE MLTC VARIABLE
###############################################################
###############################################################
### MLTC: count of doctor-diagnosed illnesses [@poole2020combined]):
  # rXangine | rXhrtatte: CHD (angina | myocardial infarction)
  # rXstroke: stroke
  # rXdiabe: diabetes/high blood glucose
  # rXlunge: pulmonary disease (chronic bronchitis, emphysema)
  # rXarthre:   - arthritis (osteoarthritis | rheumatoid | arthritis type not specified)
  # rXcancre: cancer

# additional categories used in [@torres2016wealth]
  # rXparkine: Parkinson
  # cardiovascular diseases (
  #  rxhibpe   : high blood pressure, 
  #  rXangine  : angina, 
  #  rXhrtatte : heart attack, 
  #  rXconhrtfe: heart failure, 
  #  rXhrtmre  : heart murmur 
  #  rXhrtrhme : heart rhythm)

# additional categories used in [@simpson2023eliciting]:
  # rxhibpe    : hypertension | high blood pressure
  # rXasthmae  : asthma, 
  # rXpsyche   : mental health disorders (diagnosed: depression, anxiety, emotional problems, hallucinations, schizophrenia, psychosis) 
  # rxalzhe    : Alzheimer's disease, 
  # rXdemene   : dementia 


############################# poole2020 definition 
### MLTC: count of doctor-diagnosed illnesses [@poole2020combined] with the addition of ):

H_elsa_w1 <- create_MLTC_variable(
  H_elsa_w1, 
  "r1MLTC",
  c("r1angine", "r1stroke", "r1diabe", "r1lunge", "r1arthre", "r1cancre","r1psyche")
)


H_elsa_w4_6 <- create_MLTC_variable(
  H_elsa_w4_6, 
  "r4MLTC",
  c("r4angine", "r4stroke", "r4diabe", "r4lunge", "r4arthre", "r4cancre","r4psyche")
)

H_elsa_w4_6 <- create_MLTC_variable(
  H_elsa_w4_6, 
  "r5MLTC",
  c("r5angine", "r5stroke", "r5diabe", "r5lunge", "r5arthre", "r5cancre","r5psyche")
)

H_elsa_w4_6 <- create_MLTC_variable(
  H_elsa_w4_6, 
  "r6MLTC",
  c("r6angine", "r6stroke", "r6diabe", "r6lunge", "r6arthre", "r6cancre","r6psyche")
)
```

## N of Pain Locations

```{r}
loc_varnames <- c("hepawba", "hepawhi", "hepawkn", "hepawfe", "hepawmo", "hepawot", "hepawal")


# Create for wave 4
wave4_vars <- paste0("r4", loc_varnames)
H_elsa_w4_6 <- create_pain_locations_count(H_elsa_w4_6, "r4pain_locN", wave4_vars)

# Create for waves 5 and 6
wave5_vars <- paste0("r5", loc_varnames)
H_elsa_w4_6 <- create_pain_locations_count(H_elsa_w4_6, "r5pain_locN", wave5_vars)

wave6_vars <- paste0("r6", loc_varnames)
H_elsa_w4_6 <- create_pain_locations_count(H_elsa_w4_6, "r6pain_locN", wave6_vars)

```

### Compoasite Chronic pain state + location

0 NO PAIN
1 ACUTE PAIN
3 CHRONIC
4 CHRONIC 3+ locations

```{r}
H_elsa_w4_6<- H_elsa_w4_6%>% mutate(
                            r4painchr_ext = case_when(
                                    r4painchr == 0 ~ 0,
                                    r4painchr == 1 & r4pain_locN_cat == "2 or less location"  ~ 1,
                                    r4painchr == 2 & r4pain_locN_cat == "2 or less location"  ~ 2,
                                    r4painchr == 2 & r4pain_locN_cat == "3 or more locations" ~ 3,
                                    TRUE ~ NA_real_
                                    )%>% factor(levels = 0:3, 
                                                labels = c("No pain", 
                                                           "Acute pain", #"Acute widespread",
                                                           "Chronic pain", 
                                                           "Chronic 3+locations")),
                            r5painchr_ext = case_when(
                                    r5painchr == 0 ~ 0,
                                    r5painchr == 1 & r5pain_locN_cat == "2 or less location"  ~ 1,
                                    r5painchr == 2 & r5pain_locN_cat == "2 or less location"  ~ 2,
                                    r5painchr == 2 & r5pain_locN_cat == "3 or more locations" ~ 3,
                                    TRUE ~ NA_real_
                                    )%>% factor(levels = 0:3, 
                                                labels = c("No pain", 
                                                           "Acute pain", #"Acute widespread",
                                                           "Chronic pain", 
                                                           "Chronic 3+locations")),
                            r6painchr_ext = case_when(
                                    r6painchr == 0 ~ 0,
                                    r6painchr == 1 & r6pain_locN_cat == "2 or less location"  ~ 1,
                                    r6painchr == 2 & r6pain_locN_cat == "2 or less location"  ~ 2,
                                    r6painchr == 2 & r6pain_locN_cat == "3 or more locations" ~ 3,
                                    TRUE ~ NA_real_
                                    )%>% factor(levels = 0:3, 
                                                labels = c("No pain", 
                                                           "Acute pain", #"Acute widespread",
                                                           "Chronic pain", 
                                                           "Chronic 3+locations")),
                            )

```


## Partnership classification
# we create a partnership variable:
responded is categorised as having a partner if any of the following is true:
* r4mstat == 1 ("married")
* r4mstat == 4 ("partnered")
* r4mpart == 1 ("whether the respondent partnership is implied")
* r4ssupport >0 ("partnership implied because the respondent reported the Marital strain experienced with their partner")

```{r}
H_elsa_w4_6 <- H_elsa_w4_6 %>%
  mutate(
    r4partner = case_when(r4mpart == 1 | r4mstat == 1 | r4ssupport6>=0 ~ 1 ,
                          TRUE ~ 0),
    r5partner = case_when(r5mpart == 1 | r5mstat == 1 | r5ssupport6>=0 ~ 1 ,
                          TRUE ~ 0),
    r6partner = case_when(r6mpart == 1 | r6mstat == 1 | r6ssupport6>=0 ~ 1 ,
                          TRUE ~ 0)
    )

H_elsa_w1 <- H_elsa_w1 %>%
  mutate(
    r1partner = case_when(r1mpart == 1 | r1mstat == 1 | r1ssupport6>=0 ~ 1 ,
                          TRUE ~ 0)
    )
```

## excluding respondent with all data missing

```{r}

H_elsa_w4_6$dropout_status <- create_dropout_variable(H_elsa_w4_6)

# Create a summary table of dropout status
dropout_table <- table(H_elsa_w4_6$dropout_status, useNA = "ifany")
dropout_df <- as.data.frame(dropout_table)
names(dropout_df) <- c("Dropout Status", "Frequency")
dropout_df$Percentage <- round(100 * dropout_df$Frequency / sum(dropout_df$Frequency), 2)

# Add labels for better readability
dropout_df$`Dropout Status` <- factor(dropout_df$`Dropout Status`, 
                                      levels = c("-3", 
                                                 "-2", 
                                                 "-1", 
                                                 "0", 
                                                 "5", 
                                                 "6", 
                                                 "NA"),
                                      labels = c("All waves missing", 
                                                 "Intermittent missing", 
                                                 "Missing with return",
                                                 "Complete response",
                                                 "Dropout at Wave 5",
                                                 "Dropout at Wave 6",
                                                 "Missing"))

print(dropout_df)



## exclude participants that dropped out at wave 5 without return

H_elsa_w4_6<-H_elsa_w4_6 %>% filter(dropout_status !=5) # n = 8656


#filter cases where proxy was never 1 (all zeroes) OR was 1 in only 1 wave
H_elsa_w4_6<-H_elsa_w4_6 %>%  filter((r4proxy == 0 & r5proxy == 0 & r6proxy == 0) | 
         (r4proxy + r5proxy + r6proxy) == 1) # 8014


nrow(H_elsa_w4_6)

```


```{r eval=F}
saveRDS(H_elsa_w1, paste0(paper1_rds_path,"H_elsa_w1.rds")) 
saveRDS(H_elsa_w4_6, paste0(paper1_rds_path,"H_elsa_w4_6.rds")) 

```






