---
title: "Supplementary pt.3 — Variable Definitions and Descriptive Statistics"
author: "Diego Vitali"
output:
  github_document:
    toc: True
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(naniar)
library(NbClust)
library(readr)
library(polycor)
library(patchwork)
library(corrplot)
library(GGally)
library(parallel)
library(table1)
library(knitr)
library(tableone)

source("Sources/Paper1_func.R")

```

```{r eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
paper1_rds_path<-"~/private_WP5/WP5_data/rds/"

# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

H_elsa_w4_6<-readRDS(paste0(paper1_rds_path,"H_elsa_w4_6.rds"))
H_elsa_w1<-readRDS(paste0(paper1_rds_path,"H_elsa_w1.rds"))


```

# Supplementary Material pt.3 — Variable Definitions and Descriptive Statistics {.unnumbered}

This supplementary document provides full variable definitions, data source information, and descriptive statistics for all measures used in the analysis. Sections follow the order of the main paper methods section. Code chunks are shown for reproducibility; all derivations are implemented in `suppl_pt1_Data_prep.Rmd` unless otherwise noted.

# Variable Inventory

The table below summarises all variables entering the LTA model: their analysis name, original ELSA source variable(s), data file, whether they are time-varying, and the transformation applied.

| Role | Analysis variable | ELSA source variable(s) | Source file | Time-varying | Transformation |
|---|---|---|---|---|---|
| Outcome indicator | `painchr_ext` | `rXpainfr`, `rXpainlv` (harmonised); `hepag`, `hepawXX` (wave-specific) | `Hg3_w1_9_ALL.tab`; `wave_N_elsa_data_vN.tab` | Yes | Derived: four-level (no pain / acute / localised chronic / multisite chronic); pt1 |
| Outcome indicator | `lvimp_shifted_cat` | `rXlvimp` items (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | IRT score → zero separation + PAM clustering → three levels; Wave 4 medoids fixed; pt2.1 |
| Exposure | `partner` | `rXmpart` (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | Binary recoded (0/1); pt1 |
| Exposure | `ssupport6_cat` | `rXssupport6` (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | Median cut at 1.5 (back-transformed from log1p) → three levels; pt4 |
| TE confounder | `Sex_BIN` | `ragender` (harmonised) | `Hg3_w1_9_ALL.tab` | No | Binary factor (male reference); pt2.1 |
| TE confounder | `raagey_z` | `raagey` = `r4agey` (harmonised) | `Hg3_w1_9_ALL.tab` | No | Standardised; Wave 4 baseline only; pt2.1 |
| TE confounder | `raeduc_e` | `raeduc_e` (harmonised) | `Hg3_w1_9_ALL.tab` | No | −15 recoded to “other”; five-level factor; pt1 |
| TE confounder | `wealthQ_bin` | `rXnettotw_bu_s`, `rXtotwq5_bu_s` | `wave_N_financial_derived_variables.tab` | Yes | Worst-off quintile (Q1) vs all others (Q2–Q5); pt2.1 |
| TE confounder | `MLTC` | `rXangine`, `rXstroke`, `rXdiabe`, `rXlunge`, `rXarthre`, `rXcancre`, `rXpsyche` (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | Count → categorical (0/1/2/3+); pt1 |
| DE confounder | `cesd_lat_shifted` | `rXcesd` items (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | CFA latent score (metric invariance model); shifted to non-negative before imputation; pt2 |
| DE confounder | `lnlys` | `rXlnlys` (harmonised) | `Hg3_w1_9_ALL.tab` | Yes | UCLA-4 mean, continuous; imputed in pt2, passed through pt2.1 |

Note: `hrtatte` (myocardial infarction) is present in the data but excluded from MLTC by design — CHD is represented by angina (`rXangine`) only, following [@poole2020combined].

# Variable Definitions

## Outcome Indicators

### Chronic pain status (`painchr_ext`)

Chronic pain status (`painchr_ext`) was derived from four ELSA items: `rXpainfr` ("are you often troubled by pain?"), `rXpainlv` (usual pain level: none / mild / moderate / severe), `hepag` ("how long has the pain been bothering you?"), and a count of pain body locations. Note that `hepag` and the location items are wave-specific variables not available in the harmonised dataset and were merged from the ELSA core wave files. We first classified participants as having no pain (`rXpainfr` = 0), acute pain (`rXpainfr` = 1 and `hepag` < 3 months), or chronic pain (`rXpainfr` = 1 and `hepag` ≥ 3 months) [@guideline2021chronic; @kang2023chronic; @nicholas2019iasp]. Among those meeting the chronic pain criterion, we further distinguished those reporting pain in three or more body locations from those reporting pain in two or fewer. This was not implemented as a diagnostic criterion for widespread pain because ELSA only assesses pain at seven body locations, which precludes an appropriate implementation of validated widespread pain criteria [@wolfe2019cwp]. Instead, this was a pragmatic choice to retain information about pain diffuseness within the chronic pain category. In the context of the present LTA model, this approach recognises that respondents with localised and more diffuse chronic pain may represent clinically distinct phenotypes, and allows the latent states to reflect this difference if present.

Source items:

-   `rXpainfr` — frequent pain (harmonised): 0 = no, 1 = yes
-   `rXpainlv` — pain severity (harmonised): 0 = none, 1 = mild, 2 = moderate, 3 = severe
-   `hepag` — pain duration (wave-specific): < 3 months / 3–6 months / 6–12 months / > 12 months
-   `hepawba`, `hepawhi`, `hepawkn`, `hepawfe`, `hepawmo`, `hepawot`, `hepawal` — pain locations (wave-specific): back, hips, knees, feet, mouth/teeth, elsewhere, all over

Resulting four-level indicator (`painchr_ext`):

-   0: No pain
-   1: Acute pain (pain present, duration < 3 months)
-   2: Localised chronic pain (duration ≥ 3 months, ≤ 2 locations)
-   3: Multisite chronic pain (duration ≥ 3 months, ≥ 3 locations)

### Functional health impact (`lvimp_shifted_cat`)

The continuous IRT-based health impact score (`rXlvimp`) shows partial scalar measurement invariance across waves (see supplementary material pt.1.2). The score is zero-inflated with evidence of multimodality in the non-zero portion. We categorised it into three levels using a two-part approach:

1. Zero-scorers were assigned to category 0 (no detectable functional impairment) a priori, as structural zeros on the IRT scale represent a qualitatively distinct state.
2. The non-zero portion (scores > 0.1; the 0.1 threshold excludes artefactual near-zero scores arising from freed item intercepts under partial scalar invariance) was clustered using PAM [@saxena2017review] with the number of clusters selected via NbClust [@charrad2014nbclust] and validated by silhouette width [@rousseeuw1987silhouettes]. A k = 3 solution was selected, yielding medoids of 0.879, 1.392, and 2.025 on the IRT scale.
3. Bootstrap resampling (2,000 iterations) confirmed stability of the Wave 4 medoids. These were applied as fixed reference centres to waves 5 and 6: each observation was assigned to the nearest medoid by absolute distance, ensuring a consistent categorisation scale across waves.

The final three-level indicator (`lvimp_shifted_cat3`) used in the LTA:

-   0: No functional impairment (structural zeros)
-   1: Some health impact (PAM clusters 1–2, medoids 0.879 and 1.392 merged)
-   2: High health impact (PAM cluster 3, medoid 2.025)

Full derivation code is in `suppl_pt2.1_DERIVED VARIABLES_VH.Rmd`.
    


##Exposure

### Partnership status and spousal support (`ssupport6_cat`)

Partnership quality was formalised as a three-level exposure variable combining partner status with a measure of perceived spousal support. Partner status (`rXmpart`) distinguished participants with and without a partner at each wave (0 = no partner, 1 = partner).

Among partnered participants, perceived spousal support was assessed via `rXssupport6`, a composite of six items:

-   `RwSUSTDFE` — spouse really understands the way the respondent feels
-   `RwSRELY` — respondent can rely on their spouse if they have a serious problem
-   `RwSOPENUP` — respondent can open up to their spouse when they need to talk about worries
-   `RwSCRITZE` — spouse criticises the respondent
-   `RwSLETDOW` — spouse lets the respondent down when counting on them
-   `RwSGETNEV` — spouse gets on the respondent’s nerves

Each item is scored 1–4; higher values indicate lower perceived support. The composite is right-skewed with a median of 1.5 at each wave (see distribution plots and summary table below). Partnered participants were divided at the median: those scoring ≤ 1.5 (supportive partner) and those scoring > 1.5 (less-than-supportive partner). Combined with the no-partner group, this yields a three-level exposure: no partner (reference), less-than-supportive partner, and supportive partner.

## Confounders — TE set

Full confounder selection rationale and causal DAGs are in Section 2.5 below. Variable definitions follow.

### Sex (`ragender` / `Sex_BIN`)

Binary variable; male coded as reference category. Time-invariant.

### Age (`raagey` / `raagey_z`)

Continuous, standardised. Wave 4 baseline age used throughout as a time-invariant predictor [@maharani2023household].

### Education (`raeduc_e`)

Educational attainment was classified using the IFS Derived Variables highest qualification summary, based on the first non-missing report across waves [@banks2025english]. Five levels: less than secondary, other, secondary, some college, college and above. The value −15 (ambiguous response) was recoded to category 2 ("other education"). Time-invariant.

### Wealth (`rXnettotw_bu_s` / `wealthQ_bin`)

Total net non-pension household wealth (`nettotw_bu_s`) is a summary measure of financial, physical, and housing wealth minus debt, estimated from 22 components by the ELSA data depositors [@torres2016wealth; @demakakos2016wealth]. A detailed description is available at the UK Data Service documentation. Wealth quintiles (`totwq5_bu_s`) were binarised into the worst-off quintile (Q1) versus all others (Q2–Q5; `wealthQ_bin`). Time-varying; sourced from wave-specific financial derived files (not the harmonised dataset).

### Multiple long-term conditions (`MLTC`)

Count of doctor-diagnosed conditions categorised as 0, 1, 2, or 3 or more. Conditions follow the definition in [@poole2020combined], with the addition of mental health disorders from [@simpson2023eliciting]:

-   CHD (angina; myocardial infarction excluded by design — represented by angina only)
-   Stroke
-   Diabetes / high blood glucose
-   Pulmonary disease (chronic bronchitis, emphysema)
-   Arthritis (osteoarthritis, rheumatoid, unspecified)
-   Cancer
-   Mental health disorders (`rXpsyche`: doctor-diagnosed depression, anxiety, psychosis)

Note: mental health disorders are included as a countable MLTC condition; depression is additionally included as a continuous latent score in the DE model, where it acts as a covariate rather than an outcome indicator.

## Additional confounders — DE model only

### Depression (`rXcesd` / `cesd_lat_shifted`)

Depression was assessed using the eight-item CES-D scale. `RwCESD` is the sum of `RwDEPRES`, `RwEFFORT`, `RwSLEEPR`, `RwFLONE`, `RwFSAD`, `RwGOING`, `(1 − RwWHAPPY)`, and `(1 − RwENLIFE)`; higher scores indicate more negative affect. A latent factor score was derived from a confirmatory factor analysis (metric invariance model) and shifted to be non-negative prior to imputation (`cesd_lat_shifted`) [@karim2015validation]. Time-varying.

### Loneliness (`rXlnlys`)

Loneliness was assessed as the mean of four UCLA loneliness items (`RwCOMPLAC`, `RwLEFTOUT`, `RwISOLATE`, `RwINTUNE`; composite: `RwLNLYS`). Higher scores indicate greater loneliness. The psychometric properties of short UCLA loneliness scales have been validated against the full 20-item version, showing adequate dimensional reliability and convergent validity [@gosling2024measuring]. @gosling2024measuring also demonstrate that dichotomising short UCLA scales produces variable prevalence estimates and sensitivity/specificity below acceptable thresholds depending on the cut-off chosen; accordingly, the score was retained as continuous. Time-varying.


## Confounder Selection and DAGs

Following the disjunctive cause criterion [@vanderweele2011criterion], as formalised by @pearl2009causality, variables were included as confounders if they have plausible directed paths to both the partnership quality exposure and the pain-state outcome. Full causal DAGs and dagitty-derived adjustment sets are presented in Figures 2 and 3 of the supplementary material (see `draw_dags.R`).

The LTA model specification and parametric formulas (initial state and transition probability equations) are documented in supplementary material pt.4.

# Descriptive Results

We present the baseline levels of all measures; where relevant we compare them with Wave 1 data, which was directly sampled from the 2001 Health Survey for England (HSE) and serves as a benchmark for population representativeness.

## Demographic factors

### Age and sex

Age and sex are consistently associated with chronic pain prevalence and characteristics. Advancing age is independently associated with greater multimorbidity, which in turn predicts chronic pain onset and persistence. Pain characteristics including duration, severity, and number of sites show age-dependent patterns, particularly in women. Sex differences in chronic pain are well established: women report higher pain prevalence, higher self-reported intensity, and higher pain-related disability than men, and are more likely to seek treatment [@mills2019chronic].

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
cut_age_brackets <- function(ages) {
  cut(ages, 
      breaks = c(50,60,70,80,Inf), 
      labels = c("50-59", "60-69","70-79", "80+"),
      right = F)
}

## HSE age categories
H_elsa_w1$r1agey_cat<-H_elsa_w1$r1agey %>% cut_age_brackets()

## Wave 4 age categories (baseline)
H_elsa_w4_6$r4agey_cat<-H_elsa_w4_6$r4agey %>% cut_age_brackets()

# Prepare data
plot_data <- data.frame(
  age = c(H_elsa_w1$r1agey_cat, H_elsa_w4_6$r4agey_cat),
  Sex = factor(
    c(H_elsa_w1$ragender, H_elsa_w4_6$ragender),
    levels = c(1, 2),
    labels = c("Male", "Female")
  ),
  wave = c(
    rep("Wave 1 (HSE sample)", length(H_elsa_w1$r1agey_cat)), 
    rep("Wave 4", length(H_elsa_w4_6$r4agey_cat))
  )) %>%
  group_by(wave, age, Sex) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(
    percentage = (count / sum(count)),
    signed_percentage = case_when(
      wave == "Wave 1 (HSE sample)" & Sex == "Male" ~ -percentage,
      wave == "Wave 1 (HSE sample)" & Sex == "Female" ~ -percentage,
      wave == "Wave 4" & Sex == "Male" ~ percentage,
      wave == "Wave 4" & Sex == "Female" ~ percentage
  ))

wave_totals <- plot_data %>%
  group_by(wave) %>%
  summarise(total_respondents = sum(count)) %>%
  mutate(label = paste0(wave, " (n = ", total_respondents, ")"))

# Population pyramid plot
ggplot(plot_data) +
  geom_col(aes(fill = interaction(Sex, wave, sep = ":"), 
               y = signed_percentage,
               x = age), colour = "white", 
           position = "dodge") + 
  # Add percentage labels at the end of bars
  geom_text(
    aes(
      x = age,
      y = signed_percentage,
      label = scales::percent(abs(percentage), accuracy = 0.1),
      group = interaction(Sex, wave),
      hjust = ifelse(signed_percentage < 0, 1, -0)  # Outside bars
    ),
    position = position_dodge(width = 0.9),  # Match the dodge width
    size = 4
  ) +
  coord_flip() +
  facet_wrap(.~ wave, 
             scales = "free_x",
             strip.position = "bottom") +
  theme_minimal() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    name = "Percentage of Respondents",
    expand = expansion(mult = c(0, 0))  # Add more space for labels
  )+
  scale_x_discrete(
    name = "Age Bracket"
  )+
  theme(
    legend.position = "bottom",
    panel.spacing.x = unit(0, "pt")
  )+
  labs(
    title = "Age bracket distribution by sex: Wave 1 (HSE) vs Wave 4",
    fill = "Sex:Wave",
    caption = paste0("Wave 1 (HSE) n =",nrow(H_elsa_w1),
                     "; Wave 4 n = ",nrow(H_elsa_w4_6))
  )


```

## Socioeconomic factors

### Net total wealth

The prevalence of chronic pain is inversely related to socioeconomic
factors [@mills2019chronic]. Those who are socio-economically deprived are not only more likely to
experience chronic pain than people from more affluent areas, but they
are also more likely to experience more severe pain and a greater level
of pain-related disability [@mills2019chronic]

Net total wealth (`nettotw_bu_s`) is the sum of savings, investments, physical and housing wealth minus financial and mortgage debt. The distribution is severely right-skewed; the log scale is used for visual clarity. In general there is a clear wealth gradient in disability among older English adults with the 
poorest men with disability more likely to live without a partner and have no weekly contact with children, 
family or friends compared to the wealthiest. Among women with disability, the poorest are 
more likely to report loneliness and have no partner while the wealthiest and the intermediate 
groups were more likely to be living with a partner [@torres2016wealth].


```{r echo=FALSE, warning=F,message=FALSE,error=FALSE}
Sys.setlocale("LC_ALL", "en_GB.UTF-8")   # adjust to your region

Wealth_data<- H_elsa_w1 %>% 
                select(r1nettotw_bu_s, r1totwq5_bu_s)                                                                                         
Wealth_median<-median(Wealth_data$r1nettotw_bu_s, na.rm = T)

plot_Wealth_w1<-ggplot(Wealth_data, aes(x = r1nettotw_bu_s)) +
  geom_histogram(fill = "blue", color = "black", alpha = .4) +
  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10),
                     labels = scales::comma,
                     guide = guide_axis(angle =  90),
                     breaks = c(-10^4,-10^3,-10^2,-10,0,10,10^2,10^3,10^4,10^5,10^6,10^7)) +
  geom_vline(xintercept = Wealth_median, linetype="dotted", color = "red")+
  labs(title = "Distribution of net total wealth (£, log scale) Wave 1",
       subtitle = "The sum of savings, investments, physical wealth and housing wealth, minus \nfinancial debt and mortgage debt.",
       x = "Total Net Wealth ( log scale)",
       y = "Count",
       caption = paste0("Dotted line is the median net total wealth m= £",Wealth_median, 
                        ", n = ",nrow(Wealth_data))) +
  theme_minimal()

plot_Wealth_w1
```

The distribution by quintile allows comparison of Wave 1 (HSE) and Wave 4 wealth levels.

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}


w4quintile_breaks <- quantile(H_elsa_w4_6$r4nettotw_bu_s, 
                           probs = seq(0, 1, by = 0.2),na.rm = T)


# Use these breakpoints to assign quintiles to all values
H_elsa_w4_6$r4totwq5_bu_sIMP <- cut(H_elsa_w4_6$r4nettotw_bu_s, 
                                     breaks = w4quintile_breaks, 
                                     labels = FALSE,
                                     include.lowest = TRUE)


# Where quintiles exist, keep the original values
H_elsa_w4_6$r4totwq5_bu_s <- ifelse(
  is.na(H_elsa_w4_6$r4totwq5_bu_s),
  H_elsa_w4_6$r4totwq5_bu_sIMP,
  H_elsa_w4_6$r4totwq5_bu_s
)


plot_dataw1 <- H_elsa_w1 %>% 
  select(r1nettotw_bu_s, r1totwq5_bu_s) %>%                                                         
  mutate(
    r1nettotw_bu_s = as.integer(r1nettotw_bu_s),
    r1totwq5_bu_s = as.integer(r1totwq5_bu_s)
  ) %>% 
  na.omit() %>%
  mutate(wave = "Wave 1")

plot_dataw4 <- H_elsa_w4_6 %>% 
  select(r4nettotw_bu_s, r4totwq5_bu_s) %>%                                                         
  mutate(
    r4nettotw_bu_s = as.integer(r4nettotw_bu_s),
    r4totwq5_bu_s = as.integer(r4totwq5_bu_s)
  ) %>% 
  na.omit() %>%
  mutate(wave = "Wave 4")

# Combine data
plot_data <- bind_rows(
  plot_dataw1 %>% 
    rename(
      nettotw_bu_s = r1nettotw_bu_s,
      totwq5_bu_s = r1totwq5_bu_s
    ),
  plot_dataw4 %>% 
    rename(
      nettotw_bu_s = r4nettotw_bu_s,
      totwq5_bu_s = r4totwq5_bu_s
    )
)

ggplot(plot_data, aes(x = as.factor(totwq5_bu_s), y = nettotw_bu_s , fill = wave)) +
  geom_boxplot() +
  scale_y_continuous(trans=scales::pseudo_log_trans(base = 10),
                     labels = scales::comma,
                     breaks = c(-10^4,-10^3,-10^2,-10,0,10,10^2,10^3,10^4,10^5,10^6,10^7)) +
  labs(title = "Wave1 (HSE) vs wave4: Distribution of Wealth by Quintile",
       x = "Quintile (poorer->wealthier)",
       y = "Wealth £ (log scale) ",
       caption = paste0("Wave 1 (HSE) n =",nrow(plot_dataw1),
                        "; Wave 4 n = ",nrow(plot_dataw4)))+
  theme_minimal()


```

### Education

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

# Prepare data for both waves
education_dataw1 <- H_elsa_w1 %>%
    group_by(raeduc_e) %>%
    summarise(count = n()) %>%
    mutate(
      category = case_when(
        is.na(raeduc_e)  ~ "Missing",
        raeduc_e == 1    ~ "Less than upper secondary",
        raeduc_e == 2    ~ "other education", # this was originally category "-15"
        raeduc_e == 3    ~ "Secondary",
        raeduc_e == 4    ~ "Some college",
        raeduc_e == 5    ~ "College and above",
        TRUE             ~ "Unknown"
      ),
      percentage = count / sum(count) * 100,
      wave = "Wave 1"
  )

education_dataw4 <- H_elsa_w4_6 %>%
    group_by(raeduc_e) %>%
    summarise(count = n()) %>%
    mutate(
      category = case_when(
        is.na(raeduc_e)  ~ "Missing",
        raeduc_e == 1    ~ "Less than upper secondary",
        raeduc_e == 2    ~ "other education", # this was originally category "-15"
        raeduc_e == 3    ~ "Secondary",
        raeduc_e == 4    ~ "Some college",
        raeduc_e == 5    ~ "College and above",
        TRUE             ~ "Unknown"
      ),
      percentage = count / sum(count) * 100,
      wave = "Wave 4"
  )

# Combine data
education_data <- bind_rows(education_dataw1, education_dataw4) %>%
  group_by(wave) %>%
  mutate(
    percentage = count / sum(count) * 100,
    category = factor(
      category, 
      levels = c(
        "Less than upper secondary", 
        "other education", # this was originally category "-15"
        "Secondary",
        "Some college", 
        "College and above",
        "Missing"
      )
    )
  )

edu_plot<-ggplot(education_data, aes(x = category, y = percentage, fill = wave)) +
  geom_bar(
    stat = "identity", 
    position = position_dodge(width = .9),
    alpha = 0.7
  ) +
  scale_fill_brewer(palette = "Paired")+
  geom_text(
    aes(
      label = paste0(round(percentage,1),"%"),
      group = wave
    ),
    position = position_dodge(width = 1),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "Education Level Distribution: Wave 1 (HSE) vs Wave 4",
    x = "Education Category",
    y = "Percentage of Respondents",
    caption = paste0("Wave 1 (HSE) n =",nrow(H_elsa_w1),
                     "; Wave 4 n = ",nrow(H_elsa_w4_6))
  ) +
  theme_minimal() +
  scale_y_continuous(labels = function(x) paste0(x, "%"))+
  scale_x_discrete(guide = guide_axis(angle =  90))

edu_plot
```

### Occupational status

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
work_plot<-binary_var_chart(
  data_w1 = H_elsa_w1, 
  data_w4 = H_elsa_w4_6, 
  var_w1 = "r1work", 
  var_w4 = "r4work",
  positive_code = 0,
  negative_code = 1,
  var_name = "Work Status", 
  positive_label = "Not Working", 
  negative_label = "Working for Pay"
)

work_plot

```

## Additional variables (smoking, alcohol)

### Smoking

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
smoking_plot <- binary_var_chart(
  data_w1 = H_elsa_w1, 
  data_w4 = H_elsa_w4_6, 
  var_w1 = "r1smoken", 
  var_w4 = "r4smoken",
  positive_code = 1,
  negative_code = 0,
  var_name = "Ever having smoked.", 
  positive_label = "Having smoked", 
  negative_label = "Never Smoked"
)

smoking_plot

```

The prevalence of smoking in patients with chronic pain has not declined when compared to the general population; higher prevalence has been consistently observed across commonly occurring pain diagnoses including fibromyalgia, back pain, and headache [@orhurhu2015prevalence].

### Alcohol

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

binwidth=.3
NA_count<-H_elsa_w4_6$r4drinkwn_e %>%  is.na() %>% sum
NA_perc<- round(NA_count/nrow(H_elsa_w4_6) *100,1)
drinking_plot <-ggplot(H_elsa_w4_6, aes(x = r4drinkwn_e , 
                        fill = factor(ragender, 
                                      labels = c("male","female")))) +
    geom_histogram(binwidth = binwidth,
                   position = position_dodge2(preserve = "single", width = 0.25),
                  alpha = 0.8) +
    scale_x_continuous(trans=scales::pseudo_log_trans(base = 10), 
                       breaks = c(0,1,2,4,8,16,32,64,128))+
    theme_minimal() +
    labs(
      title = "Distribution of number of drinks/week by gender at baseline (w4)",
      subtitle = "x axis represent the count of beer, wine, spirit drunk each week",
      x = "N. of beer, wine, spirit drunk each week (log scale)",
      y = "Frequency",
      fill = "Gender",
      caption = paste0("Wave 4 n = ", nrow(H_elsa_w4_6),
                        "; missing values n = ", NA_count, 
                       " (",NA_perc,"%)"))+
    theme_minimal() 

drinking_plot

```

## Exposure: partnership and spousal support

### Partnership status

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
Partnership_plot <- binary_var_chart(
  data_w1 = H_elsa_w1, 
  data_w4 = H_elsa_w4_6, 
  var_w1 = "r1partner", 
  var_w4 = "r4partner",
  positive_code = 1,
  negative_code = 0,
  var_name = "Having a partner.", 
  positive_label = "Have a partner/spouse", 
  negative_label = "Does not have a partner/spouse"
)

Partnership_plot

```

### Spousal support score

```{r ssupport-summary-table, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
# Reshape wide → long: one row per person-wave, partnered observations only
ssup_long <- H_elsa_w4_6 %>%
  select(idauniq, ragender,
         r4partner, r5partner, r6partner,
         r4ssupport6, r5ssupport6, r6ssupport6) %>%
  pivot_longer(
    cols      = c(r4ssupport6, r5ssupport6, r6ssupport6),
    names_to  = "wave",
    values_to = "ssupport6"
  ) %>%
  mutate(
    wave    = recode(wave,
                     r4ssupport6 = "Wave 4",
                     r5ssupport6 = "Wave 5",
                     r6ssupport6 = "Wave 6"),
    partner = case_when(
      wave == "Wave 4" ~ r4partner,
      wave == "Wave 5" ~ r5partner,
      wave == "Wave 6" ~ r6partner
    )
  ) %>%
  filter(partner == 1) %>%
  select(idauniq, ragender, wave, ssupport6)

ssup_summary <- ssup_long %>%
  group_by(wave) %>%
  summarise(
    n         = n(),
    n_missing = sum(is.na(ssupport6)),
    pct_miss  = round(n_missing / n * 100, 1),
    med       = median(ssupport6, na.rm = TRUE),
    q25       = quantile(ssupport6, 0.25, na.rm = TRUE),
    q75       = quantile(ssupport6, 0.75, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(
    `Median [IQR]`  = paste0(round(med, 2), " [", round(q25, 2), ", ", round(q75, 2), "]"),
    `Missing n (%)`  = paste0(n_missing, " (", pct_miss, "%)")
  ) %>%
  select(Wave = wave, `n (partnered)` = n, `Median [IQR]`, `Missing n (%)`)

kable(ssup_summary,
      caption = "Perceived spousal support score (RwSSUPPORT6) at each wave, among partnered participants. Higher scores indicate greater strain (range 1–4). Median = 1.5 at all three waves.")
```

```{r ssupport-distribution-plot, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
wave_medians <- ssup_long %>%
  group_by(wave) %>%
  summarise(med = median(ssupport6, na.rm = TRUE), .groups = "drop")

ggplot(ssup_long, aes(x = ssupport6,
                      fill = factor(ragender, labels = c("Male", "Female")))) +
  geom_histogram(binwidth = 0.2,
                 position = position_dodge2(preserve = "single", width = 0.25),
                 alpha = 0.8) +
  geom_vline(data = wave_medians, aes(xintercept = med),
             linetype = "dotted", colour = "blue") +
  facet_wrap(~ wave, ncol = 3) +
  scale_x_continuous(breaks = seq(1, 4, 0.5)) +
  theme_minimal() +
  labs(
    title    = "Perceived spousal support score by wave and sex (partnered participants)",
    subtitle = "Dotted line = median per wave; higher score = greater strain (range 1–4)",
    x        = "Spousal Support Score",
    y        = "Frequency",
    fill     = "Sex"
  )
```

## Clinical and health factors

### Sleep

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
sleep_plot <- binary_var_chart(
  data_w1 = H_elsa_w1, 
  data_w4 = H_elsa_w4_6, 
  var_w1 = "r1sleepr", 
  var_w4 = "r4sleepr",
  positive_code = 1,
  negative_code = 0,
  var_name = "Sleep is restless.", 
  positive_label = "Restless sleep", 
  negative_label = "No restless sleep"
)

sleep_plot

```

### BMI

*Moderate obesity* defined as BMI >= 30 [@andreyeva2007obesity]. Missing values arise when a nurse visit did not take place.

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

binwidth=3
NA_count<-H_elsa_w4_6$r4mbmi %>%  is.na() %>% sum
NA_perc<- round(NA_count/nrow(H_elsa_w4_6) *100,1)
MedianBMI<- median(H_elsa_w4_6$r4mbmi, na.rm = T)
BMI_plot <-ggplot(H_elsa_w4_6, aes(x = r4mbmi , 
                        fill = factor(ragender, 
                                      labels = c("male","female")))) +
    geom_histogram(binwidth = binwidth,
                   position = position_dodge2(preserve = "single", width = 0.25),
                  alpha = 0.8) +
    scale_x_continuous(breaks = seq(0,60,5))+
    geom_vline(xintercept = MedianBMI, linetype="dotted", color = "blue")+
    theme_minimal() +
    labs(
      title = "Distribution of Body Mass Index (BMI) by gender at baseline (w4)",
      subtitle = paste0("Dotted line indicate median BMI = ", round(MedianBMI,2)),
      x = "BMI",
      y = "Frequency",
      fill = "Gender",
      caption = paste0("Wave 4 n = ", nrow(H_elsa_w4_6),
                        "; missing values n = ", NA_count, 
                       " (",NA_perc,"%)"))+
    theme_minimal() 

BMI_plot
```

### Physical activity

Participants were asked to self-report the frequency of their
participation in sports or activities that were vigorous, moderately
energetic and mildly energetic (more than once a week, once a week, one
to three times a month, hardly ever or never). Physical activity was
then categorised into four groups in accordance with previous literature
[@dhalwani2016long; @malkowski2023socioeconomic]: (1) inactive (no
activity on a weekly basis); (2) only mild activity at least once per
week; (3) at least moderate but no vigorous activity at least once per
week; and (4) vigorous activity at least once per week.

```{r}


# Create a data frame with percentages by category and gender
PA_data <- H_elsa_w4_6 %>%
  filter(!is.na(r4PA_cat)) %>%
  group_by(ragender, r4PA_cat) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ragender) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(category = factor(r4PA_cat,
                          levels = 1:4,
                          labels = c("Inactive", 
                                     "Only mild activity",
                                     "At least moderate but not vigorous activity",
                                     "At least one vigorous activity every week")))

# Calculate missing data count
missing_PA_count <- sum(is.na(H_elsa_w4_6$r4PA_cat))

# Create the plot
plot_PA_data<-ggplot(PA_data, aes(x = category, y = percentage, fill = factor(ragender, labels = c("male", "female")))) +
  geom_bar(
    stat = "identity", 
    position = position_dodge(width = 0.9),
    alpha = 0.7
  ) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", percentage),
      group = factor(ragender, labels = c("male", "female"))
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "Physical Activity Categories by Gender at Baseline (Wave 4)",
    x = "Physical Activity Category",
    y = "Percentage of Respondents",
    fill = "Gender",
    caption = paste0("Participants with missing PA data n= ", missing_PA_count, " of N = ", nrow(H_elsa_w4_6))
  ) +
  theme_minimal() +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_PA_data
```

### Depression (CES-D 8)

Depression is strongly linked with chronic pain [@mills2019chronic]. The CES-D-8 is a valid and reliable measure of depressive symptoms in community-dwelling older people [@briggs2018validation]; a score of >= 3 is generally accepted as the threshold for caseness [@turvey1999revised; @white2016duration].

```{r}

binwidth=1
NA_count<-H_elsa_w4_6$r4cesd %>%  is.na() %>% sum
NA_perc<- round(NA_count/nrow(H_elsa_w4_6) *100,1)
MedianCESD<- median(H_elsa_w4_6$r4cesd, na.rm = T)
Threshold_CESD<-3
CESD_plot <-ggplot(H_elsa_w4_6, aes(x = r4cesd , 
                        fill = factor(ragender, 
                                      labels = c("male","female")))) +
    geom_histogram(binwidth = binwidth,
                   position = position_dodge2(preserve = "single", width = 0.25),
                  alpha = 0.8) +
    scale_x_continuous(breaks = seq(0,7,1))+
    geom_vline(xintercept = Threshold_CESD, linetype="dashed", color = "red")+
    theme_minimal() +
    labs(
      title = "Distribution of CES-D 8 scores by sex at baseline (Wave 4)",
      subtitle = paste0("Dotted line = median CES-D 8 = ", round(MedianCESD,2),"; dashed line = caseness threshold (score >= 3)"),
      x = "CESD",
      y = "Frequency",
      fill = "Gender",
      caption = paste0("Wave 4 n = ", nrow(H_elsa_w4_6),
                        "; missing values n = ", NA_count,
                       " (", NA_perc, "%)"))+
    theme_minimal() 

CESD_plot
```

### MLTC

Count of doctor-diagnosed conditions. Full item list in Section 2.3.

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

################################ Not adopted
# additional categories used in [@torres2016wealth]
  # rXparkine: Parkinson
  # cardiovascular diseases (
  #  rxhibpe   : high blood pressure, 
  #  rXangine  : angina, 
  #  rXhrtatte : heart attack, 
  #  rXconhrtfe: heart failure, 
  #  rXhrtmre  : heart murmur 
  #  rXhrtrhme : heart rhythm)

# additional categories used in [@simpson2023eliciting]:
  # rxhibpe    : hypertension | high blood pressure
  # rXasthmae  : asthma, 
  # rXpsyche   : mental health disorders (diagnosed: depression, anxiety, emotional problems, hallucinations, schizophrenia, psychosis) 
  # rxalzhe    : Alzheimer's disease, 
  # rXdemene   : dementia 
################################## Not adopted

MLTC_plot <-ggplot(H_elsa_w4_6, aes(x = r4MLTC , 
                        fill = factor(ragender, 
                                      labels = c("male","female")))) +
    geom_histogram(binwidth = binwidth,
                   position = position_dodge2(preserve = "single", width = 0.25),
                  alpha = 0.8) +
    scale_x_continuous(breaks = seq(0,5,1))+
   theme_minimal() +
    labs(
      title = "Distribution of MLTC count by gender at baseline (w4)",
      subtitle = paste0(""),
      x = "N of MLTC",
      y = "Frequency",
      fill = "Gender",
      caption = paste0("Wave 4 n = ", nrow(H_elsa_w4_6), ". missing = ", sum(is.na(H_elsa_w4_6$r4MLTC)))
      )+
    theme_minimal() 

MLTC_plot
```

#### MLTC in Wave 4 and Wave 1 (HSE)

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

wave1 <- H_elsa_w1 %>% 
  select(r1MLTC) %>% 
  mutate(MLTC = r1MLTC, wave = "Wave 1 (HSE)") %>%
  group_by(MLTC) %>%
  summarise(count = n(), wave = "Wave 1 (HSE)")

wave4 <- H_elsa_w4_6 %>% 
  select(r4MLTC) %>% 
  mutate(MLTC = r4MLTC, wave = "Wave 4") %>%
  group_by(MLTC) %>%
  summarise(count = n(), wave = "Wave 4")

plot_data <- bind_rows(wave1, wave4) %>%
  group_by(wave) %>%
  mutate(percentage = count / sum(count) * 100)
    

ggplot(plot_data, aes(x = MLTC, y = percentage, fill = wave)) +  # Changed y to percentage
  geom_bar(
    stat = "identity", 
    position = position_dodge(width = 0.9),
    alpha = 0.7
  ) +
  scale_fill_brewer(palette = "Paired")+
  geom_text(
    aes(
      label = sprintf("%.1f%%", percentage),  # Changed label format
      group = wave
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "Multiple long-term conditions (MLTC): Wave 1 (HSE) vs Wave 4",
    x = "Number of Conditions",
    y = "Percentage of Respondents",  # Changed y-axis label
    fill = "Wave",
    caption = paste0("Wave 1 n = ", nrow(H_elsa_w1), "; Wave 4 n = ", nrow(H_elsa_w4_6))
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%"))+  # Added percentage formatting  
  scale_x_continuous(breaks = c(0,1,2,3,4,5)) +
    theme_minimal()


plotMLTCxAgexGender <- ggpairs(
  H_elsa_w4_6,
  title = "Age x Gender x count of Long Term Conditions (MLTC)",
  mapping = aes(group = as.factor(ragender), 
                fill = as.factor(ragender), 
                color = as.factor(ragender),
                alpha = .6),
  columns = c("r4MLTC", "r4agey")
)

plotMLTCxAgexGender

```

### Number of pain locations

-   hepawba *Whether feel pain in back*
-   hepawhi *Whether feel pain in hips*
-   hepawkn *Whether feel pain in knees*
-   hepawfe *Whether feel pain in feet*
-   hepawmo *Whether feel pain in mouth or teeth*
-   hepawot *Whether feel pain elsewhere*
-   hepawal *Whether feel pain all over*

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}

PainlocN_data <- H_elsa_w4_6 %>%
  group_by(pain_locN = r4pain_locN, gender = factor(ragender, labels = c("Male", "Female"))) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(gender) %>%
  mutate(percentage = count / sum(count) * 100)


ggplot(PainlocN_data, aes(x = pain_locN, y = percentage, fill = gender)) +  # Changed y to percentage
  geom_bar(
    stat = "identity", 
    position = position_dodge(width = 0.9),
    alpha = 0.7
  ) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", percentage),  # Changed label format
      group = gender
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "Number of pain locations by gender",
    x = "Number of Pain locations",
    y = "Percentage of Respondents",  # Changed y-axis label
    fill = "Wave",
    caption = paste0("Wave 4 n = ", nrow(H_elsa_w4_6))
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%"))+  # Added percentage formatting  
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7)) +
    theme_minimal()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
# Local recoding for descriptive display only — not pipeline derivations.
# r4wealthBIN: binary from raw pre-imputation quintile (r4totwq5_bu_s) for Table 1.
# r4mbmiBIN: BMI category for descriptive plot. Neither variable is saved.
H_elsa_w4_6 <- H_elsa_w4_6 %>%
  mutate(
    r4wealthBIN = case_when(
      r4totwq5_bu_s == 1 ~ "most deprived Q",
      r4totwq5_bu_s >  1 ~ "2:5 least deprived Q",
      TRUE ~ NA_character_
    ) %>% as.factor(),
  )

table(H_elsa_w4_6$r4wealthBIN)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
H_elsa_w4_6 <- H_elsa_w4_6 %>% mutate(
  r4mbmiBIN = case_when(
    is.na(r4mbmi)  ~ "No nurse data",
    r4mbmi >= 30   ~ "moderate obesity",
    r4mbmi <  30   ~ "sub threshold")
)

plot_data <- H_elsa_w4_6 %>%
    group_by(r4mbmiBIN, ragender) %>%
    summarise(count = n()) %>%
    group_by(ragender) %>%
    mutate(
      percentage = count / sum(count) * 100,
      category = factor(
        r4mbmiBIN,
        levels = c("moderate obesity", "sub threshold", "No nurse data")
      )
    )

missing_nurse_count <- sum(H_elsa_w4_6$r4mbmiBIN == "No nurse data", na.rm = TRUE)

BMI_binplot <- ggplot(plot_data, aes(x = category, y = percentage, fill = factor(ragender, labels = c("male", "female")))) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9),
    alpha = 0.7
  ) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", percentage),
      group = factor(ragender, labels = c("male", "female"))
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "BMI categories by sex at baseline (Wave 4)",
    x = "BMI Category",
    y = "Percentage of Respondents",
    fill = "Sex",
    caption = paste0("Participants missing nurse data n= ", missing_nurse_count, " of N = ", nrow(H_elsa_w4_6))
  ) +
  theme_minimal() +
  scale_y_continuous(labels = function(x) paste0(x, "%"))

BMI_binplot
```

### Validity of pain location count as a subdivision of chronic pain

The use of pain location count (`r4pain_locN_cat`) to subdivide chronic pain into localised and multisite categories was validated by examining its association with the chronic pain indicator (`r4painchr`). A high polychoric correlation between the two would confirm that the location-based subdivision captures a meaningful gradation within the chronic pain group rather than introducing noise.

```{r}
painLoc_data <- H_elsa_w4_6 %>%
  select(
    r4pain_locN_cat,
    r4painchr) %>%
  # Remove NA values
  filter(!is.na(r4painchr))

fact_var<-c("r4painchr",
            "r4pain_locN_cat"
            )

painLoc_data<-painLoc_data%>% mutate(across(all_of(fact_var), factor))

str(painLoc_data)

```

Both variables are ordinal; polychoric correlation is therefore used.

```{r}

# hetcor computes a heterogenous correlation matrix, consisting of: 
# * Pearson product-moment correlations between numeric variables, 
# * polyserial correlations between numeric and ordinal variables, 
# * polychoric correlations between ordinal variables

#str(cor_df_out)
cor_df_out<-as.data.frame(painLoc_data) # because hetcor only takes dataframes and not tibbles
het_cor_out_cat <- hetcor(cor_df_out%>%select(r4pain_locN_cat,r4painchr), 
                          use = c("pairwise.complete.obs"),
                          ML = TRUE,
                          ncores=detectCores(logical=T))


het_cor_out_cat

table(painLoc_data$r4pain_locN_cat,painLoc_data$r4painchr)

```

The polychoric correlation is 0.963 (p < 0.001), indicating near-perfect ordinal agreement between location count category and chronic pain status. The contingency table confirms complete separation between pain-free respondents and those with any chronic pain, with location count differentiating within the chronic pain group. Polychoric correlation is robust to moderate violations of bivariate normality.

### Distribution of `painchr_ext`

```{r}
NA_count_pain<-is.na(H_elsa_w4_6$r4painchr_ext) %>% sum    
NA_perc_pain<-NA_count_pain/nrow(H_elsa_w4_6) * 100               
# r4painchr_ext extends r4painchr by splitting chronic pain (class 2) into localised vs widespread (3+ locations)

ggplot(H_elsa_w4_6, aes(x = r4painchr_ext, 
                       fill = factor(ragender, 
                                    labels = c("Male", "Female")))) +
  geom_bar(position = position_dodge(width = 0.9),
           alpha = 0.8) +
  geom_text(aes(label = after_stat(count)), 
            stat = "count",
            size = 3,
            position = position_dodge(width = 0.9),
            vjust = -0.5) +
  theme_minimal() +
  labs(
    title = "Distribution of painchr_ext by sex at baseline (Wave 4)",
    subtitle = "0 = no pain; 1 = acute; 2 = localised chronic (≤2 locations); 3 = multisite chronic (≥3 locations)",
    x = "Pain state Category",
    y = "Frequency",
    fill = "Gender",
    caption = paste0("Wave 4 n = ",nrow(H_elsa_w4_6), 
                   "; missing data n = ", NA_count_pain, 
                   " (", round(NA_perc_pain,1), "%)")
  )
    
    
```


### Distribution of `Health impact`

```{r}

plot_data<- H_elsa_w4_6 %>% filter(!is.na(r4painchr_ext))
plot_data<-plot_data %>% mutate(r4lvimp_shifted = r4lvimp - min(r4lvimp))

ggplot(plot_data, 
       aes(x = r4lvimp_shifted,
           fill = factor(r4painchr_ext))) +
  geom_density(alpha = 0.8) +
  theme_minimal() +
  labs(title = "Distribution of Health impact scores by chronic pain status at baseline (Wave 4)",
       subtitle = "0 = no pain; 1 = acute; 2 = localised chronic (≤2 locations); 3 = multisite chronic (≥3 locations)",
       x = "Health impact IRT score",
       y = "Density",
       fill = "Chronic Pain Status",
       caption = paste0("Wave 4 n = ",nrow(H_elsa_w4_6),
                        "; missing data n = ", NA_count_pain,
                        " (", round(NA_perc_pain,1), "%)")) 




```

# Sample Characteristics — Table 1

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
# r4lvimp, r4pain_locN_cat, r4painchr_ext: pass-through from H_elsa_w4_6.rds
# derived in suppl_pt1_Data_prep.Rmd:
#   r4lvimp          — IRT pain impact score (pt1 Section "IMPORT impact IRT data")
H_elsa_w4_6$r4lvimp_shifted<-H_elsa_w4_6$r4lvimp - min(H_elsa_w4_6$r4lvimp)
#   r4pain_locN_cat  — pain location count category (pt1 Section "N of Pain Locations")
#   r4painchr_ext    — extended 4-category classifier (pt1 Section "Composite Chronic pain state + location")
var_Ys<-c("r4painchr","r4lvimp_shifted")

var_Xs<-c("r4ssupport6","r4partner","r4pain_locN_cat","r4pain_locN")

var_Covs<-c("r4agey","r4agey_cat",
              "ragender",
              "raracem",
              "r4cesd",
              "r4nettotw_bu_s","r4totwq5_bu_s","r4wealthBIN",
              "raeduc_e","r4work",
              "r4smoken","r4drinkwn_e","r4lnlys","r4mbmi","r4MLTC")

# Prepare data — note: r4MLTC_cat is created inside prepare_table1_data() from r4MLTC (see Paper1_func.R)
table1_data <- prepare_table1_data(H_elsa_w4_6, var_Ys, var_Xs, var_Covs)

# Create the table
table1_1 <- table1(~ r4lvimp_shifted + r4pain_locN_cat + r4cesd + r4agey + ragender + raracem +
                      r4partner + r4totwq5_bu_s + r4wealthBIN + raeduc_e + r4work +
                      r4smoken + r4drinkwn_e + r4mbmi + r4lnlys+ r4MLTC_cat |
                     r4painchr,
                    data = table1_data,
                    render.continuous = my_render_continuous,
                    render.categorical = my_render_categorical,
                    overall = "Total",
                    topclass = "Rtable1-times")


kable(as.data.frame(table1_1),
      caption = paste0(
        "Table 1. Wave 4 analytic sample characteristics by observed pain category. ",
        "n = 7,963 (51 participants excluded due to unclassifiable pain status [missing r4painfr]). ",
        "Full analytic sample n = 8,014. Missing proportions are based on n = 7,963."
      ))


# Restrict to partnered only so missing % denominator = n partnered, not full sample
table1_2 <- table1(~ r4ssupport6 | r4painchr,
                    data = table1_data %>% filter(r4partner == "With partner"),
                    render.continuous = my_render_continuous,
                    render.categorical = my_render_categorical,
                    overall = "Total",
                    topclass = "Rtable1-times")



kable(as.data.frame(table1_2))

```

# Wave 4 vs Wave 1 (HSE) comparison — Table 2

Wave 1 of ELSA was drawn directly from the Health Survey for England (HSE) 2001 and is therefore approximately representative of the English population aged 50+. Wave 4 is our analytic sample (participants completing all three waves). Greater similarity between the two samples supports the external validity of our findings. Chronic pain prevalence is not compared here as the HSE did not include an equivalent measure.

```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}

# Build harmonised dataset from both waves using only variables present in both
hse_sub <- H_elsa_w1 %>%
  transmute(
    Wave      = factor("Wave 1 (HSE)", levels = c("Wave 1 (HSE)", "Wave 4")),
    Age       = r1agey,
    Age_cat   = r1agey_cat,
    Sex       = factor(ragender, levels = c(1, 2), labels = c("Male", "Female")),
    Education = factor(raeduc_e, levels = 1:5,
                       labels = c("Less than secondary", "Other", "Secondary",
                                  "Some college", "College and above")),
    WealthQ   = factor(r1totwq5_bu_s, levels = 1:5,
                       labels = c("Q1 (most deprived)", "Q2", "Q3", "Q4",
                                  "Q5 (least deprived)")),
    MLTC      = factor(case_when(
                  is.na(r1MLTC) ~ NA_character_,
                  r1MLTC == 0   ~ "No conditions",
                  r1MLTC == 1   ~ "One condition",
                  r1MLTC == 2   ~ "Two conditions",
                  r1MLTC >= 3   ~ "Three or more"
                ), levels = c("No conditions", "One condition",
                              "Two conditions", "Three or more")),
    Employed  = factor(r1work,   levels = c(0, 1), labels = c("Not working", "Working")),
    Smoker    = factor(r1smoken, levels = c(0, 1), labels = c("Non-smoker", "Current smoker")),
    Partner   = factor(r1partner,levels = c(0, 1), labels = c("Without partner", "With partner")),
    Race      = factor(raracem,  levels = c(1, 4),  labels = c("White", "Non-White")),
    # r1cesd not cleaned in pt1 (deprVars starts r3); apply guard here
    CESD      = if_else(r1cesd < 0, NA_real_, as.double(r1cesd)),
    # Loneliness: add when r1lnlys confirmed available — if_else(r1lnlys < 0, NA_real_, as.double(r1lnlys))
    # r1ssupport6 clean in saved rds (min=1, NAs are genuine non-partnered)
    Spousal_support = r1ssupport6,
    # r1painlv has negative ELSA codes (-18, -16, -4); exclude here
    Pain_level = factor(if_else(r1painlv < 0, NA_integer_, as.integer(r1painlv)),
                        levels = c(0, 1, 2, 3),
                        labels = c("Not troubled", "Mild", "Moderate", "Severe"))
  )

w4_sub <- H_elsa_w4_6 %>%
  transmute(
    Wave      = factor("Wave 4", levels = c("Wave 1 (HSE)", "Wave 4")),
    Age       = r4agey,
    Age_cat   = r4agey_cat,
    Sex       = factor(ragender, levels = c(1, 2), labels = c("Male", "Female")),
    Education = factor(raeduc_e, levels = 1:5,
                       labels = c("Less than secondary", "Other", "Secondary",
                                  "Some college", "College and above")),
    WealthQ   = factor(r4totwq5_bu_s, levels = 1:5,
                       labels = c("Q1 (most deprived)", "Q2", "Q3", "Q4",
                                  "Q5 (least deprived)")),
    MLTC      = factor(case_when(
                  is.na(r4MLTC) ~ NA_character_,
                  r4MLTC == 0   ~ "No conditions",
                  r4MLTC == 1   ~ "One condition",
                  r4MLTC == 2   ~ "Two conditions",
                  r4MLTC >= 3   ~ "Three or more"
                ), levels = c("No conditions", "One condition",
                              "Two conditions", "Three or more")),
    Employed  = factor(r4work,   levels = c(0, 1), labels = c("Not working", "Working")),
    Smoker    = factor(r4smoken, levels = c(0, 1), labels = c("Non-smoker", "Current smoker")),
    Partner   = factor(r4partner,levels = c(0, 1), labels = c("Without partner", "With partner")),
    Race      = factor(raracem,  levels = c(1, 4),  labels = c("White", "Non-White")),
    # r4cesd cleaned in pt1 (deprVars r3:6); r4ssupport6 intentionally not cleaned — guard here
    CESD      = r4cesd,
    # Loneliness: add when r1lnlys confirmed — Loneliness = r4lnlys
    Spousal_support = if_else(r4ssupport6 >= 0, as.double(r4ssupport6), NA_real_),
    # r4painlv already clean (values 0–3 only)
    Pain_level = factor(r4painlv,
                        levels = c(0, 1, 2, 3),
                        labels = c("Not troubled", "Mild", "Moderate", "Severe"))
  )

hse_w4_df <- bind_rows(hse_sub, w4_sub)

tab2 <- CreateTableOne(
  vars       = c("Age", "Age_cat", "Sex", "Race", "Education", "WealthQ",
                 "MLTC", "Employed", "Smoker", "Partner",
                 "CESD", "Spousal_support", "Pain_level"),
  factorVars = c("Age_cat", "Sex", "Race", "Education", "WealthQ",
                 "MLTC", "Employed", "Smoker", "Partner", "Pain_level"),
  strata     = "Wave",
  data       = hse_w4_df,
  addOverall = FALSE
)

# SMD only, no p-values
# |SMD| < 0.10 = negligible imbalance; 0.10-0.20 = small; > 0.20 = meaningful
n_partnered_w1 <- sum(hse_sub$Partner == "With partner", na.rm = TRUE)
n_partnered_w4 <- sum(w4_sub$Partner  == "With partner", na.rm = TRUE)

tab2_mat <- print(tab2,
                  smd         = TRUE,
                  test        = FALSE,
                  printToggle = FALSE,
                  noSpaces    = TRUE)

# Rename spousal support row so the label signals it is a conditional statistic
rownames(tab2_mat) <- sub("Spousal_support",
                          "Spousal support (partnered)",
                          rownames(tab2_mat))

kable(tab2_mat,
      caption = paste0(
        "Table 2. Wave 4 analytic sample vs Wave 1 (HSE) reference population. ",
        "SMD = standardised mean difference; |SMD| < 0.10 indicates negligible imbalance. ",
        "Wave 1 (HSE) n = ", nrow(hse_sub), "; Wave 4 n = ", nrow(w4_sub), ". ",
        "MLTC: count of doctor-diagnosed long-term conditions (angina, stroke, diabetes, ",
        "pulmonary disease, arthritis, cancer, mental health disorders; Poole et al., 2020). ",
        "CESD: 8-item CES-D sum score (0–8). ",
        "Spousal support: marital quality scale (1–4), reported among those with a partner ",
        "(Wave 1 n = ", n_partnered_w1, "; Wave 4 n = ", n_partnered_w4, ")."
      ))
```

