---
title: "Multiple imputation Mice"
author: "Diego Vitali"
output: 
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib

---
```{r setup, eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(lavaan)
library(tidyr)
library(dplyr)
library(ggplot2)
library(naniar)
library(NbClust)
library(readr)
library(polycor)
library(patchwork)
library(corrplot)
library(GGally)
library(lme4)
library(broom.mixed)

source("Sources/Paper1_func.R")

paper1_rds_path<-"~/private_WP5/WP5_data/rds/"

H_elsa_w4_6<-readRDS(paste0(paper1_rds_path,"H_elsa_w4_6.rds"))

paper1_MODELS_rds_path<-"~/private_WP5/WP5_data/model_fits/"
# refreshments: 3, 4, 6, 7, 9 and 10

# Nurse Visits: 2,4,6,8,9

```

This documentation describes our R-based pipeline for handling missing data using 
Multiple Imputation by Chained Equations (MICE). 

This pipeline operates on data in the long data format (one row per participant per wave, with a time/wave indicator).

*Features:*

* Conditional imputation: The code addresses conditional imputation scenarios, such as imputing "Marital strain" only for partnered individuals. This is implemented via the where_matrix in MICE, ensuring that imputation is only performed where appropriate.
* Data types: Variables are explicitly cast to factor (with or without ordering) or numeric  types as required. This is essential for MICE to select the correct imputation model and for downstream analyses.
* Multicollinearity: The script includes steps for checking multicollinearity using heterogeneous correlation matrices (polycor::hetcor). This is a good practice, as highly collinear predictors can destabilize imputation models.
* Minimal Missingness handling: For variables with very few missing values, simple imputation (median for numeric, mode for categorical) is performed before running MICE. This is efficient and avoids unnecessary complexity for almost-complete variables.
* Missingness visualisation: The use of naniar::vis_miss to visualize missingness patterns before and after imputation is excellent for transparency and quality control.

## Age and immortal cohort assumption

We assume no further death event than those recorded in ELSA and we take the baseline age as a timeinvariant predictor 

```{r}

H_elsa_w4_6<- H_elsa_w4_6%>% mutate(
                            r5agey = case_when(
                                    is.na(r5agey) ~ r4agey + 2,
                                    TRUE ~ r5agey
                                    ),
                            r6agey = case_when(
                                    is.na(r6agey) ~ r5agey + 2,
                                    TRUE ~ r6agey
                                    )
                            )
                            

H_elsa_w4_6$raagey<- H_elsa_w4_6$r4agey
      

```


## defining the imputation dataset

Select relevant variables: 
(race is excluded from the auxillary variable because sample is 97% white)
```{r}
Missingdata_wide <- H_elsa_w4_6 %>%
  select(c("idauniq",
          #varnames                          # description    (type); whether (comp)lete or (inc)omplete data
           paste0("r",4:6,"painlv"),         #                (categorical); incomp 
           paste0("r",4:6,"painchr_ext"),    #                (categorical); incomp 
           paste0("r",4:6,"lvimp"),          #                (categorical); incomp 
           paste0("r",4:6,"shlt"),           # self reported health               (categorical); incomp 
           paste0("r",4:6,"adltot6"),        # ADL total score                    (numeric); incomp 
           paste0("r",4:6,"partner"),        # having a partner                   (binary); comp 
           paste0("r",4:6,"ssupport6"),      # Marital strain score             (numeric, skewed) incomplete;
           paste0("r",4:6,"lnlys"),          # UCLA - loneliness                (numeric, skewed) incomplete;
           "raagey",                         # Baseline age                     (numeric, normal); comp 
           "ragender",                       # Gender                           (binary); comp 
           #"raracem",                        # Race/ethnicity                   (binary); comp
           paste0("r",4:6,"nettotw_bu_s"),     # Net wealth                     (numeric, skewed); incomp
           "raeduc_e",                       # Education                        (categorical); incomp
           paste0("r",c(4:6),"smoken"),      # smoking                          (binary); incomp
           paste0("r",4:6,"work"),           # Working status                   (binary); incomp
           paste0("r",4:6,"drinkwn_e"),      # Drinking (weekly units)          (continuous); incomp
#           paste0("r",c(4,6),"mbmi"),        # BMI                              (numeric); incomp
#           paste0("r",c(4,6),"hscrp"),       # CRP                              (numeric); incomp
           paste0("r",4:6,"PA_cat"),         # Physical activity                (categorical); incomp
           paste0("r",4:6,"cesd_lat"),       # Depression latent score          (numeric, skewed); incomp
           paste0("r",4:6,"MLTC")            # Multiple long-term condit. count (numeric/count, skewed); incomp
  ))

#str(Missingdata_wide)
```

### Transform skewed variables

Wealth: Shift to positive and log-transform.

Impact and drinking: Shift/log-transform as appropriate.

Create auxiliary variables for change (e.g., MLTC_change_w5, MLTC_change_w6) to 
capture longitudinal change and reduce collinearity.

```{r}
###### TRANSFORM ###################
### Transform wealth data

# Calculate global minimums once
global_min_wealth <- min(c(Missingdata_wide$r4nettotw_bu_s, 
                          Missingdata_wide$r5nettotw_bu_s, 
                          Missingdata_wide$r6nettotw_bu_s), na.rm = TRUE)

global_min_lvimp <- min(c(Missingdata_wide$r4lvimp, 
                          Missingdata_wide$r5lvimp, 
                          Missingdata_wide$r6lvimp), na.rm = TRUE)

global_min_cesd_lat <- min(c(Missingdata_wide$r4cesd_lat, 
                          Missingdata_wide$r5cesd_lat, 
                          Missingdata_wide$r6cesd_lat), na.rm = TRUE)



# Apply same shift to all waves
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4nettotw_bu_s_log = log(r4nettotw_bu_s - global_min_wealth + 1),
    r5nettotw_bu_s_log = log(r5nettotw_bu_s - global_min_wealth + 1),
    r6nettotw_bu_s_log = log(r6nettotw_bu_s - global_min_wealth + 1)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4nettotw_bu_s,-r5nettotw_bu_s,-r6nettotw_bu_s)


### shift imp_data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4lvimp_shifted = r4lvimp - global_min_lvimp,
    r5lvimp_shifted = r5lvimp - global_min_lvimp,
    r6lvimp_shifted = r6lvimp - global_min_lvimp,
  )
# Remove original impact variables

Missingdata_wide<- Missingdata_wide %>%
  select(-r4lvimp,-r5lvimp,-r6lvimp)

### shift depr_lat data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4cesd_lat_shifted = r4cesd_lat - global_min_cesd_lat,
    r5cesd_lat_shifted = r5cesd_lat - global_min_cesd_lat,
    r6cesd_lat_shifted = r6cesd_lat - global_min_cesd_lat,
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4cesd_lat,-r5cesd_lat,-r6cesd_lat)


### Transform drinkning data
#Missingdata_wide$r4drinkwn_e %>% log() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4drinkwn_e_log  = log1p(r4drinkwn_e),
    r5drinkwn_e_log  = log1p(r5drinkwn_e),
    r6drinkwn_e_log  = log1p(r6drinkwn_e)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4drinkwn_e,-r5drinkwn_e,-r6drinkwn_e)


### Transform support data
#Missingdata_wide$r4ssupport6 %>% log1p() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4ssupport6_log1p  = log1p(r4ssupport6),
    r5ssupport6_log1p  = log1p(r5ssupport6),
    r6ssupport6_log1p  = log1p(r6ssupport6)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4ssupport6,-r5ssupport6,-r6ssupport6)


#str(Missingdata_wide)

# wave 4 ssupport missingness
ss4<-vis_miss(Missingdata_wide[which(Missingdata_wide$r4partner == 1),] %>% select(r4ssupport6_log1p))
# wave 5 ssupport missingness
ss5<-vis_miss(Missingdata_wide[which(Missingdata_wide$r5partner == 1),] %>% select(r5ssupport6_log1p))
# wave 5 ssupport missingness
ss6<-vis_miss(Missingdata_wide[which(Missingdata_wide$r6partner == 1),] %>% select(r6ssupport6_log1p))

(ss4 + ss5 + ss6)+
  plot_annotation(
    title = "Missingness in conditional variable `Marital strain` total score"
  )


Missingdata_wide %>% select(-r4ssupport6_log1p,-r5ssupport6_log1p,-r6ssupport6_log1p) %>% vis_miss()

```

## Defining variables type
```{r}

bin.vars<-  c(paste0("r",4:6,"work"),         # currently working (bl only)       (binary); incomp 
              paste0("r",4:6,"partner"),      # having a partner (imp condition)  (binary); comp 
              paste0("r",4:6,"smoken"),       # smoking (bl only)                 (binary); incomp
              "ragender"                     # gender                            (binary); incomp
        #      "raracem"                       # race                              (binary); comp
              )

ord.vars<-c(paste0("r",4:6,"PA_cat"),          # Physical Activity                (cat-ordinal); incomp
#            paste0("r",4:6,"shlt"),           # self-reported health             (cat-ordinal); incomp
            "raeduc_e",                        # education                        (cat-ordinal); incomp
            paste0("r",4:6,"painchr_ext")
            )        


Missingdata_wide<-Missingdata_wide %>% mutate(
                                ragender = case_when(
                                          ragender == 1 ~ 0, # male
                                          ragender == 2 ~ 1, # female
                                          TRUE ~ NA_real_   # unreachable: only codes 1/2 exist in ELSA
                                          )#,
#                                raracem = case_when(
#                                          raracem == 1 ~ 1, # white
#                                          raracem == 4 ~ 0, #non-white
#                                          TRUE ~ NA_real_
#                                          ),
                                )

```
## Variable type definition

```{r}


# categorical variables,# The first step is that all binary and categorical incomplete variables need to be stored as type “factor”;  @wijesuriya2025multiple


Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(ord.vars), ~factor(., ordered = T)))

Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(bin.vars), ~factor(.)))

Missingdata_wide<-as.data.frame(Missingdata_wide)

str(Missingdata_wide)

```


# Minimal Missingmess

For variables with ≤5 missing values (0.06% of total sample N = 8014), impute using median (numeric) or mode (categorical).

While efficient, this approach doesn't propagate imputation uncertainty for these variables and should perhaps be reconsidered (?)
[@azur2011multiple]

```{r eval = T}
# Identify variables with minimal missingness
minimal_missing <- names(which(colSums(is.na(Missingdata_wide)) > 0 & 
                              colSums(is.na(Missingdata_wide)) <= 5))


# Simple imputation for these variables
for(var in minimal_missing) {
  if(is.numeric(Missingdata_wide[[var]])) {
    # For numeric variables: median imputation
    Missingdata_wide[[var]][is.na(Missingdata_wide[[var]])] <- 
      median(Missingdata_wide[[var]], na.rm = TRUE)
  } else if(is.factor(Missingdata_wide[[var]]) || is.character(Missingdata_wide[[var]])) {
    # For categorical variables: mode imputation
    mode_val <- names(sort(table(Missingdata_wide[[var]]), decreasing = TRUE))[1]
    Missingdata_wide[[var]][is.na(Missingdata_wide[[var]])] <- mode_val
  }
}


# Use r6smoken to fill missing values in earlier waves
# Assumption: smoking status is treated as time-stable (ever-smoker); wave 6 is used as
# a last-observation carry-back. This is applied outside MICE and does not propagate uncertainty.
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Fill r4smoken missing values with r6smoken
    # Note: if r6smoken is also NA, the carry-back propagates NA (acceptable;
    # residual missingness handled by MICE).
    r4smoken = case_when(
      !is.na(r4smoken) ~ r4smoken,  # Keep existing values
      is.na(r4smoken) ~ r6smoken    # Fill missing with wave 6
    ),

    # Fill r5smoken missing values with r6smoken
    r5smoken = case_when(
      !is.na(r5smoken) ~ r5smoken,  # Keep existing values
      is.na(r5smoken) ~ r6smoken    # Fill missing with wave 6
    )
  )

#str(Missingdata_wide)
#summary(Missingdata_wide)
```


## wide to long

Define time-invariant variables (ID, gender, race, education).

Use pivot_longer() to convert repeated measures to long format, extracting the wave number and aligning variables accordingly.

Visualize missingness patterns overall and for conditional variables (e.g., Marital Strain for partnered individuals) using naniar::vis_miss.
```{r}

vars_time_invariant <- c("idauniq", "ragender", 
                         #"raracem", 
                         "raeduc_e","raagey")

Missingdata_long <- Missingdata_wide %>%
  pivot_longer(
    cols = -all_of(vars_time_invariant),
    names_to = c("wave", ".value"),
     # Improved pattern that ensures we match the full variable name
    names_pattern = "^r([456])(.+)$") %>%
      # Convert wave to numeric if needed
      mutate(wave = as.numeric(wave))


Missingdata_long  %>% vis_miss()


```



|Variable          |Type                      |Time Status     |Waves Available  |Missing % (Long Format) |
|------------------|--------------------------|----------------|-----------------|---------|
|idauniq           |Integer                   |Structural      |All waves        |Complete |
|ragender          |Binary                    |Time-invariant  |Fixed            |Complete |
|raeduc_e          |Categorical (5 levels)    |Time-invariant  |Fixed            |Variable |
|wave              |Numeric                   |Structural      |4, 5, 6          |Complete |
|painlv            |Continuous                |Time-varying    |4, 5, 6          |Variable |
|painchr_ext       |Ordinal (4 levels)        |Time-varying    |4, 5, 6          |Variable |
|shlt              |Ordinal (5 levels)        |Time-varying    |4, 5, 6          |Variable |
|adltot6           |Count/Continuous          |Time-varying    |4, 5, 6          |Variable |
|partner           |Binary                    |Time-varying    |4, 5, 6          |Complete |
|ssupport6_log1p   |Continuous                |Time-varying    |4, 5, 6          |Variable (conditional) |
|raagey            |Continuous                |Time-invariant  |Fixed            |Complete |
|smoken            |Binary                    |Time-varying    |4, 5, 6          |Variable |
|work              |Binary                    |Time-varying    |4, 5, 6          |Variable |
|PA_cat            |Ordinal (4 levels)        |Time-varying    |4, 5, 6          |Variable |
|cesd_lat_shifted  |Continuous (shifted)      |Time-varying    |4, 5, 6          |Variable |
|MLTC              |Count/Continuous          |Time-varying    |4, 5, 6          |Variable |
|nettotw_bu_s_log  |Continuous (log-trans.)   |Time-varying    |4, 5, 6          |Variable |
|lvimp_shifted     |Continuous (shifted)      |Time-varying    |4, 5, 6          |Variable |
|drinkwn_e_log   |Continuous (log1p-trans.) |Time-varying    |4, 5, 6          |Variable |


### Data Structure

* Level 1: Observations (waves) nested within individuals
* Level 2: Individuals (participants)

**Level 1 Predictors** (Time-Varying): These are variables that change over time within the same individual across waves 4, 5, and 6. level 1 structural variable is therefore "wave"


```{r}
############################
#### **LEVEL 1 Predictors** 
###########################
# Binary time-varying
bin_TimeVar.long <- c("work","smoken","partner")

# Continuous time-varying
cont_TimeVar.long <- c("painlv",        # ordinal pain severity (5 levels) treated as continuous integer for 2l.pmm; PMM draws from observed values ensuring valid imputed integers
                       "shlt",          # ordinal self-reported health (5 levels) treated as continuous integer for 2l.pmm; same rationale as painlv
                       "drinkwn_e_log","nettotw_bu_s_log",
                       "ssupport6_log1p","lvimp_shifted","cesd_lat_shifted",
                       "lnlys",
                       # the following are integer ordinal/count variables, we re ordered but treating as integer here to facilitate mice predictions
                       "adltot6","MLTC",
                       "PA_cat","painchr_ext") 

lv1_vars<-c(cont_TimeVar.long,bin_TimeVar.long)
```


**Level 2 Predictors** (Time-Invariant): These are variables that remain constant for each individual across all waves (or are measured once at baseline). Level 2 structural variable is therefore "idauniq"


```{r}
############################
#### **LEVEL 2 Predictors** 
###########################

# Binary time-invariant
bin_TimeInvar.long <- c("ragender") ## taking race out because it sample is 97% white

# Continuous time-invariant
cont_TimeInvar.long<- c("raagey", # Age at baseline (though technically increases, treated as baseline)
                        "raeduc_e") # this is ordered but treating as integer here to facilitate mice predictions

lv2_vars<-c(bin_TimeInvar.long,cont_TimeInvar.long)


```


Complete and incomplete vars
```{r}

complete.vars<-names(Missingdata_long)[colSums(is.na(Missingdata_long)) == 0]
incomplete.vars<-names(Missingdata_long)[colSums(is.na(Missingdata_long)) != 0]

# in mice the id variable is required to be integer format
Missingdata_long$idauniq <- as.integer(Missingdata_long$idauniq)  # Convert idauniq to integer

# converting ordered factors to integers
Missingdata_long<-Missingdata_long %>%
                    mutate(
                      adltot6 = as.integer(adltot6),
                      raeduc_e = as.integer(raeduc_e),
                      PA_cat = as.integer(PA_cat),
                      MLTC = as.integer(MLTC),
                      painchr_ext = as.integer(painchr_ext)
                    )

```

## MICE imputation

Linear Mixed-Model-based FCS [@wijesuriya2025multiple]

The FCS-2L approach imputes missing values using a series of conditional LMMs specified for each incomplete variable, tailored according to the distribution of the variable. 

The mice, micemd and miceadds R packages contain functions for implementing FCS-2L, identified by names starting with "2l" (2-level imputation).



In addition in our case we have a conditional variable, that is a variable having structural 
missingness that depend on the value of another variable: 

* Marital Strain (conditional on having a partner)


As with FCS-1L-wide method, the variables first need to be stored as correct type before 
generating the imputations, as described earlier. *The cluster variable must be of type integer*.

Therefore we configure the variable types so that each variable is imputed using an appropriate method: logistic regression for binary variables, and predictive mean matching (`2l.pmm`) for ordinal and continuous variables. For ordinal variables (`painchr_ext`, `PA_cat`) integer-coded with a bounded observed range, PMM is preferred over ordered logistic regression because it draws imputed values directly from observed donor cases, preserving the observed scale without imposing distributional assumptions.

```{r}
# Check missingness by variable
vars_struct_missing<-c("ssupport6_log1p")
incomplete.vars_noStruct<-incomplete.vars[!incomplete.vars %in% vars_struct_missing]


missingness_by_var <- Missingdata_long %>%
  mutate(
    # Mark structural NAs that shouldn't count
    ssupport6_log1p_missing = ifelse(partner == 0, FALSE, is.na(ssupport6_log1p)),
#    mbmi_missing = ifelse(wave == 5, FALSE, is.na(mbmi)),
#    hscrp_missing = ifelse(wave == 5, FALSE, is.na(hscrp))
  ) %>%
  summarise(
    # For conditional variables - only count eligible cases
    ssupport6_log1p_pct = mean(ssupport6_log1p_missing[partner == 1], na.rm = TRUE) * 100,
#    mbmi_pct = mean(mbmi_missing[wave != 5], na.rm = TRUE) * 100,
#    hscrp_pct = mean(hscrp_missing[wave != 5], na.rm = TRUE) * 100,
    
    # For other variables - standard calculation
    across(all_of(incomplete.vars_noStruct), 
           ~mean(is.na(.)) * 100, .names = "{.col}_pct")
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missingness_by_var)


# Variables that enter the LTA model (analytic set only)
lta_vars <- c("idauniq", "wave", "raagey", "ragender", "raeduc_e",
              "nettotw_bu_s_log", "MLTC", "cesd_lat_shifted", "lnlys",
              "partner", "ssupport6_log1p", "painchr_ext", "lvimp_shifted")

# Diagnostic only: percentage of incomplete cases on analytic variables.
# M = 60 is determined by the von Hippel (2020) two-stage FMI procedure
# in chunks 'imputation-stage1' / 'imputation-stage2' below, not by this rule.
analytic_vars_present <- intersect(lta_vars, names(Missingdata_long))
pct_incomplete <- mean(!complete.cases(Missingdata_long[, analytic_vars_present])) * 100
print(paste("Pct incomplete cases (analytic vars):", round(pct_incomplete, 1)))
```


```{r}
library(mice)
library(miceadds)
library(micemd)
library(lme4)       # lmer() for FMI proxy models (longitudinal structure)
library(broom.mixed) # pool() support for lmer objects
# Define imputation methods
meth1 <- make.method(Missingdata_long)

meth1[complete.vars] <- ""  # No imputation for complete variables

## Time variant methods
Lv1_incomplete.vars<-incomplete.vars[incomplete.vars %in% lv1_vars]

incomplete_continuous<-cont_TimeVar.long[cont_TimeVar.long %in% Lv1_incomplete.vars]
meth1[incomplete_continuous] <- "2l.pmm" # Two-level predictive mean matching for incomplete continuous variables

incomplete_binary<-bin_TimeVar.long[bin_TimeVar.long %in% Lv1_incomplete.vars]

meth1[incomplete_binary] <- "2l.bin"  # Two-level logistic regression

## structural missingness will fail with 2l.xxx methods
#meth1["mbmi"] <- "pmm"        # Single-level instead of 2l.pmm (wave 5 structural missing)
#meth1["hscrp"] <- "pmm"       # Single-level instead of 2l.pmm (wave 5 structural missing)
meth1["ssupport6_log1p"] <- "pmm"   # Single-level PMM: 2l.pmm causes convergence failures for
                                   # structurally missing variables (ssupport6 absent for partner==0)
meth1["raeduc_e"] <- "2lonly.pmm"  # Safer if raeduc_e has specific values



## time invariant methods
Lv2_incomplete.vars<-incomplete.vars[incomplete.vars %in% lv2_vars]


# Define predictor Matrix
pred <- make.predictorMatrix(Missingdata_long)

# with "0" we tell which variables are not predictors
# leaving blank the row means all rows
#pred["predicted","predictor"]
# idauniq set to -2 below as cluster variable; this intermediate 0 is overwritten
pred[, "idauniq"] <- -2  # Cluster grouping variable

# time-fixed variables should not be imputed using wave as a predictor —
# wave has no meaning for variables that do not vary within person.
# note: raeduc_e is incomplete and is imputed with 2lonly.pmm (see below);
# ragender and raagey are complete and do not require imputation.
pred[lv2_vars, c("wave")] <- 0



# To specify random slopes for a predictor in a univariate imputation model for a certain variable, 
# number “2” should be included in the corresponding column variable for the row variable that 
# is being imputed. 

# Only use random slopes for variables with sufficient complete cases (20,000+)
# However, we need to prevent overfitting because with random slops we fit 2 parameters 
# per person (random intercept + random slope). With 8014 individuals we need to fit 16,028 random effect parameters, 
# but we only have 23,808 observations. This creates an overfitting situation.

# Apply random slopes selectively
#pred[c("ssupport6_log1p"),"wave"] <- 2 

### Special Exclusions ###
# Prevent painchr_ext from predicting painlv
pred["painchr_ext", "painlv"] <- 0  
pred["painlv", "painchr_ext"] <- 0  
## this was added after I noticed that pmm struggled (numerical instability) with conditional variables as predictors given that these have systematic missingness
conditional_vars <- c("ssupport6_log1p")
pred[, conditional_vars] <- 0

# Exclude partner as predictor OF ssupport6_log1p: the where_matrix restricts imputation to                                                                                
# partner==1 rows only, making partner constant within the imputed subset — singular predictor                                                                             
pred["ssupport6_log1p", "partner"] <- 0                                                                                                                                    


#pred[ord_TimeVar.long, ] <- 0



################################################################################
##### CONDITIONAL IMPUTATION STEP 2 - "WHERE" MATRIX
#################################################################################


where_matrix <- make.where(Missingdata_long, "missing")

Missingdata_long <- Missingdata_long %>%
  mutate(
    # Set support to NA for non-partnered individuals
    ssupport6_log1p = ifelse(partner == 0, NA, ssupport6_log1p),
    # Set BMI and SCRP to NA for wave 5 because there was no nurse visit at that wave
#    mbmi  = ifelse(wave == 5, NA, mbmi),
#    hscrp = ifelse(wave == 5, NA, hscrp)
  )




# Modify where matrix to prevent imputation when partner=0
where_matrix[Missingdata_long$partner == 0 & is.na(Missingdata_long$ssupport6_log1p), "ssupport6_log1p"] <- FALSE
# Modify where matrix to prevent imputation when wave=5 from bmi and HSCRP
#where_matrix[Missingdata_long$wave == 5 & is.na(Missingdata_long$mbmi), "mbmi"] <- FALSE
#where_matrix[Missingdata_long$wave == 5 & is.na(Missingdata_long$hscrp), "hscrp"] <- FALSE



```

We determined the number of imputations using the two-stage quadratic procedure recommended by von Hippel (2020). 

In stage 1, we ran a pilot analysis with M = 20 imputations and estimated the fraction of missing information (γ_mis) using mixed-effects proxy regression on three key variables: both LTA outcome indicators (`painchr_ext` and `lvimp_shifted`) and the primary exposure variable (`ssupport6_log1p`). Because LMest is not compatible with the `mice` pooling interface, proxy regression approximates analysis-level FMI substantially better than intercept-only marginal models. Because the imputed data are in long format with three repeated observations per person, we used `lmer()` with a random intercept per individual rather than OLS, which would treat repeated observations as independent and understate standard errors. Each outcome indicator was regressed on the fixed-effect LTA predictor set across the full dataset; `ssupport6_log1p` was assessed in the partnered subsample only, since by design it is not imputed for unpartnered participants. To be conservative, we used the upper bound of the 95% confidence interval for γ_mis on the logit scale [@harel2007inferences] when calculating the required M for stage 2. The number of imputations was selected to achieve a coefficient of variation of SE ≤ 0.05 (equivalent to ≥ 200 effective degrees of freedom), ensuring that SE estimates would not change materially if the data were re-imputed [@vonhippel2020many]. The worst-case γ_mis across all coefficients and all three proxy models determined M.

The final M used in stage 2 is reported in the diagnostic output below.

```{r imputation-stage1, eval=FALSE}
# ── Stage 1: Pilot imputation (von Hippel 2018) ──────────────────────────────
# Run m=20 pilot, extract FMI via OLS proxy regression on both LTA outcome
# indicators, compute 95% CI upper bound on logit scale (Harel 2007), apply
# quadratic rule to identify the worst-case parameter and recommend M for stage 2.

# Quick sanity check (3 imputations, 5 iterations) — run before the pilot
imp_test <- mice(data = Missingdata_long,
                 m = 3,
                 predictorMatrix = pred,
                 method = meth1,
                 where = where_matrix,
                 diagnostics = TRUE,
                 printFlag = TRUE,
                 maxit = 5)

# Pilot imputation — von Hippel (2018)
set.seed(200720)
imp_pilot <- mice(data = Missingdata_long,
                  m = 20,                  # von Hippel (2018): pilot m=20
                  predictorMatrix = pred,
                  method = meth1,
                  where = where_matrix,
                  diagnostics = TRUE,
                  printFlag = TRUE,
                  maxit = 20) # 5–20 iterations sufficient [@vanbuuren2018flexible]
```


```{r}
# Helper functions — von Hippel (2018) / Harel (2007)
logit     <- function(p) log(p / (1 - p))                          # von Hippel (2018)
inv_logit <- function(x) exp(x) / (1 + exp(x))                     # von Hippel (2018)

ci_upper_fmi <- function(gamma_hat, M) {                            # Harel (2007)
  se_logit    <- sqrt(2 / M)
  upper_logit <- logit(gamma_hat) + 1.96 * se_logit
  inv_logit(upper_logit)
}

recommended_M <- function(gamma_upper, CV_target = 0.05) {         # von Hippel (2018)
  ceiling(1 + 0.5 * (gamma_upper / CV_target)^2)                   # CV_target=0.05 ↔ df=200
}

# ── FMI extraction via lmer proxy models ─────────────────────────────────────
# LMest is not mice-compatible so OLS/lmer proxies are used to approximate
# analysis-level FMI. The data are in long format (3 waves per person), so
# plain lm() would treat repeated observations as independent — understating
# SEs and deflating FMI. lmer() with a random intercept per person is the
# appropriate proxy for this data structure.
#
# Three proxy models:
#   (1) painchr_ext    ~ fixed predictors + (1|idauniq) — full dataset
#   (2) lvimp_shifted  ~ fixed predictors + (1|idauniq) — full dataset
#   (3) ssupport6_log1p ~ fixed predictors + (1|idauniq) — partner==1 subset
#       ssupport6_log1p is one of the most-imputed variables. It is not
#       imputed for partner=0 cases by design (where_matrix), so those rows
#       retain NA in complete() datasets. Restricting to partner==1 gives the
#       imputed values; partner is dropped from predictors (no variation).
#
# pool() accepts a plain list of fitted models and works with lmer via broom.mixed.

# Fixed predictors shared by outcomes (1) and (2)
proxy_pred_outcomes <- "partner + cesd_lat_shifted + MLTC +
                        lnlys + 
                        nettotw_bu_s_log + raagey + ragender + raeduc_e + wave +
                        (1 | idauniq)"

proxy_pred_outcomes_ss <- "ssupport6_log1p + cesd_lat_shifted + MLTC +
                        lnlys + 
                        nettotw_bu_s_log + raagey + ragender + raeduc_e + wave +
                        (1 | idauniq)"

# Fixed predictors for ssupport6 — partner==1 subset, partner omitted
proxy_pred_ss <- "cesd_lat_shifted + MLTC +
                  lnlys + 
                  nettotw_bu_s_log + raagey + ragender + raeduc_e + wave +
                  (1 | idauniq)"

extract_fmi_lmer <- function(mids_obj, outcome, predictors, M,
                              subset_expr = NULL) {
  tryCatch({
    fml  <- as.formula(paste(outcome, "~", predictors))
    fits <- lapply(seq_len(M), function(i) {
      dat <- mice::complete(mids_obj, i)   # tidyr::complete() masks mice::complete()
      if (!is.null(subset_expr)) dat <- dat[eval(subset_expr, dat), ]
      lmer(fml, data = dat, REML = FALSE)
    })
    pooled <- mice::pool(fits)
    params <- pooled$pooled
    data.frame(
      outcome     = outcome,
      parameter   = params$term,
      gamma_hat   = params$fmi,
      gamma_upper = ci_upper_fmi(params$fmi, M = M),     # Harel (2007)
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    warning("Proxy model failed for outcome: ", outcome, " — ", conditionMessage(e))
    data.frame(outcome = outcome, parameter = NA_character_,
               gamma_hat = NA_real_, gamma_upper = NA_real_)
  })
}

fmi_painchr  <- extract_fmi_lmer(imp_pilot, "painchr_ext",     proxy_pred_outcomes, M = 20)
fmi_lvimp    <- extract_fmi_lmer(imp_pilot, "lvimp_shifted",   proxy_pred_outcomes, M = 20)
fmi_painchr_ss  <- extract_fmi_lmer(imp_pilot, "painchr_ext",     proxy_pred_outcomes_ss, M = 20, subset_expr = quote(partner == 1))
fmi_lvimp_ss    <- extract_fmi_lmer(imp_pilot, "lvimp_shifted",   proxy_pred_outcomes_ss, M = 20, subset_expr = quote(partner == 1))


fmi_ssupport <- extract_fmi_lmer(imp_pilot, "ssupport6_log1p", proxy_pred_ss,       M = 20,
                                  subset_expr = quote(partner == 1))

# Combine across all proxy models and apply quadratic rule
fmi_pilot <- rbind(fmi_painchr, fmi_lvimp, 
                   fmi_painchr_ss, fmi_lvimp_ss, 
                   fmi_ssupport)
fmi_pilot$M_recommended <- recommended_M(fmi_pilot$gamma_upper)  # von Hippel (2018)

worst <- fmi_pilot %>%
  filter(!parameter %in% c("lnlys", "wave")) %>%  # excluded: lnlys is auxiliary covariate not LTA outcome;
                                                    # wave is a design variable, not analytically uncertain
  slice_max(gamma_upper, n = 1)

cat("=== Stage 1 Imputation Diagnostics (von Hippel 2018) ===\n")
cat("Pilot imputations : 20\n")
cat("Proxy models      : lmer() with random intercept per person (REML=FALSE)\n")
cat("  (1) painchr_ext      ~ fixed LTA predictors + (1|idauniq)\n")
cat("  (2) lvimp_shifted    ~ fixed LTA predictors + (1|idauniq)\n")
cat("  (3) ssupport6_log1p  ~ predictors + (1|idauniq) [partner==1 subset]\n\n")

cat(sprintf("Worst-case parameter : %s ~ %s\n", worst$outcome, worst$parameter))
cat(sprintf("  gamma_hat          : %.3f\n", worst$gamma_hat))
cat(sprintf("  95%% CI upper       : %.3f\n", worst$gamma_upper))
cat(sprintf("  Recommended M      : %d\n\n", worst$M_recommended))

cat("All parameters (ordered by gamma_upper):\n")
fmi_pilot %>%
  arrange(desc(gamma_upper)) %>%
  print()

cat("\n Conclusion: 'imputation-stage2' only if M_recomended is greater than stage 2.\n")
cat(ifelse(
  worst$M_recommended <= 20,
  "M_recommended <= 20 — pilot imputations are sufficient.\n",
  "M_recommended > 20  — proceed with stage 2 imputations.\n"
))

M_recommended <- worst$M_recommended  # scalar for stage-2 chunk
```

```{r imputation-stage2, eval=FALSE}
# ── Stage 2: Final imputation (von Hippel 2018) ───────────────────────────────
# M set by quadratic rule applied to worst-case gamma_mis upper CI bound from
# Stage 1. See chunk 'imputation-stage1' for diagnostics.

if (!exists("M_recommended") || is.na(M_recommended)) {
  stop("M_recommended not found — run chunk 'imputation-stage1' first.")
}

set.seed(200720)
imp_1 <- mice(data = Missingdata_long,
              m = M_recommended,           # Stage 2 M: von Hippel (2018) quadratic rule
              predictorMatrix = pred,
              method = meth1,
              where = where_matrix,
              diagnostics = TRUE,
              printFlag = TRUE,
              maxit = 20) # 5–20 iterations sufficient [@vanbuuren2018flexible]

saveRDS(imp_1, paste0(paper1_rds_path,"mice_imputationlong_VH.rds"))
```


```{r eval=T}
imp_1<-readRDS(paste0(paper1_rds_path,"mice_imputationlong_VH.rds"))
summary(imp_1)
#saveRDS(imp, paste0(paper1_rds_path,"mice_imputation.rds"))
plot(imp_1)



# Extract one completed dataset
imputed_result <- mice::complete(imp_1, 1)

Missingdata_long %>% vis_miss()
imputed_result %>% vis_miss()
# wave 4 ssupport missingness
imputed_result %>% filter(partner == 1) %>% select(ssupport6_log1p) %>% vis_miss

## ── Post-imputation diagnostics ─────────────────────────────────────────────

# 2lonly.pmm consistency check: raeduc_e is a time-invariant variable imputed
# at the person level. All 3 rows per person must receive the same imputed value.
# The following should return 0 rows — if not, within-person consistency has failed.
raeduc_consistency_check <- mice::complete(imp_1, 1) %>%
  group_by(idauniq) %>%
  summarise(n_unique_educ = n_distinct(raeduc_e), .groups = "drop") %>%
  filter(n_unique_educ > 1)
stopifnot("raeduc_e imputed inconsistently within person" = nrow(raeduc_consistency_check) == 0)

# 1. FMI (Fraction of Missing Information) per analytic variable
# White et al. 2011: FMI > 0.5 suggests m may be inadequate

analytic_in_mids <- intersect(lta_vars, names(imp_1$data))

if (length(analytic_in_mids) > 0) {
  # Fit a null model (intercept only) per variable to extract FMI
  fmi_list <- lapply(analytic_in_mids, function(v) {
    tryCatch({
      fit <- with(imp_1, lm(as.formula(paste(v, "~ 1"))))
      pooled <- pool(fit)
      data.frame(variable = v,
                 fmi = mean(pooled$pooled$fmi, na.rm = TRUE))
    }, error = function(e) data.frame(variable = v, fmi = NA_real_))
  })
  fmi_df <- do.call(rbind, fmi_list) %>% filter(!is.nan(fmi))
}

 fmi_df %>% arrange(desc(fmi)) %>%
  mutate(fmi = round(fmi, 3)) 
  if (any(fmi_df$fmi > 0.5, na.rm = TRUE)){
    warning("FMI > 0.5 for: ", paste(fmi_df$variable[fmi_df$fmi > 0.5], collapse = ", ")," — consider increasing m")
  }else{
    cat(sprintf("All FMI values acceptable (< 0.50), Max FMI : %.4f (%s)\n", max(fmi_df$fmi), fmi_df$variable[which.max(fmi_df$fmi)]))
    }
    
 
 

#2. Density plots: observed vs imputed distributions for continuous variables
densityplot(imp_1, ~ painlv + painchr_ext + shlt + raeduc_e + 
                     #cesd_lat_shifted + lvimp_shifted  # these two variables were complete and are excluded as not imputed
                     lnlys + nettotw_bu_s_log + drinkwn_e_log + ssupport6_log1p,
            main = "Observed (blue) vs Imputed (red) — continuous vars")

densityplot(imp_1, ~ nettotw_bu_s_log + ssupport6_log1p + drinkwn_e_log,
            main = "Observed (blue) vs Imputed (red) — log-transformed vars")

bwplot(imp_1, painlv + painchr_ext + shlt + raeduc_e + nettotw_bu_s_log + ssupport6_log1p + drinkwn_e_log ~ .imp,
       main = "Observed vs imputed — distribution per imputation")

```

