---
title: "LTA Post-processing and MI Pooling"
author: "Diego Vitali"
output:
  github_document:
    toc: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r setup, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(ggplot2)
library(mice)
library(LMest)
library(knitr)

source("Sources/Paper1_func.R")
```

```{r paths, eval=TRUE}
rds_path    <- path.expand("~/private_WP5/WP5_data/rds/")
models_path <- path.expand("~/private_WP5/WP5_data/model_fits/imputations/")

n_imputations <- 22
```

---

# Overview

This document covers the final stage of the LTA pipeline: transforming
the 22 × 2 (TE and DE) raw model fits from
`suppl_pt4.1_imputation_loop.r` into pooled, reportable results.

The pipeline proceeds in three phases:

**Phase A — Per-imputation extraction (Sections 3–6).** For each completed dataset, the raw `lmest` solution is:
1. mapped onto a common severity ordering of the four latent states, 
2. refitted at that ordering to obtain numerical Hessian standard errors, and 
2. summarised into clean coefficient tibbles. Results are saved as `summaries_{TE|DE}_imp{i}.rds`.

**Phase B — Quality checks (Sections 4b–7).** Two cross-imputation checks must pass before pooling: 

* (a) a log-likelihood consistency check confirms no imputation converged to a local maximum; 
* (b) a state ordering consistency check confirms the severity-based reordering produced the same state-to-content mapping across all 22 imputations.

If either check fails, pooling is aborted.

**Phase C — Multiple-imputation pooling (Section 8).** Rubin's rules are applied via `mice::pool.scalar()`, yielding pooled coefficients, 
standard errors, Barnard-Rubin degrees of freedom, p-values, and imputation diagnostics (fmi, r, lambda) for every regression
coefficient in both models.

---

# Formula reconstruction

The formulas passed to `lmest()` must be identical here and in `pt4.1`. They are rebuilt from s
cratch from one completed dataset so that `lmestFormula()` encodes the same factor contrasts that were used
during model fitting. Any drift between the two files would cause `rebase_and_refit_lmest()` to fail or silently misalign parameters.

```{r formulas, eval=TRUE}
updated_mids <- readRDS(paste0(rds_path, "mice_updatedMIDS_VH.rds"))

prep_lta_data <- function(imp_df) {
  imp_df %>%
    mutate(
      ssupport6     = expm1(ssupport6_log1p),
      ssupport6_cat = case_when(
        partner == "No-partner"  ~ "No-partner",
        ssupport6 <= 1.5         ~ "supportive partner",
        ssupport6 >  1.5         ~ "less than supportive partner",
        TRUE                     ~ NA_character_
      ) %>% factor(levels = c("No-partner",
                               "less than supportive partner",
                               "supportive partner"))
    ) %>%
    select(all_of(Variables))
}

Variables <- c(
  "idauniq", "time",
  "raagey_z", "Sex_BIN", "raeduc_e",
  "wealthQ_bin", "MLTC", "cesd_lat_shifted", "loneliness_group",
  "partner", "ssupport6_cat",
  "painchr_ext", "lvimp_shifted_cat3"
)

one_imp           <- prep_lta_data(mice::complete(updated_mids, 1))
latent_state_vars <- c("lvimp_shifted_cat3", "painchr_ext")

TE_initial_prob_vars <- c("Sex_BIN", "raeduc_e", "wealthQ_bin",
                           "ssupport6_cat", "MLTC", "raagey_z")
DE_vars              <- c(TE_initial_prob_vars, "cesd_lat_shifted", "loneliness_group")

fmLatent_TE <- lmestFormula(data = one_imp, response = latent_state_vars,
                             LatentInitial    = TE_initial_prob_vars,
                             LatentTransition = TE_initial_prob_vars)

fmLatent_DE <- lmestFormula(data = one_imp, response = latent_state_vars,
                             LatentInitial    = DE_vars,
                             LatentTransition = DE_vars)
```

```{r ordering-config, eval=TRUE}
ordering_indicators <- c("lvimp_shifted_cat3", "painchr_ext")
ordering_weights    <- c(0.6, 0.4)
```

The composite severity score used to order states is:

$$\text{severity}_k = 0.6 \times E[\texttt{lvimp}|k] + 0.4 \times E[\texttt{painchr}|k]$$

The two indicators have different ranges (`lvimp_shifted_cat3`: 0–2, `painchr_ext`: 0–1). Equal weights would let the wider-range indicator
dominate. The 0.6/0.4 split gives functional impact slightly more pull than the chronic/acute binary. This is consistent with the clinical 
interpretation that impact rather than pain intensity drives more heavily the "severity" dimension. Weights need only be consistent across 
imputations — a state swap can only occur if two states have nearly identical composite scores, and this is caught by `check_state_ordering()`. 

---

# 3. Per-imputation extraction: `extract_one()`

`extract_one()` is the core unit of Phase A (per-imputation extraction). This function performs five steps for
a single imputation:

1.  **Load** `raw_results_{label}_imp{i}.rds` — the best `lmest`
    solution selected from the multi-start grid in `pt4.1`.
2.  **Order states** using `Determine_state_ordering()`: compute the
    composite severity score from the conditional response probability
    matrix (Psi), rank the four states, and print the ordered scores
    with the minimum adjacent gap as an inline sanity check. A gap
    below 0.05 between any two adjacent states is flagged as a
    near-tie — a warning that the boundary between those states is
    numerically fragile.
3.  **Rebase and refit**: `rebase_and_refit_lmest()` permutes the
    parameter matrices (Piv, Pi, Ga, Be) to match the severity order,
    then refits with `out_SE=TRUE` to allow lmest to compute the appropriate 
    standard errors at the rebased solution. 
4.  **Extract summaries**: `regression_summaries()` returns clean
    tibbles of transition and initial-state multinomial logit
    coefficients with per-coefficient N of observation and Event Per Variable (EPV)*.
5.  **Save** `summaries_{label}_imp{i}.rds` and return a 4-row log
    tibble recording which original lmest state label was mapped to
    each new severity slot.

```{r extract-one-def, eval=TRUE}
extract_one <- function(imp, label, fm_latent, fm_responses) {

  cat(sprintf("\n[%s] Imputation %d of %d\n", label, imp, n_imputations))

  raw <- readRDS(paste0(models_path, "raw_results_", label, "_imp", imp, ".rds"))

  # ── state ordering ──────────────────────────────────────────────────────────
  ordering    <- Determine_state_ordering(
    raw$best_model,
    indicator_names   = ordering_indicators,
    indicator_weights = ordering_weights
  )
  state_order <- ordering$severity_order

  scores_ordered <- round(ordering$composite_severity[state_order], 3)
  gaps           <- diff(scores_ordered)
  min_gap_idx    <- which.min(gaps)
  min_gap_label  <- sprintf("S%d-S%d=%.3f", min_gap_idx, min_gap_idx + 1, gaps[min_gap_idx])
  gap_flag       <- if (any(gaps < 0.05)) " !!! NEAR-TIE — check ordering" else " OK"
  cat(sprintf("  S1=%.3f  S2=%.3f  S3=%.3f  S4=%.3f  (orig order: %s)  min gap: %s%s\n",
              scores_ordered[1], scores_ordered[2],
              scores_ordered[3], scores_ordered[4],
              paste(state_order, collapse = "-"),
              min_gap_label, gap_flag))

  # ── rebase + SE ─────────────────────────────────────────────────────────────
  model_rebased <- rebase_and_refit_lmest(
    fit_model        = raw$best_model,
    state_order      = state_order,
    latent_formula   = fm_latent,
    response_formula = fm_responses,
    out_SE           = TRUE,
    plot             = FALSE
  )

  # ── summaries ───────────────────────────────────────────────────────────────
  summaries <- regression_summaries(Model_fit = model_rebased)

  saveRDS(summaries,
          paste0(models_path, "summaries_", label, "_imp", imp, ".rds"))

  # ── state order log entry ───────────────────────────────────────────────────
  log_entry <- tibble(
    imp            = imp,
    new_state      = 1:4,
    orig_state     = state_order,
    severity_score = ordering$composite_severity[state_order]
  )

  rm(raw, model_rebased, summaries)
  gc()

  log_entry
}
```

---

# 4. Cross-imputation log-likelihood check

Before post-processing, we verify that all 22 imputations converged to
the same region of the likelihood surface. Each imputation is an
independently completed dataset drawn from the same MICE posterior, so
all 22 should produce similar best log-likelihoods (same n, same model
structure, similar data).

A large negative outlier — defined as a best LL more than *2 SDs* below
the median. For example, one such negative outlier would indicate that the 
optimiser settled in a local maximum for that imputation. Its parameters describe 
a qualitatively different solution and should not be pooled with the other 21. 
If flagged, options are (a) refit that imputation with more random starts, or (b)
exclude from pooling and document the decision.

`check_ll_consistency()` is defined in `Sources/Paper1_func.R` and
reads `best_model$lk` from every `raw_results_{label}_imp{i}.rds`.

```{r ll-check, eval=TRUE, cache=TRUE}
lls_TE <- check_ll_consistency("TE", n_imputations, models_path)
lls_DE <- check_ll_consistency("DE", n_imputations, models_path)
```

---

# 5. Extraction loop — Total Effects (TE)

The TE model includes the baseline covariates and the partnership/spousal support 
exposure (`ssupport6_cat`) in both the initial-state and transition equations. 
Depression (`cesd_lat_shifted`) and loneliness (`loneliness_group`) are **not** 
included — they are potential mediators on the pathway from partnership to pain, so
conditioning on them in the TE model would partially block the path of
interest.

`bind_rows()` collects the 22 × 4 log tibbles into `state_order_log_TE`
(88 rows), saved for the ordering check in Section 7.

```{r loop-TE, eval=FALSE}
cat("\n", rep("#", 80), "\n")
cat("POST-PROCESSING: TOTAL EFFECTS\n")
cat(rep("#", 80), "\n")

state_order_log_TE <- bind_rows(map(1:n_imputations, function(i) {
  extract_one(i, "TE", fmLatent_TE$latentFormula, fmLatent_TE$responsesFormula)
}))
saveRDS(state_order_log_TE, paste0(models_path, "state_order_log_TE.rds"))
cat("\nTE post-processing complete.\n")
```

---

# 6. Extraction loop — Direct Effects (DE)

The DE model adds depression and loneliness as covariates, estimating the association 
between partnership / spousal support and pain transitions *net* of these potential 
mediators. Comparing TE and DE coefficients for `ssupport6_cat` provides an informal 
indication of how much of the partnership effect (if any) is mediated through depression
or loneliness.

```{r loop-DE, eval=FALSE}
cat("\n", rep("#", 80), "\n")
cat("POST-PROCESSING: DIRECT EFFECTS\n")
cat(rep("#", 80), "\n")

state_order_log_DE <- bind_rows(map(1:n_imputations, function(i) {
  extract_one(i, "DE", fmLatent_DE$latentFormula, fmLatent_DE$responsesFormula)
}))
saveRDS(state_order_log_DE, paste0(models_path, "state_order_log_DE.rds"))
cat("\nDE post-processing complete.\n")
```

---

# 7. State ordering consistency check

Before pooling, we verify that the severity-based reordering assigned the same latent 
content to the same slot across all 22 imputations. This is a prerequisite for 
Rubin's rules: pooling averages parameters labelled "State 2 transition 
coefficient" across imputations, which is only valid if "State 2" refers to 
the same clinical content in every imputation.

`check_state_ordering()` (defined in `Sources/Paper1_func.R`) assesses
this in a deterministic way via a **score range overlap test**. 
For each severity slot (new_state 1–4), it records the minimum and maximum 
composite score that landed in that slot across the 22 imputations. If the maximum
score in slot *k* exceeds the minimum score in slot *k+1*, the boundaries between 
those two states are unstable — some imputations assigned higher-severity cases 
to slot *k* than lower-severity cases assigned to slot *k+1* in other imputations.

Note: `n_unique_orig` in the output table counts how many distinct original lmest 
labels (1–4) mapped to each severity slot. Because lmest assigns arbitrary integer 
labels at each fit, we expext that all four labels will appear in each slot across 
22 imputations — `n_unique_orig = 4`.

If any overlap is detected, `check_state_ordering()` calls `stop()`
to abort execution before pooling runs.

```{r ordering-load, eval=TRUE}
## loading the state_order_data
state_order_log_TE <- readRDS(paste0(models_path, "state_order_log_TE.rds"))
state_order_log_DE <- readRDS(paste0(models_path, "state_order_log_DE.rds"))
```

```{r ordering-check, eval=TRUE}
check_state_ordering(state_order_log_TE, "TE", n_imputations)
check_state_ordering(state_order_log_DE, "DE", n_imputations)
```

---

# 8. Multiple-imputation pooling: Rubin's rules

`pool_summaries()` applies Rubin's rules to pool the 22 per-imputation coefficient 
tibbles into a single set of estimates. Pooling is performed via `mice::pool.scalar()`, 
which implements the Barnard-Rubin (1999) degrees of freedom correction. The function is
called separately for each (varname × Transition) or (varname × State) cell using 
`group_modify()`, which avoids a nest/unnest overhead and keeps the grouping keys in scope.

## 8.1 Output columns

Both `$transitions` and `$initial_states` contain:

| Column | Description |
|--------|-------------|
| `m` | Number of imputations contributing (should equal `n_imputations`) |
| `Q_bar` | Pooled coefficient: mean across *m* imputations ($\bar{Q}$, Rubin 1987 eq. 3.1.2) |
| `U_bar` | Within-imputation variance: mean of *m* squared SEs ($\bar{U}$, eq. 3.1.3) |
| `B` | Between-imputation variance: variance of *m* coefficients (eq. 3.1.4) |
| `T_var` | Total variance: $\bar{U} + (1 + 1/m) \times B$ (eq. 3.1.5) |
| `SE_pool` | Pooled SE: $\sqrt{T_{var}}$ |
| `t_pool` | Pooled t-statistic: $\bar{Q} / \text{SE}_{pool}$ |
| `nu_BR` | Barnard-Rubin df: $(m-1)(1 + \bar{U}/((1+1/m)B))^2$ |
| `p_pool` | Two-sided p-value from $t(\nu_{BR})$ |
| `fmi` | Fraction of missing information (eq. 3.1.10) |
| `r` | Relative increase in variance due to missingness: $(1+1/m)B / \bar{U}$ |
| `lambda` | Proportion of total variance due to missingness: $(1+1/m)B / T_{var}$ |
| `n_NA_SE` | Imputations where SE was NaN (Hessian inversion failed) |
| `N_obs` | Mean observed count in from-state / state across imputations |
| `EPV` | Events per variable: $N_{obs} / n_{predictors}$ |
| `Sep_Flag` | Data quality flag: "OK", "LOW EPV=x", "SE NA in k/22 imps" |
| `Signif` | Significance stars: \*\*\* / \*\* / \* for p < 0.01 / 0.05 / 0.10 |
| `Comment` | "Reliable", "n.s", "Exclude — low EPV", or "Unreliable SE" |

## 8.2 Interpreting the imputation diagnostics

For each pooled estimate, the key diagnostics to inspect are:

- **`r`** (relative increase in variance due to non-response): how
  much larger is the between-imputation variance relative to the
  within-imputation variance? Values near zero mean missingness
  contributes little extra uncertainty for this parameter.
- **`nu_BR`** (residual df for hypothesis testing): should be much
  larger than *m* − 1 = 21. Values approaching 21 indicate that
  between-imputation variance dominates — the Barnard-Rubin
  correction widens the confidence interval relative to a normal
  approximation.
- **`lambda`** (proportion of total variance due to missingness): how
  much of the pooled SE comes from missing data? High lambda for a
  specific transition suggests that parameter is poorly identified
  given the incomplete data.
- **`fmi`** (fraction of missing information): directly interpretable
  as the proportion of statistical information about that parameter
  lost to missingness.

## 8.3 EPV exclusion

Coefficients with EPV < 10 are flagged in `Sep_Flag` and excluded from `Signif` and `Comment`. EPV is the number of observed transitions
from a given state divided by the number of predictors in the transition equation. 
Sparse transitions (few observed events relative to predictors) produce unreliable 
coefficient estimates and inflated SEs regardless of pooling. S
ee Peduzzi et al. (1996) and Vittinghoff & McCulloch (2007).

## 8.4 Run pooling

These chunks save the pooled results to disk. They are set to `eval=FALSE` because 
they write `.rds` files; results are loaded in Section 8.5.

```{r pool-run, eval=FALSE}
pooled_TE <- pool_summaries("TE", n_imputations, models_path)
pooled_DE <- pool_summaries("DE", n_imputations, models_path)

saveRDS(pooled_TE, paste0(models_path, "pooled_TE.rds"))
saveRDS(pooled_DE, paste0(models_path, "pooled_DE.rds"))
```

## 8.5 Load pooled results

```{r pool-load, eval=TRUE}
pooled_TE <- readRDS(paste0(models_path, "pooled_TE.rds"))
pooled_DE <- readRDS(paste0(models_path, "pooled_DE.rds"))
```

---

# 9. Barnard-Rubin df diagnostic

This section is essentially a sensitivity analysis that checks whether the finite-df 
Barnard-Rubin correction materially changes significance conclusions relative to 
the naive z-threshold (|t| > 1.96).

The pooled table already contains `nu_BR`, `p_pool`, `fmi`, `r`, and `lambda` per row, 
but with ~84 transition coefficients per model it is impractical to identify 
borderline cases by eye. `diagnose_barnard_rubin()` scans for two actionable issues:

1. **Low df (nu_BR < 10):** the t-distribution has heavy tails; the critical value at α = 0.05 exceeds 2.23. These estimates have high
   between-imputation variance.
   
2. **Significance flips:** rows where |t_pool| > 1.96 (z says significant) but p_pool > 0.05 
   (t(nu_BR) says not significant), or vice versa. 
   *The predominant case is the anti-conservative direction*: a borderline result that looks significant under z is
   revealed to be non-significant once the finite-df correction is applied.

```{r br-diagnostic, eval=TRUE}
diagnose_barnard_rubin(pooled_TE, "TE")
diagnose_barnard_rubin(pooled_DE, "DE")
```

---

# 10. Key results tables

Filtered to the primary exposure (`ssupport6_cat`) and secondary exposures 
(`cesd_lat_shifted`, `loneliness_group` in DE only), and restricted to EPV ≥ 10.

```{r key-tables-def, eval=TRUE}
print_key_tables <- function(pooled, label, EPV_threshold = 10, round = 3) {

  hicp_transitions     <- c("1 -> 2", "1 -> 3", "1 -> 4", "2 -> 4", "3 -> 4")
  recovery_transitions <- c("4 -> 1", "4 -> 2", "4 -> 3")

  cat(sprintf("\n=== %s KEY RESULTS ===\n", label))

  cat("\n-- Initial state probabilities (ssupport6_cat) --\n")
  pooled$initial_states %>%
    filter(grepl("^ssu", varname), EPV >= EPV_threshold) %>%
    arrange(State, varname) %>%
    mutate(across(where(is.numeric), ~ round(., round))) %>%
    kable(caption = sprintf("%s — initial state probabilities (ssupport6_cat)", label)) %>%
    print()

  cat("\n-- Transitions toward HICP (ssupport6_cat) --\n")
  pooled$transitions %>%
    filter(Transition %in% hicp_transitions, grepl("^ssu", varname), EPV >= EPV_threshold) %>%
    arrange(Transition, varname) %>%
    mutate(across(where(is.numeric), ~ round(., round))) %>%
    kable(caption = sprintf("%s — transitions toward HICP", label)) %>%
    print()

  cat("\n-- Transitions out of HICP / recovery (ssupport6_cat) --\n")
  pooled$transitions %>%
    filter(Transition %in% recovery_transitions, grepl("^ssu", varname), EPV >= EPV_threshold) %>%
    arrange(Transition, varname) %>%
    mutate(across(where(is.numeric), ~ round(., round))) %>%
    kable(caption = sprintf("%s — transitions out of HICP (recovery)", label)) %>%
    print()
}
```

```{r key-tables-run, eval=TRUE}
print_key_tables(pooled_TE, "TE")
print_key_tables(pooled_DE, "DE")
```

---

# 10.5 Observed transition counts and model-implied probabilities

Descriptive tables pooled across all 22 TE-model imputations. State labels use
the pre-verified severity ordering from `state_order_log_TE.rds` — 
Here we do not need re-computation of `Determine_state_ordering()` because we done 
so in the previous phase.

**Imputation uncertainty** is captured by computing each quantity for every
imputation and reporting mean ± SD across the 22 values. This reflects
variability due to the imputation model; within-imputation model estimation
uncertainty is not separately quantified here (Rubin's rules apply to
inferential quantities, not descriptive counts or marginal probabilities).

**Note:** Both count and probability tables are based on Viterbi
(modal/hard-assignment) decoding. Hard assignment underestimates classification
uncertainty relative to soft posterior probabilities, but is standard for
descriptive transition matrices.

```{r trans-matrices, eval=TRUE, cache=TRUE}
# Load pre-verified state ordering
state_order_log_TE <- readRDS(paste0(models_path, "state_order_log_TE.rds"))

state_labels_sev <- c(
  "S1: No pain",
  "S2: Pain, no impairment",
  "S3: Impairment, no CP",
  "S4: HICP"
)

# For imputation i: integer vector s.t. order_i[j] = original LMest label
# whose severity rank is j (1 = least severe).
get_order_i <- function(i) {
  state_order_log_TE %>%
    filter(.data$imp == i) %>%
    arrange(new_state) %>%
    pull(orig_state)
}

# Accumulate across imputations
init_counts_list  <- vector("list", n_imputations)
trans_counts_list <- vector("list", n_imputations)
trans_probs_list  <- vector("list", n_imputations)

for (i in seq_len(n_imputations)) {
  raw     <- readRDS(paste0(models_path, "raw_results_TE_imp", i, ".rds"))
  model   <- raw$best_model
  order_i <- get_order_i(i)

  # ── Initial state counts ──────────────────────────────────────────────────
  ic <- get_observed_InitialStates_counts(model)
  # ic is named "State 1" … "State 4" using LMest original labels; reorder
  init_counts_list[[i]] <- ic[paste("State", order_i)]

  # ── Transition counts ─────────────────────────────────────────────────────
  invisible(capture.output(
    tc <- get_observed_transition_counts(model, option = "global")
  ))
  # tc rows/cols named "From/To State {orig_label}"; reorder to severity
  trans_counts_list[[i]] <-
    tc[paste0("From State ", order_i), paste0("To State ", order_i)]

  # ── Model-implied transition probabilities ────────────────────────────────
  tp <- calc_transition_matrix_prob(model)
  trans_probs_list[[i]] <-
    tp[paste0("From State ", order_i), paste0("To State ", order_i)]
}

# ── Pool: mean ± SD ──────────────────────────────────────────────────────────

# Initial state counts
init_mat   <- do.call(rbind, lapply(init_counts_list, as.numeric))
init_mean  <- colMeans(init_mat)
init_sd    <- apply(init_mat, 2, sd)

init_df <- data.frame(
  State = state_labels_sev,
  `Mean count (SD)` = sprintf("%.1f (%.1f)", init_mean, init_sd),
  check.names = FALSE
)

# Transition counts
tc_array  <- simplify2array(trans_counts_list)   # 4 × 4 × 22
tc_mean   <- apply(tc_array, c(1, 2), mean)
tc_sd     <- apply(tc_array, c(1, 2), sd)
tc_table  <- matrix(sprintf("%.1f (%.1f)", tc_mean, tc_sd), 4, 4,
                    dimnames = list(
                      paste("From", state_labels_sev),
                      paste("To",   state_labels_sev)
                    ))

# Transition probabilities
tp_array  <- simplify2array(trans_probs_list)    # 4 × 4 × 22
tp_mean   <- apply(tp_array, c(1, 2), mean)
tp_sd     <- apply(tp_array, c(1, 2), sd)
tp_table  <- matrix(sprintf("%.3f (%.3f)", tp_mean, tp_sd), 4, 4,
                    dimnames = list(
                      paste("From", state_labels_sev),
                      paste("To",   state_labels_sev)
                    ))
```

## Initial state counts

```{r init-counts-table, eval=TRUE}
kable(init_df,
      caption = "Initial state membership counts — mean (SD) across 22 imputations")
```

## Observed transition counts

```{r trans-counts-table, eval=TRUE}
kable(tc_table,
      caption = "Observed transition counts (Viterbi) — mean (SD) across 22 imputations")
```

## Model-implied transition probabilities

```{r trans-probs-table, eval=TRUE, cache=TRUE}
kable(tp_table,
      caption = "Model-implied marginal transition probabilities — mean (SD) across 22 imputations. Averaged over subjects and time points 2:TT from model$PI.")
```

---

# 11. Effect plots

Three figures display the pooled log-odds ratios with 95% confidence
intervals (computed using the Barnard-Rubin df: `qt(0.975, nu_BR)`).
Estimates with EPV < 10 are excluded. 

```{r}
EPV_PLOT_THRESHOLD <- 20
```

TE and DE model estimates are plotted side by side for direct comparison.

**Figure 1** shows initial-state membership log-ORs for all covariates,
faceted by state (States 2–4 vs State 1 reference). This summarises
which baseline characteristics are associated with starting in a given
pain state.

**Figure 2** shows transition log-ORs for all covariates, faceted by
transition label and grouped by clinical direction (toward HICP /
recovery from HICP). Maintenance transitions (state-to-same-state) are
excluded to focus on clinically meaningful change.

**Figure 3** summarises the primary exposure (`ssupport6_cat`) only,
combining initial states and key transitions in a single panel layout
for the main results interpretation.

```{r plot-config, eval=TRUE}

hicp_transitions     <- c("1 -> 2", "1 -> 3", "1 -> 4", "2 -> 4", "3 -> 4")
recovery_transitions <- c("4 -> 1", "4 -> 2", "4 -> 3")

varname_labels <- c(
  "ssupport6_catless than supportive partner" = "Less than supportive partner \n vs no partner",
  "ssupport6_catsupportive partner"           = "Supportive partner \n vs no partner",
  "cesd_lat_shifted"                          = "CES-D (continuous)",
  "loneliness_groupHigh"                      = "High loneliness \n vs not high loneliness"
)

state_labels <- c(
  "State 1" = "No pain, No impact on functioning",
  "State 2" = "AP or CP, no high-impact \n vs No pain, no impact",
  "State 3" = "NO CP, high-impact \n vs No pain, no impact",
  "State 4" = "HICP vs \nNo pain, no impact"
)

state_caption <- str_wrap(paste(
  "Latent pain states — State 1: no pain and no functional limitations;",
  "State 2: acute or chronic pain without high-impact limitations;",
  "State 3: high-impact limitations without chronic pain;",
  "State 4: high-impact chronic pain (HICP).",
  "CES-D and loneliness included in the direct effects (DE) model only.",
  "Estimates with EPV < 10 excluded. Error bars: 95% CI (Barnard-Rubin df)."
), width = 100)

plot_theme <- theme_minimal(base_size = 11) +
  theme(legend.position  = "bottom",
        plot.caption      = element_text(hjust = 0, size = 8),
        strip.text.y      = element_text(angle = 0))
```

```{r plot-data-prep, eval=TRUE}
init_combined <- bind_rows(
  pooled_TE$initial_states %>% mutate(Model = "TE"),
  pooled_DE$initial_states %>% mutate(Model = "DE")
) %>%
  filter(EPV >= EPV_PLOT_THRESHOLD) %>%
  filter(grepl("^ssu", varname) | grepl("^loneli", varname) | grepl("^cesd", varname)) %>%
  mutate(
    CI_lo   = Q_bar - qt(0.975, df = nu_BR) * SE_pool,
    CI_hi   = Q_bar + qt(0.975, df = nu_BR) * SE_pool,
    varname = recode(varname, !!!varname_labels),
    State   = recode(State,   !!!state_labels)
  )

trans_combined <- bind_rows(
  pooled_TE$transitions %>% mutate(Model = "TE"),
  pooled_DE$transitions %>% mutate(Model = "DE")
) %>%
  filter(EPV >= EPV_PLOT_THRESHOLD) %>%
  filter(grepl("^ssu", varname) | grepl("^loneli", varname) | grepl("^cesd", varname)) %>%
  mutate(
    CI_lo     = Q_bar - qt(0.975, df = nu_BR) * SE_pool,
    CI_hi     = Q_bar + qt(0.975, df = nu_BR) * SE_pool,
    varname   = recode(varname, !!!varname_labels),
    Direction = case_when(
      Transition %in% hicp_transitions     ~ "Toward HICP",
      Transition %in% recovery_transitions ~ "Recovery from HICP",
      Transition == "3 -> 1"               ~ "Improvement",
      TRUE                                  ~ "Maintenance"
    )
  )
```

## Figure 1: Initial state membership

```{r fig1-init, eval=TRUE, fig.width=12, fig.height=6}
p_init <- ggplot(init_combined,
                 aes(x = Q_bar, y = varname, colour = Model,
                     xmin = CI_lo, xmax = CI_hi)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(position = position_dodge(width = 0.3), size = 0.4, linewidth = .6) +
  facet_wrap(~ State, nrow = 1) +
  labs(title   = "Initial state membership — pooled log-ORs (TE vs DE)",
       x       = "Log-OR (95% CI)", y = NULL, colour = "Model",
       caption = state_caption) +
  plot_theme

print(p_init)
```

## Figure 2: Transitions by clinical direction

```{r fig2-trans, eval=TRUE, fig.width=12, fig.height=10}
p_trans <- trans_combined %>%
  filter(Direction != "Maintenance") %>%
  ggplot(aes(x = Q_bar, y = varname, colour = Model,
             xmin = CI_lo, xmax = CI_hi)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(position = position_dodge(width = 0.3), size = 0.4, linewidth = .6) +
  facet_grid(Transition ~ Direction, scales = "free_y", space = "free_y") +
  labs(title   = "Transitions — pooled log-ORs (TE vs DE)",
       x       = "Log-OR (95% CI)", y = NULL, colour = "Model",
       caption = state_caption) +
  plot_theme

print(p_trans)
```

