---
title: "PAPER 1 — pt1.1 MAR Exploration"
author: "Diego Vitali"
output:
  github_document:
    toc: True
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib

---

```{r setup, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(tidyr)      # data reshaping (pivot_longer)
library(dplyr)      # data wrangling
library(ggplot2)    # plotting
library(naniar)     # missingness visualisation (vis_miss, gg_miss_var)
library(readr)      # file I/O
library(mice)       # multiple imputation; mice::complete() for completed datasets
library(patchwork)  # combining ggplot panels
library(ggrepel)    # non-overlapping labels on ggplot (geom_label_repel)
library(broom)      # tidy() for model output (missingness logistic regressions)

source("~/ELSA-partners/Sources/Paper1_func.R")

paper1_rds_path <- "~/private_WP5/WP5_data/rds/"

H_elsa_w4_6 <- readRDS(paste0(paper1_rds_path, "H_elsa_w4_6.rds"))
```

# Overview

This file explores the missing data structure in the analytic dataset
prior to multiple imputation (pt2). It asks two questions:

1. **Pre-imputation:** What is the pattern and magnitude of missingness?
   Is missingness plausibly MAR conditional on observed covariates?

2. **Post-imputation:** Do the imputed values look plausible? Under MAR,
   imputed values should distribute similarly to observed values at
   equivalent missingness probability levels (Bondarenko & Raghunathan
   2016).

The post-imputation section reads `mice_imputationlong_VH.rds`
(output of pt2) and therefore must be run after pt2.

---

# Pre-imputation: missingness patterns

## Variable selection

Select relevant variables: 
(race is excluded as an auxillary variable because sample is 97% white)
```{r}
# This block mirrors the pt2 data prep — keep in sync if pt2 changes.
Missingdata_wide <- H_elsa_w4_6 %>%
  select(c("idauniq",
          #varnames                          # description    (type); whether (comp)lete or (inc)omplete data
           paste0("r",4:6,"painlv"),         #                (categorical); incomp 
           paste0("r",4:6,"painchr_ext"),    #                (categorical); incomp 
           paste0("r",4:6,"lvimp"),          #                (categorical); incomp 
           paste0("r",4:6,"shlt"),           # self reported health               (categorical); incomp 
           paste0("r",4:6,"adltot6"),        # ADL total score                    (numeric); incomp 
           paste0("r",4:6,"partner"),        # having a partner                   (binary); comp 
           paste0("r",4:6,"ssupport6"),      # Marital strain score             (numeric, skewed) incomplete;
           paste0("r",4:6,"lnlys"),          # UCLA - loneliness                (numeric, skewed) incomplete;
           "raagey",                         # Baseline age                     (numeric, normal); comp 
           "ragender",                       # Gender                           (binary); comp 
           #"raracem",                        # Race/ethnicity                   (binary); comp
           paste0("r",4:6,"nettotw_bu_s"),     # Net wealth                     (numeric, skewed); incomp
           "raeduc_e",                       # Education                        (categorical); incomp
           paste0("r",c(4:6),"smoken"),      # smoking                          (binary); incomp
           paste0("r",4:6,"work"),           # Working status                   (binary); incomp
           paste0("r",4:6,"drinkwn_e"),      # Drinking (weekly units)          (continuous); incomp
#           paste0("r",c(4,6),"mbmi"),        # BMI                              (numeric); incomp
#           paste0("r",c(4,6),"hscrp"),       # CRP                              (numeric); incomp
           paste0("r",4:6,"PA_cat"),         # Physical activity                (categorical); incomp
           paste0("r",4:6,"cesd_lat"),       # Depression latent score          (numeric, skewed); incomp
           paste0("r",4:6,"MLTC")            # Multiple long-term condit. count (numeric/count, skewed); incomp
  ))

#str(Missingdata_wide)
```

### Transform skewed variables

Wealth: Shift to positive and log-transform.

Impact and drinking: Shift/log-transform as appropriate.

Create auxiliary variables for change (e.g., MLTC_change_w5, MLTC_change_w6) to 
capture longitudinal change and reduce collinearity.

```{r}
###### TRANSFORM ###################
### Transform wealth data

# Calculate global minimums once
global_min_wealth <- min(c(Missingdata_wide$r4nettotw_bu_s, 
                          Missingdata_wide$r5nettotw_bu_s, 
                          Missingdata_wide$r6nettotw_bu_s), na.rm = TRUE)

global_min_lvimp <- min(c(Missingdata_wide$r4lvimp, 
                          Missingdata_wide$r5lvimp, 
                          Missingdata_wide$r6lvimp), na.rm = TRUE)

global_min_cesd_lat <- min(c(Missingdata_wide$r4cesd_lat, 
                          Missingdata_wide$r5cesd_lat, 
                          Missingdata_wide$r6cesd_lat), na.rm = TRUE)



# Apply same shift to all waves
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4nettotw_bu_s_log = log(r4nettotw_bu_s - global_min_wealth + 1),
    r5nettotw_bu_s_log = log(r5nettotw_bu_s - global_min_wealth + 1),
    r6nettotw_bu_s_log = log(r6nettotw_bu_s - global_min_wealth + 1)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4nettotw_bu_s,-r5nettotw_bu_s,-r6nettotw_bu_s)


### shift imp_data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4lvimp_shifted = r4lvimp - global_min_lvimp,
    r5lvimp_shifted = r5lvimp - global_min_lvimp,
    r6lvimp_shifted = r6lvimp - global_min_lvimp,
  )
# Remove original impact variables

Missingdata_wide<- Missingdata_wide %>%
  select(-r4lvimp,-r5lvimp,-r6lvimp)

### shift depr_lat data
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    # Shift data to make all values positive (if necessary)
    r4cesd_lat_shifted = r4cesd_lat - global_min_cesd_lat,
    r5cesd_lat_shifted = r5cesd_lat - global_min_cesd_lat,
    r6cesd_lat_shifted = r6cesd_lat - global_min_cesd_lat,
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4cesd_lat,-r5cesd_lat,-r6cesd_lat)


### Transform drinkning data
#Missingdata_wide$r4drinkwn_e %>% log() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4drinkwn_e_log  = log1p(r4drinkwn_e),
    r5drinkwn_e_log  = log1p(r5drinkwn_e),
    r6drinkwn_e_log  = log1p(r6drinkwn_e)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4drinkwn_e,-r5drinkwn_e,-r6drinkwn_e)


### Transform support data
#Missingdata_wide$r4ssupport6 %>% log1p() %>% hist(breaks = seq(-1,2,0.1))
Missingdata_wide <- Missingdata_wide %>%
  mutate(
    r4ssupport6_log1p  = log1p(r4ssupport6),
    r5ssupport6_log1p  = log1p(r5ssupport6),
    r6ssupport6_log1p  = log1p(r6ssupport6)
  )
Missingdata_wide<- Missingdata_wide %>%
  select(-r4ssupport6,-r5ssupport6,-r6ssupport6)


#str(Missingdata_wide)

# wave 4 ssupport missingness
ss4<-vis_miss(Missingdata_wide[which(Missingdata_wide$r4partner == 1),] %>% select(r4ssupport6_log1p))
# wave 5 ssupport missingness
ss5<-vis_miss(Missingdata_wide[which(Missingdata_wide$r5partner == 1),] %>% select(r5ssupport6_log1p))
# wave 5 ssupport missingness
ss6<-vis_miss(Missingdata_wide[which(Missingdata_wide$r6partner == 1),] %>% select(r6ssupport6_log1p))

(ss4 + ss5 + ss6)+
  plot_annotation(
    title = "Missingness in conditional variable `Marital strain` total score"
  )


Missingdata_wide %>% select(-r4ssupport6_log1p,-r5ssupport6_log1p,-r6ssupport6_log1p) %>% vis_miss()

```

## Defining variables type
```{r}

bin.vars<-  c(paste0("r",4:6,"work"),         # currently working (bl only)       (binary); incomp 
              paste0("r",4:6,"partner"),      # having a partner (imp condition)  (binary); comp 
              paste0("r",4:6,"smoken"),       # smoking (bl only)                 (binary); incomp
              "ragender"                     # gender                            (binary); incomp
        #      "raracem"                       # race                              (binary); comp
              )

ord.vars<-c(paste0("r",4:6,"PA_cat"),          # Physical Activity                (cat-ordinal); incomp
#            paste0("r",4:6,"shlt"),           # self-reported health             (cat-ordinal); incomp
            "raeduc_e",                        # education                        (cat-ordinal); incomp
            paste0("r",4:6,"painchr_ext")
            )        


Missingdata_wide<-Missingdata_wide %>% mutate(
                                ragender = case_when(
                                          ragender == 1 ~ 0, # male
                                          ragender == 2 ~ 1, # female
                                          TRUE ~ NA_real_
                                          )#,
#                                raracem = case_when(
#                                          raracem == 1 ~ 1, # white
#                                          raracem == 4 ~ 0, #non-white
#                                          TRUE ~ NA_real_
#                                          ),
                                )

```
## Variable type definition

```{r}


# categorical variables,# The first step is that all binary and categorical incomplete variables need to be stored as type “factor”;  @wijesuriya2025multiple


Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(ord.vars), ~factor(., ordered = T)))

Missingdata_wide<- Missingdata_wide %>% mutate(across(all_of(bin.vars), ~factor(.)))

Missingdata_wide<-as.data.frame(Missingdata_wide)

str(Missingdata_wide)

```

```{r}
vars_time_invariant <- c("idauniq", "ragender", 
                         #"raracem", 
                         "raeduc_e","raagey")

Missingdata_long <- Missingdata_wide %>%
  pivot_longer(
    cols = -all_of(vars_time_invariant),
    names_to = c("wave", ".value"),
     # Improved pattern that ensures we match the full variable name
    names_pattern = "^r([456])(.+)$") %>%
      # Convert wave to numeric if needed
      mutate(wave = as.numeric(wave))

```


```{r missingness-overview, fig.width=12, fig.height=6}
# Overall missingness heatmap
vis_miss(Missingdata_long) +
  labs(title = "Missingness map — long format (waves 4–6)",
       caption = "Note: ssupport6_log1p includes structural missingness (partner==0 never collected by design); conditional plot follows.")
```

```{r missingness-by-variable, fig.width=8, fig.height=6}
# ssupport6_log1p has structural missingness: partner==0 cases are never
# collected and never imputed by design. Showing it alongside other variables
# inflates its apparent % missing. Exclude it here; its conditional missingness
# (among partner==1) is shown separately below.
p1<-gg_miss_var(Missingdata_long %>% select(-ssupport6_log1p), show_pct = TRUE) +
  labs(title = "% missing per variable (8014 subjects over 3 waves, tot N rows = 24042)")
```

```{r missingness-ssupport-conditional, fig.width=8, fig.height=3}
# ssupport6_log1p: % missing among partner==1 only (the population for whom
# it was collected)
p2<-gg_miss_var(Missingdata_long %>% filter(partner == 1) %>% select(ssupport6_log1p),
            show_pct = TRUE) +
  labs(title = "% missing: partner support data(among partner==1 only, N rows = 17247)")


p1 / p2 +
  plot_layout(heights = c(15, 1)) +
  plot_annotation(
    caption = "* ssupport6_log1p shown among partner==1 only; structural missingness in partner==0 excluded by design"
  )
```

```{r missingness-table}
# For all variables except ssupport6: % missing over the full long dataset
miss_general <- Missingdata_long %>%
  select(-ssupport6_log1p) %>%
  summarise(across(everything(), ~ round(mean(is.na(.)) * 100, 1))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  mutate(denominator = "all rows")

# For ssupport6: % missing among partner==1 only
miss_ss <- Missingdata_long %>%
  filter(partner == 1) %>%
  summarise(pct_missing = round(mean(is.na(ssupport6_log1p)) * 100, 1)) %>%
  mutate(variable = "ssupport6_log1p", denominator = "partner==1 rows only")

miss_tbl <- bind_rows(miss_general, miss_ss) %>%
  arrange(desc(pct_missing)) %>%
  filter(pct_missing > 0)

knitr::kable(miss_tbl,
             caption = "Variables with any missingness. ssupport6_log1p reported among partner==1 only (structural missingness for partner==0 is by design).")
```

```{r missingness-by-wave}
# Missingness proportion per key analytic variable by wave.
# ssupport6_log1p is computed among partner==1 rows only (structural
# missingness for partner==0 is by design, not informative).
key_vars_no_ss <- c("painchr_ext", "lvimp_shifted", "cesd_lat_shifted",
                    "raeduc_e", "lnlys", "nettotw_bu_s_log", "MLTC", "partner")

miss_wave <- Missingdata_long %>%
  select(wave, all_of(intersect(key_vars_no_ss, names(Missingdata_long)))) %>%
  group_by(wave) %>%
  summarise(across(everything(), ~ round(mean(is.na(.)) * 100, 1)))

miss_wave_ss <- Missingdata_long %>%
  filter(partner == 1) %>%
  group_by(wave) %>%
  summarise(ssupport6_log1p = round(mean(is.na(ssupport6_log1p)) * 100, 1))

miss_wave <- left_join(miss_wave, miss_wave_ss, by = "wave")

knitr::kable(miss_wave,
             caption = "% missing per wave for key analytic variables. ssupport6_log1p: among partner==1 only.")
```

---

# Pre-imputation tests

## Wave-by-Wave attrition

```{r}
miss_wave_long <- miss_wave %>%                                                                                                                       
    pivot_longer(-wave, names_to = "variable", values_to = "pct_missing") %>%                                                                           
    filter(variable %in% c(key_vars_no_ss,"ssupport6_log1p"))                                                                                   

miss_wave_long<- miss_wave_long %>% mutate(
  SCQ = case_when(variable %in% c("lnlys","ssupport6_log1p") ~ "SelfCompQ",
                  TRUE ~ "Interview")
)                                                                                                                   

  ggplot(miss_wave_long, aes(x = wave, y = pct_missing, group = variable, colour = SCQ)) +
    geom_line() + geom_point() +                                                                                         geom_label_repel(
    data = . %>% filter(wave == max(wave)),
    aes(label = variable),
    size = 3, show.legend = FALSE
    ) +
    scale_x_continuous(breaks = c(4, 5, 6)) +                                                                   
    labs(x = "Wave", y = "% missing",
         title = "Missingness by wave — monotonic attrition check") +                                                                                   
    theme_minimal()       
```

## predictors of missingness (MAR justification)

Under MAR, missingness may depend on *observed* data but not on the
missing values themselves. We assess this by fitting logistic regression
models predicting the missingness indicator of each key variable from
observed covariates. Significant predictors support MAR (missingness is
explainable from observed data) rather than MCAR.

Note: logistic models below are fitted on the long-format dataset (3 rows
per person). Standard errors are not cluster-adjusted; this analysis is
exploratory and is not reported in the paper.

We focus on the two variables with the highest expected FMI:
`ssupport6_log1p` (among partnered participants only, where imputation
was attempted) and `painchr_ext`. `lvimp_shifted` was complete and
requires no MAR assessment.

```{r mar-helpers}
# Returns a tidy logistic regression of R(missing) ~ predictors
# Works on the raw (pre-imputation) data.
missingness_logit <- function(data, outcome_var, predictor_vars) {
  R <- as.integer(is.na(data[[outcome_var]]))
  # Drop predictors that are themselves completely missing in this subset
  available_preds <- predictor_vars[
    sapply(predictor_vars, function(v) v %in% names(data) &&
             mean(is.na(data[[v]])) < 1)
  ]
  fml <- as.formula(paste("R ~", paste(available_preds, collapse = " + ")))
  fit <- glm(fml, data = data, family = binomial)
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(outcome = outcome_var, .before = 1)
}
```

```{r mar-ssupport6}
# ssupport6_log1p: restrict to partner==1 (not imputed for partner=0
# by design; MAR question is whether, among partnered participants,
# missingness is predicted by observed characteristics)
preds_ss <- c("cesd_lat_shifted", "MLTC", "lnlys",
              "nettotw_bu_s_log", "raagey", "ragender", "raeduc_e", "wave")

mar_ss <- Missingdata_long %>%
  filter(partner == 1) %>%
  missingness_logit("ssupport6_log1p", preds_ss)

knitr::kable(mar_ss %>% select(outcome, term, estimate, conf.low, conf.high, p.value) %>%
               mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 2)),
                      p.value = round(p.value, 3)),
             caption = "Predictors of P(ssupport6_log1p missing) among partner==1 [OR, 95% CI]")
```

```{r mar-pain-lvimp}
preds_out <- c("partner", "cesd_lat_shifted", "MLTC", "lnlys",
               "nettotw_bu_s_log", "raagey", "ragender", "raeduc_e", "wave")

mar_pain <- missingness_logit(Missingdata_long, "painchr_ext",  preds_out)

knitr::kable(
    mar_pain  %>%
    select(outcome, term, estimate, conf.low, conf.high, p.value) %>%
    mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 2)),
           p.value = round(p.value, 3)),
  caption = "Predictors of missingness for LTA outcome indicators (lv impact omitted as complete) [OR, 95% CI]"
)
```

# Post-imputation: MAR diagnostic (Bondarenko & Raghunathan 2016)

Under MAR, after fitting a good imputation model, the conditional
distribution of imputed values should resemble the conditional
distribution of observed values at similar missingness probabilities.
We assess this using propensity score plots following van Buuren &
Groothuis-Oudshoorn (2011).

For each target variable:

1. Fit a logistic model predicting its missingness from observed
   covariates (predictors vary across imputations; missingness
   indicator is constant).
2. Average fitted propensity scores across M imputations.
3. Plot imputed (filled) vs observed (open) values against propensity
   score — similar distributions support MAR.

```{r load-imputed}
imp_1 <- readRDS(paste0(paper1_rds_path, "mice_imputationlong_VH.rds"))
```

```{r br-helpers}
# propensity_scores(): fits P(variable missing) ~ observed predictors on each
# completed dataset and returns per-row scores averaged over M imputations.
# miss_indicator: full-length 0/1 vector (length = nrow(mids$data))
# subset_rows:    integer index — restrict both predictors and R to these rows
propensity_scores <- function(mids_obj, miss_indicator, predictor_vars,
                               subset_rows = NULL) {
  M   <- mids_obj$m
  fml <- as.formula(paste("R ~", paste(predictor_vars, collapse = " + ")))

  ps_mat <- sapply(seq_len(M), function(i) {
    dat   <- mice::complete(mids_obj, i)
    dat$R <- miss_indicator                                 # full-length assignment
    if (!is.null(subset_rows)) dat <- dat[subset_rows, ]   # subset rows + R together
    fit <- glm(fml, data = dat, family = binomial)
    fitted.values(fit)
  })

  rowMeans(ps_mat)
}

# plot_br(): Bondarenko-Raghunathan diagnostic plot (ggplot2).
# Stacks one completed dataset; colours observed (grey) vs imputed (orange).
# ps: propensity score vector, length = number of rows used (full or subset).
plot_br <- function(mids_obj, var, ps, title, subset_rows = NULL) {
  orig <- mids_obj$data
  imp  <- mice::complete(mids_obj, 1)   # single completed dataset is sufficient

  if (!is.null(subset_rows)) {
    orig <- orig[subset_rows, ]
    imp  <- imp[subset_rows, ]
  }

  plot_dat <- data.frame(
    value = imp[[var]],
    ps    = ps,
    type  = ifelse(is.na(orig[[var]]), "imputed", "observed")
  )

  ggplot(plot_dat, aes(x = ps, y = value, colour = type, shape = type)) +
    geom_point(alpha = 0.4, size = 2.9) +
    scale_colour_manual(values = c(observed = "darkgrey", imputed = "blue")) +
    scale_shape_manual(values  = c(observed = 0, imputed = 19)) +
    labs(x = "P(missing | observed covariates)", y = var,
         title = title, colour = NULL, shape = NULL) +
    theme_minimal()
}
```

```{r br-painchr, fig.width=7, fig.height=4}
R_pain <- as.integer(is.na(imp_1$data$painchr_ext))

ps_pain <- propensity_scores(
  imp_1,
  miss_indicator = R_pain,
  predictor_vars = c("partner", "cesd_lat_shifted", "MLTC", "lnlys",
                     "nettotw_bu_s_log", "raagey", "ragender", "raeduc_e", "wave")
)

plot_br(imp_1, "painchr_ext", ps_pain,
        title = "MAR diagnostic: painchr_ext")
```

Impact scores (`lvimp_shifted`) were complete — no propensity plot needed.

```{r br-ssupport6, fig.width=7, fig.height=4}
# ssupport6_log1p: restrict to partner==1 rows only
partner1_rows <- which(imp_1$data$partner == 1)
R_ss          <- as.integer(is.na(imp_1$data$ssupport6_log1p))

ps_ss <- propensity_scores(
  imp_1,
  miss_indicator = R_ss,
  predictor_vars = c("cesd_lat_shifted", "MLTC", "lnlys",
                     "nettotw_bu_s_log", "raagey", "ragender", "raeduc_e", "wave"),
  subset_rows    = partner1_rows
)

plot_br(imp_1, "ssupport6_log1p", ps_ss,
        title = "MAR diagnostic: ssupport6_log1p (partner==1 only)",
        subset_rows = partner1_rows)
```

---

# Summary

| Variable | Pre-imputation MAR evidence | Post-imputation B&R |
|---|---|---|
| `ssupport6_log1p` | See table above (partner==1 subset) | See plot above |
| `painchr_ext` | See table above | See plot above |
| `lvimp_shifted` | Complete — no missingness; MAR test not applicable | Complete — no propensity plot needed |

Interpret plots: if MAR holds, filled (imputed) and open (observed)
circles should overlap within propensity score bands, with no systematic
vertical separation.
